### 2021-11-04 Rmarkdown/preprocessing/Biocrates_Preprocessing_211103.Rmd
Updated Biocrates preprocessing pipeline

---
title: "Normalize Biocrates Data"
author: "Kristen James"
date: "11/2/2021"
output:
  html_document:
    toc: true
    toc_float: true
---

# About

**Project:** FL100

**About:** We received metabolomics data from the Newman lab run using Biocrates kits. The original file needs to be manipulated to make it data-science ready. Once we format the data for easier manipulation, we will assess its compliance to the normal distribution, per Dr. John Newman's suggestion. He suggested Johnson transformations or log transformations - we will proceed with log transformations because it is easier to understand.

**Inputs:** Biocrates data in a "messy" format.

**Outputs:** Outputs include the following:

1) Data-science approved Biocrates data, non-transformed, BD1 only [bio_bd1]
2) Data-science approved Biocrates data, non-transformed, BD1-4 long format [bio5]
3) Data-science approved Biocrates data, non-transformed, BD1-4 wide format, all cases including NAs [bd.all.w]
4) Data-science approved Biocrates data, non-transformed, BD1-4 wide format, complete cases only [bd.complete.w]
5) Data-science approved Biocrates data, natural log transformed, BD1 only [bio_bd1.ln2]
6) Data-science approved Biocrates data, natural log transformed, BD1-4 wide format, complete cases only[bd.complete.w.ln]

# Set Up

Load libraries.

```{r libraries, results='hide', message=FALSE}
library(dplyr)
```

Load data.

```{r load biocrates data, results='hide'}
# Load data from Jon Newman on 8/5/21
bio <- read.csv("../../data/raw/from_JohnNewman/Prelim_Biocretes_dataset_v3_corrected_PPTime.csv", na.strings=c(""," ","NA"))

bio[1:10,1:10]
```
# Preprocess

This file is ugly. It needs to be tailored before we can confidently maneuver in it.

```{r Beautify biocrates file, results='hide'}
# Set column names
colnames(bio) <- bio[3,]

# Note, can't set row names to subject ID yet bc this data-set has multiple blood draws/person so IDs are not unique. 
# Will subset to BD1 in our first pass. See chunk below.

# Trim junk rows on top
dim(bio) #683 591
bio <- bio[4:677,] #674 591

# Use gsub to get rid of spaces in column names - YUCK!
colnames(bio)
colnames(bio) <- gsub(" ", "_", colnames(bio))
colnames(bio) <- gsub("-", "_", colnames(bio))

# Get rid of blank columns, if needed
emptycols <- sapply(bio, function (k) all(is.na(k)))
table(emptycols) # no empty cols now
bio2 <- bio[!emptycols]
dim(bio2) #674 591

# Look at structure
str(bio2) # Imported as chr. Not what we want. Check that out.
colnames(bio2[,1:10])
# First, subset to the columns we want to turn into numeric values
bio3 <- bio2[,c(7:length(bio2))]
# Then, lapply the combined character->numeric function to turn them into numbers
bio4 <- lapply(bio3, function(x) as.numeric(as.character(x)))
str(bio4)
# Finally, turn your list into a data frame using cbind
bio5 <- do.call(cbind.data.frame, bio4)

```

View handiwork. 
```{r View processed data}
head(bio5[,1:10])
```

So bio5 is in the long format - participants with blood draws at each time point have their ID listed multiple times in the subject ID column. Because there are ~340 people with BD1 and only ~100 people with BD2-4 we should create multiple objects.

Subset to BD1 samples and set row names.

```{r Subset to biocrates BD1, results='hide'}
# Look at individuals by blood draw
table(bio5$pp_Time_min)
# Subset to BD1
bio_bd1 <- bio5[bio5$pp_Time_min == 0,]
head(bio_bd1[,1:10])

# Now set row names to subject ID
rownames(bio_bd1) <- bio_bd1[,"Subject_ID"]
bio_bd1 <- subset(bio_bd1, select = -c(Subject_ID))
```

Save cleaned Biocrates data for future use.

```{r Save Biocrates data}
# BD1
write.csv(bio_bd1, "../../data/processed/biocrates/FL100_biocrates_bd1_rawCounts_211102.csv")
saveRDS(bio_bd1, "../../data/processed/biocrates/FL100_biocrates_bd1_rawCounts_211102.rds")
# all data
write.csv(bio5, "../../data/processed/biocrates/FL100_biocrates_rawCounts_211102.csv")
saveRDS(bio5, "../../data/processed/biocrates/FL100_biocrates_rawCounts_211102.rds")
```

Another file format that will be useful to have is the BD1-4 long data. Reshape the wide data (bio5) and save.

```{r Reshape BD1-4 long to wide}
# Give the object with BD1-4 a more useful name
bd.all.l <- bio5
# Make a wide data frame with the long data
bd.all.w <- reshape(bd.all.l, idvar = "Subject_ID", timevar = "pp_Time_min", direction = "wide")
# Set rownames, which reset during the reshape
rownames(bd.all.w) <- bd.all.w[,"Subject_ID"]
bd.all.w <- subset(bd.all.w, select = -c(Subject_ID))

# Check it out
colnames(bd.all.w[,1:10])
colnames(bd.all.w[,2330:2332])
head(bd.all.w[,585:590])

# Save
write.csv(bd.all.w, "../../data/processed/biocrates/FL100_biocrates_rawCounts_BD1-4_211103.csv")
saveRDS(bd.all.w, "../../data/processed/biocrates/FL100_biocrates_rawCounts_BD1-4_211103.rds")
```

The wide data looks good but once we get into BD2-4 there are a lot of NAs (simply because those samples weren't run at those time points). It will be nice to also have a complete (no NA) wide data frame as well as the full (all samples data frame).

```{r subset to complete wide data}
# Who has complete TMAO data? 
bd.complete.w <- bd.all.w[complete.cases(bd.all.w$TMAO.0)==TRUE &
                            complete.cases(bd.all.w$TMAO.30)==TRUE &
                            complete.cases(bd.all.w$TMAO.180)==TRUE &
                            complete.cases(bd.all.w$TMAO.180)==TRUE,]

# Save
write.csv(bd.complete.w, "../../data/processed/biocrates/FL100_biocrates_rawCounts_BD1-4CompleteCases_211103.csv")
saveRDS(bd.complete.w, "../../data/processed/biocrates/FL100_biocrates_rawCounts_BD1-4CompleteCases_211103.rds")
```

# Transform

Recall, this data has gone through general QC by the Newman lab but it has not been assessed for compliance to the normal distribution. Next, we will complete log transformations and save output.

```{r Log transform BD1}
# Check data
bio_bd1[1:10,1:10] #C0 starts in column 3
dim(bio_bd1)

# Transform continuous columns
bio_bd1.ln <- log(bio_bd1[,2:584])

# Combine back with first 2 rows
bio_bd1.cat <- bio_bd1[,1, drop = FALSE]
bio_bd1.ln2 <- merge(bio_bd1.cat, bio_bd1.ln, by=0)
rownames(bio_bd1.ln2) <- bio_bd1.ln2[,"Row.names"] 
bio_bd1.ln2 <- subset(bio_bd1.ln2, select = -c(Row.names))

# Save bd1.ln2
write.csv(bio_bd1.ln2, "../../data/processed/biocrates/FL100_biocrates_lnTransformed_BD1_211103.csv")
saveRDS(bio_bd1.ln2, "../../data/processed/biocrates/FL100_biocrates_lnTransformed_BD1_211103.rds")
```

Log transform BD1-4 data

```{r Log transform BD1-4 complete data}
# Check data
bd.complete.w[1:10,1:10] #C0 starts in column 3
dim(bd.complete.w)

# Transform continuous columns
bd.complete.w.ln <- log(bd.complete.w)

# Save bd1.ln2
write.csv(bd.complete.w.ln, "../../data/processed/biocrates/FL100_biocrates_lnTransformed_BD1-4_CompleteCases_211103.csv")
saveRDS(bd.complete.w.ln, "../../data/processed/biocrates/FL100_biocrates_lnTransformed_BD1-4_CompleteCases_211103.rds")
```

We have an a priori hypothesis about select uremic toxins, TMAO, p-cresol-SO4, and indoxyl-sulfate. We will handle these variables more carefully. Assess their individual raw distributions, transform, and save.

```{r Uremic Toxins}
shapiro.test(bio_bd1$TMAO)
shapiro.test(bio_bd1$p_Cresol_SO4)
shapiro.test(bio_bd1$Ind_SO4)

shapiro.test(log(bio_bd1$TMAO))
shapiro.test(log(bio_bd1$p_Cresol_SO4))
shapiro.test(log(bio_bd1$Ind_SO4))

shapiro.test((bio_bd1$TMAO)^.5)
shapiro.test((bio_bd1$p_Cresol_SO4)^.5)
shapiro.test((bio_bd1$Ind_SO4)^.5)
```

### 2021-11-05 Rmarkdown/preprocessing/Phyloseq_Preprocessing_211104.Rmd
Update workflows.

---
title: "Phyloseq Preprocessing"
author: "Kristen James"
date: "11/2/2021"
output: html_document
---

# About

**Project:** FL100

**About:** Make Phyloseq objects with the phenotype data and the inputs from Dr. Mary Kable.

**Inputs:** QIIME2 outputs from Dr. Mary Kable

**Outputs:** Outputs include:

1) Processed QIIME2 inputs in R .rds and .csv formats
2) Phyloseq object with XXX phenotypes

# Set Up

Load libraries.

```{r libraries, warning=FALSE}
# Load libraries
library(phyloseq)
library(qiime2R)
library(tibble)
```

Load raw data from Dr. Mary Kable.

```{r Original Data from Mary}
## Metadata
# Metadata file from Mary Kable
metadata <- read.delim("../../data/raw/from_MaryKable/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.txt", stringsAsFactors = F)

# For qza files, make a temporary directory
dir.create("../../data/raw/from_MaryKable/Q2Rtemp")

## OTU Table
# Load the OTU count table 
countTbl <- read_qza(file = "../../data/raw/from_MaryKable/merged-table-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza", tmp = "../../data/raw/from_MaryKable/Q2Rtemp", rm = F)
names(countTbl) # look at names of the list components
countTbl$uuid # unique identifier of the artifact
countTbl$data[1:4,1:4] # the raw data (OTU table), look at a subset
dim(countTbl$data)
# Save the 'data' portion 
countData <- countTbl$data
colnames(countData)
# Note, many subject IDs have odd characters after their ID number. For example, "-X", "r", "-Z" etc. 
# These represent sample duplicates, repeats, etc. The metadata mapping file links the sampleID (XXXXr or XXXX-F10) and the subjectID (XXXX). To simplify downstream analysis, lets use the subject ID. (Can always revert if needed downstream).

# Create a copy to save in next chunk
countData_originalMK <- countData

# Now modify (and save downstream)
colnames(countData) <- gsub("[r,x]$","",colnames(countData))
colnames(countData) <- gsub("\\-.*$","",colnames(countData))
colnames(countData) # check that column names are only subjectID now; good.

## Taxonomy Table
# Load the taxonomy table
taxonTbl <- read_qza(file = "../../data/raw/from_MaryKable/merged-taxonomy-dada2-retrained.qza", tmp = "../../data/raw/from_MaryKable/Q2Rtemp", rm = F)
names(taxonTbl)
taxonTbl$uuid
head(taxonTbl$data) # this reveals that the taxonomy table has classifications to species level (in few cases) (click the right arrow to see). 
# Make taxonomy table human readable
taxTbl<-taxonTbl$data %>% as_tibble() %>% tidyr::separate(Taxon, sep=";", c("Kingdom","Phylum","Class","Order","Family","Genus","Species"))
# View
taxTbl
```

Save

```{r Save QIIME2 files as csv and rds}
# First, save the count table (with sampleID names) - we won't plan to use this but good to have a copy
write.csv(countData_originalMK, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm.csv")
saveRDS(countData_originalMK, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm.rds")

# Save 3 items, which we will use:

# Save the metadata .txt as a csv so that it can be easily altered. 
write.csv(metadata, "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.csv")
saveRDS(metadata, "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")

# Save the count table (with subjectID names)
write.csv(countData, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.csv")
saveRDS(countData, "../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")

# Save the human readable taxonomy table
write.csv(taxTbl, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.csv")
saveRDS(taxTbl, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
```

# Phyloseq Objects

We want to make a phyloseq object for each type of data we are interested in. I find this easier than making one master phyloseq object since sometimes the commands don't handle NAs well. We will make new categorical columns based on the cohort's median value of a given metabolite. Once that is complete, we can create the phyloseq object and save for future use.

## Biocrates Phyloseq Object

Make a phyloseq object for the log transformed biocrates BD1 data. First, we have to do additional preprocessing.

```{r Biocrates BD1 log transformed phyloseq preprocess}
# Load Mary's original memtadata file... that we converted into a csv in the above chunk. 
md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")
rownames(md) <- md[,"Description"]

# Load phenotype data HERE 
df <- readRDS("../../data/processed/biocrates/FL100_biocrates_lnTransformed_BD1_211103.rds")
colnames(df[,1:10])

# Trim df to subjects WITH TMAO (primary variable)
#df <- df[which(complete.cases(df$p.Cresol.SO4.0) == TRUE),]

# First, check the structure of the data. Make sure everything is numeric/integer. 
str(df[,1:15])

# In the event that some variables are factors, run the following code. (Can run regardless, if the original is already numeric the output will be identical)
# In order to make them numeric we first have to turn them to characters. 
df1 <- df # save a copy
# This loops says, "if you are a factor, then first become a character, then a number; if you aren't a factor, stay how you are"
df1[] <- lapply(df1, function(x) {
    if(is.factor(x)) as.numeric(as.character(x)) else x
})

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], 0, 1) # assign 0 if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a dataframe and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])
# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
    if(is.numeric(x)) as.factor(x) else x
})
str(df3[,1:15])
# Visually check that df2 and df3 appear the same; pass check

# Merge categorical data with numeric data
df4 <- merge(df, df3, by = 0)
rownames(df4) <- df4[,1]
df4 <- df4[,-1]

# Merge with metadata file
md2 <- merge(md, df4, by.x = "Description", by.y = 0) # Description = subjectID in the md object
# 103 subjects. These are subjects with biocrates AND microbiome data.

# Assign 'Description' (subject ID in XXXX terms) to row names
rownames(md2) <- md2$Description
```

### Save sample data

```{r Save ln biocrates data}
write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.csv")
saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.rds")
```

### Make Phyloseq object - log biocrates BD1

```{r Make Phyloseq object- log biocrates BD1}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

# Create phyloseq object
PSOlogBioc <- phyloseq(
  otu_table(ASVr, taxa_are_rows = T), 
  phy_tree(treeR$data), 
  tax_table(as.data.frame(taxtable) %>% dplyr::select(-Confidence) %>% column_to_rownames("Feature.ID") %>% as.matrix()), #moving the taxonomy to the way phyloseq wants it
  sample_data(metadata_a)
)
print(PSOlogBioc)

# Save the phyloseq object
save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211104.RData")
```

## TMAO Phyloseq Object

```{r Biocrates BD1 log transformed phyloseq preprocess}
# Load Mary's original memtadata file... that we converted into a csv in the above chunk. 
md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")
rownames(md) <- md[,"Description"]

# Load phenotype data HERE 
df <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
colnames(df[,1:10])

# Trim df to subjects WITH TMAO (primary variable)
df <- df[which(complete.cases(df$tmao) == TRUE),]

# Take a subset
df <- subset(df, select = c( pa_ee_total, endo_ln_rhi, endo_al75, betaine, carnitine, tmao, choline, creatinine, phosphocholine, hei_asa24_totalveg, hei_asa24_totalproteinfoods, hei_asa24_sodium, hei_asa24_refinedgrains, hei_asa24_sfat, hei_asa24_addsug, hei_asa24_wholegrain, hei_asa24_totalscore ))

# Transform
df$tmao_log <- log(df$tmao)

# First, check the structure of the data. Make sure everything is numeric/integer. 
str(df[,1:15])

# In the event that some variables are factors, run the following code. (Can run regardless, if the original is already numeric the output will be identical)
# In order to make them numeric we first have to turn them to characters. 
df1 <- df # save a copy
# This loops says, "if you are a factor, then first become a character, then a number; if you aren't a factor, stay how you are"
df1[] <- lapply(df1, function(x) {
    if(is.factor(x)) as.numeric(as.character(x)) else x
})

# Make categorical variables
input_data <- df1
out <- list()
for(i in seq_along(input_data)){
  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
  out[[i]] <- ifelse(input_data[,i] <= mdn[i], 0, 1) # assign 0 if value is less than or equal to median
}
res <- do.call("cbind",out)
colnames(res) <- paste0(colnames(input_data), "_mdn")
rownames(res) <- rownames(input_data)

# Turn to a dataframe and rename. Then turn variables into factors as they should be.
str(res[,1:15])
df2 <- as.data.frame(res)
str(df2[,1:15])
# Factor
df3 <- df2 # back up
df3[] <- lapply(df3, function(x) {
    if(is.numeric(x)) as.factor(x) else x
})
str(df3[,1:15])
# Visually check that df2 and df3 appear the same; pass check

# Merge categorical data with numeric data
df4 <- merge(df, df3, by = 0)
rownames(df4) <- df4[,1]
df4 <- df4[,-1]

# Merge with metadata file
md2 <- merge(md, df4, by.x = "Description", by.y = 0) # Description = subjectID in the md object
# 103 subjects. These are subjects with biocrates AND microbiome data.

# Assign 'Description' (subject ID in XXXX terms) to row names
rownames(md2) <- md2$Description
```

### Save sample data

```{r Save ln biocrates data}
write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.csv")
saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.rds")
```

### Make Phyloseq object - specific TMAO variables

```{r Make Phyloseq object- specific TMAO variables}
# Load files (metadata, count table, taxonomy table, tree)
# Metadata
metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.rds")
# Count table
ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
# Taxonomy table
taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
# Phylogenetic Tree, rooted
treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)

# Create phyloseq object
PSOtmao <- phyloseq(
  otu_table(ASVr, taxa_are_rows = T), 
  phy_tree(treeR$data), 
  tax_table(as.data.frame(taxtable) %>% dplyr::select(-Confidence) %>% column_to_rownames("Feature.ID") %>% as.matrix()), #moving the taxonomy to the way phyloseq wants it
  sample_data(metadata_a)
)
print(PSOtmao)

# Save the phyloseq object
save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211105.RData")
```

### 2021-11-05 Rmarkdown/preprocessing/Phyloseq_Preprocessing_211104.Rmd
Added alpha diversity to meta data and downstream work flow

@@
 1) Processed QIIME2 inputs in R .rds and .csv formats
-2) Phyloseq object with XXX phenotypes
+2) Phyloseq object with log transformed Biocrates BD1 data [PSOlogBioc]
+3) Phyloseq object with TMAO releated phenotypes [PSOtmao]
 
@@
 
+Alpha diversity statistics
+
+```{r Load Diversity .qza files}
+md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")
+rownames(md) <- md[,"Description"]
+
+# Faith PD
+FaithPD <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/faith_pd_vector.qza", tmp = "Q2RTemp", rm = F)
+FaithPD_DF <- FaithPD$data #353, not 355...
+
+# Shannon
+shannon <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/shannon_vector.qza", tmp = "Q2RTemp", rm = F)
+shannon_DF <- shannon$data #353, not 355...
+
+# Observed ASV
+Observed <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/observed_otus_vector.qza", tmp = "Q2RTemp", rm = F)
+Observed_DF <- Observed$data #353, not 355...
+
+# Evenness ASV
+evenness <- read_qza(file = "../../data/raw/from_MaryKable/core-metrics-results-dada2-tmao/evenness_vector.qza", tmp = "Q2RTemp", rm = F)
+evenness_DF <- evenness$data #353, not 355...
+
+# Merge all of the alpha diversity measures by row names (or a common column name)
+l = list(FaithPD_DF, shannon_DF, Observed_DF, evenness_DF)
+# Make into one DF
+Q2_alpha_DF <- Reduce(merge, lapply(l, function(x) data.frame(x, rn = row.names(x))))
+
+# Now modify (and save downstream)
+Q2_alpha_DF$rn <- gsub("[r,x]$","",Q2_alpha_DF$rn)
+Q2_alpha_DF$rn <- gsub("\\-.*$","",Q2_alpha_DF$rn)
+
+# Merge with MetaData
+metadata2 <- merge(md, Q2_alpha_DF,
+                   by.x = 0,
+                   by.y = "rn",
+                   all.x = TRUE)
+
+rownames(metadata2) <- metadata2[,"Row.names"]
+metadata2 <- subset(metadata2, select =-c(Row.names))
+
+# comment out the all.x line in the merge command and run this code; this shows you that 6072 and 7063 don't have alpha stats.
+#md2[!(rownames(md2) %in% metadata2$Row.names),] # 6072 and 7063 get left out... no alpha stats for them
+
+# Save
+write.csv(metadata2, file = "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.csv")
+saveRDS(metadata2, file = "../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.rds")
+
+```
+
 # Phyloseq Objects
@@
 # Load Mary's original memtadata file... that we converted into a csv in the above chunk. 
-md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min.rds")
+md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.rds")
 rownames(md) <- md[,"Description"]
@@
 # Save the phyloseq object
-save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211104.RData")
+save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211105.RData")
 ```

### 2021-11-10 Rmarkdown/preprocessing/Master_Phenotype_File_211108.Rmd
Update microbiome workflow.

---
title: "Master Pheno FIle"
author: "Kristen James"
date: "11/8/2021"
output: html_document
---

```{r Load TMAO pheno file}
tmao <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
```

Load tobacco data and combine it with the master object.
```{r Load tobbaco data, results='hide'}
dem <- read.csv("../../data/raw/from_RedCAP/demographic/CTSC24532USDAWHNRCNu-DemographicData_DATA_2020-08-28_1058.csv")

dem_tobbaco <- dem[,c(1,26)] #26 in tobbaco use
rownames(dem_tobbaco) <- dem_tobbaco[,"subject_id"]
dem_tobbaco <- subset(dem_tobbaco, select = -c(subject_id))

master <- merge(tmao, dem_tobbaco, by = 0)
rownames(master) <- master[,1]
master <- subset(master, select=-c(Row.names))
```

Load ethnicity data and combine it with the master object.
```{r Load ethnicity data, results='hide'}
eth <- read.csv("../../data/processed/ethnicity/ethnicity_210715.csv",row.names = 1)

# We will reorder the levels so that it matches the order from the demographic survey. This may help prevent mistakes in the future.
str(eth$ethnicity)
eth$ethnicity <- ordered(eth$ethnicity, levels = c("Caucasian", "Hispanic", "African.American", "Asian", "Multi", "Other", "NA"))
# This also turns ethnicity into a factor (not a character)

# Merge
master <- merge(master, eth, by = 0)
rownames(master) <- master[,1]
master <- subset(master, select=-c(Row.names))
```

```{r}
colnames(master)
```
Save
```{r}
write.csv(master, "../../data/processed/phenotype/Phenotypes_211108.csv")
saveRDS(master, "../../data/processed/phenotype/Phenotypes_211108.rds")
```

### 2021-11-10 Rmarkdown/preprocessing/Phyloseq_Preprocessing_211104.Rmd
Update microbiome workflow.

@@
   mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
-  out[[i]] <- ifelse(input_data[,i] <= mdn[i], 0, 1) # assign 0 if value is less than or equal to median
+  out[[i]] <- ifelse(input_data[,i] <= mdn[i], "below", "above") # assign below if value is less than or equal to median
 }
@@
 
-# Turn to a dataframe and rename. Then turn variables into factors as they should be.
+# Turn to a data frame and rename. Then turn variables into factors as they should be.
 str(res[,1:15])
@@
 str(df2[,1:15])
+
 # Factor
@@
 df3[] <- lapply(df3, function(x) {
-    if(is.numeric(x)) as.factor(x) else x
+  if(is.character(x)) as.factor(x) else x
 })
+# Check the levels. R will factor alphabetically but we want the "control" factor to be the below level
 str(df3[,1:15])
-# Visually check that df2 and df3 appear the same; pass check
+
+# Relevel so that "below" is the "control". This matters downstream in regressions when contrasts matter.
+df3.5 <- df3
+df3.5[] <- lapply(df3, function(x) {
+  relevel(x, ref = "below")
+})
+str(df3.5)
 
 # Merge categorical data with numeric data
-df4 <- merge(df, df3, by = 0)
+df4 <- merge(df, df3.5, by = 0)
 rownames(df4) <- df4[,1]
@@
 ```{r Save ln biocrates data}
-write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.csv")
-saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.rds")
+write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.csv")
+saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.rds")
 ```
@@
 # Metadata
-metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211105.rds")
+metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.rds")
 # Count table
@@
 # Save the phyloseq object
-save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211105.RData")
+save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211108.RData")
 ```
@@
 # Load phenotype data HERE 
-df <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
-colnames(df[,1:10])
+df_tmao <- read.csv("../../data/raw/from_RedCAP/TMAO/CTSC24532USDAWHNRCNu-TMAOKJ012220_DATA_2020-07-15_0915_NAedited.csv", row.names = 1)
+colnames(df_tmao[,1:10])
 
 # Trim df to subjects WITH TMAO (primary variable)
-df <- df[which(complete.cases(df$tmao) == TRUE),]
+df_tmao <- df_tmao[which(complete.cases(df_tmao$tmao) == TRUE),]
 
 # Take a subset
+df <- df_tmao
 df <- subset(df, select = c( pa_ee_total, endo_ln_rhi, endo_al75, betaine, carnitine, tmao, choline, creatinine, phosphocholine, hei_asa24_totalveg, hei_asa24_totalproteinfoods, hei_asa24_sodium, hei_asa24_refinedgrains, hei_asa24_sfat, hei_asa24_addsug, hei_asa24_wholegrain, hei_asa24_totalscore ))
@@
   mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
-  out[[i]] <- ifelse(input_data[,i] <= mdn[i], 0, 1) # assign 0 if value is less than or equal to median
+  out[[i]] <- ifelse(input_data[,i] <= mdn[i], "below", "above") # assign 0 if value is less than or equal to median
 }
@@
 df3[] <- lapply(df3, function(x) {
-    if(is.numeric(x)) as.factor(x) else x
+  if(is.character(x)) as.factor(x) else x
 })
+# Check the levels. R will factor alphabetically but we want the "control" factor to be the below level
 str(df3[,1:15])
-# Visually check that df2 and df3 appear the same; pass check
+
+# Relevel so that "below" is the "control". This matters downstream in regressions when contrasts matter.
+df3.5 <- df3
+df3.5[] <- lapply(df3, function(x) {
+  relevel(x, ref = "below")
+})
+str(df3.5)
 
@@
 rownames(md2) <- md2$Description
+
+# Make pathologic variables based on TMAO
+md2$TMAO_5 <- ifelse(md2$tmao <= 5, "below5", "above5")
+md2$TMAO_5 <- as.factor(md2$TMAO_5)
+md2$TMAO_5 <- relevel(md2$TMAO_5, ref = "below5")
+# 6
+md2$TMAO_6 <- ifelse(md2$tmao <= 6, "below6", "above6")
+md2$TMAO_6 <- as.factor(md2$TMAO_6)
+md2$TMAO_6 <- relevel(md2$TMAO_6, ref = "below6")
+# 7
+md2$TMAO_7 <- ifelse(md2$tmao <= 7, "below7", "above7")
+md2$TMAO_7 <- as.factor(md2$TMAO_7)
+md2$TMAO_7 <- relevel(md2$TMAO_7, ref = "below7")
+
+# Lastly, add age, sex, bmi_final to dataframe
+df_bins <- subset(df_tmao, select = c(sex, age, bmi_final))
+# Merge
+md3 <- merge(md2, df_bins, by=0)
+row.names(md3) <- md3[,"Row.names"]
+md3 <- subset(md3, select = -c(Row.names))
 ```
@@
 ```{r Save ln biocrates data}
-write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.csv")
-saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.rds")
+write.csv(md3, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.csv")
+saveRDS(md3, "../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.rds")
 ```
@@
 # Metadata
-metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211105.rds")
+metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/TMAO_specificVariables_mdn_sampledata_211108.rds")
 # Count table
@@
 # Save the phyloseq object
-save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211105.RData")
+save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData")
 ```

### 2021-11-12 Rmarkdown/preprocessing/ASA24_Preprocessing_211111.Rmd
Add ASA24 preprocessing workflow

---
title: "ASA24 Preprocessing"
author: "Kristen James"
date: "11/11/2021"
output:
  html_document:
    toc: true
    toc_float: true
---

# About

**Project:** FL100

**About:** We have QCed ASA24 dietary recalls and we want to find the mean intake

**Inputs:** Inputs include:

1) QCed ASA24 data in wide format with recalls 1, 2, and 3 in one file (referred to as recalls 2, 3, and 4).

**Outputs:** Outputs include:

1) ASA24 data that represents average intake

**Sources:**

1) NA

# Set Up
```{r}
library(dplyr)
```

```{r Load QCed ASA24 recalls}
# Read in ASA24 file with all recalls
dat <- read.csv(file = "../../data/raw/from_RedCAP/asa24/CTSC24532USDAWHNRCNu-ASA24RecallsQCPass_DATA_2020-04-21_1710.csv", row.names = 1)
```

```{r}
# Separate by recall
dat2 <- dat[,grep("2$", colnames(dat))]
dat3 <- dat[,grep("3$", colnames(dat))]
dat4 <- dat[,grep("4$", colnames(dat))] 
#the dimensions are all the same, good; 385x268

# Remove the recall designation
colnames(dat2) <- gsub("2$", "", colnames(dat2))
colnames(dat3) <- gsub("3$", "", colnames(dat3))
colnames(dat4) <- gsub("4$", "", colnames(dat4))

# Save the separated recalls? 
#write.csv(dat2, "../data_processed/factor_analysis/ASA24_Recalls/ASA24_raw_recall2.csv")
#write.csv(dat3, "../data_processed/factor_analysis/ASA24_Recalls/ASA24_raw_recall3.csv")
#write.csv(dat4, "../data_processed/factor_analysis/ASA24_Recalls/ASA24_raw_recall4.csv")

# Check that all column and row names are identical
identical(rownames(dat2), rownames(dat3)) # TRUE
identical(rownames(dat2), rownames(dat4)) # TRUE
identical(colnames(dat2), colnames(dat3)) # TRUE
identical(colnames(dat2), colnames(dat4)) # TRUE

# Before you rowbind, make a new column with subject ID
# R will not let row names be the same. Therefore, if you don't do this,
# then you will get wacky rownames that will cause problems in the future
dat2$subject_id <- rownames(dat2)
dat3$subject_id <- rownames(dat3)
dat4$subject_id <- rownames(dat4)

# Move subject ID to the 1st column for easy viewing
dat2 <- dat2 %>% select(subject_id, everything())
dat3 <- dat3 %>% select(subject_id, everything())
dat4 <- dat4 %>% select(subject_id, everything())

# Rowbind 
dat23 <- rbind(dat2, dat3)
dat_long <- rbind(dat23, dat4)

# Remove columns that are not continuous 
# Remove 1:saltused_tnf  
colnames(dat_long)
dat_long <- dat_long[,-c(2:16)]
# Remove datacomp_totals:numsuppls_tns
colnames(dat_long)
dat_long <- dat_long[,-c(104:114)]
# Remove datacomp_tns:saltused_tnfs
colnames(dat_long)
dat_long <- dat_long[,-c(157:172)]
# Remove datacomp_tnfs:recallqc_passed_
colnames(dat_long)
dat_long <- dat_long[, -c(222:227)]

# Check structure (want all numeric)
str(dat_long)
```

```{r Calculate mean intake}
# Calculate mean intake
# All data
dat_mean <- stats::aggregate(dat_long[,2:221], list(dat_long$subject_id), mean, na.rm = TRUE)
row.names(dat_mean) <- dat_mean[,"Group.1"]
dat_mean <- subset(dat_mean, select = -c(Group.1))
# Note, by specifying na.rm = TRUE, then when there are only 2 entries it will divide by 2 to get mean
# Otherwise, it will put an NA in the mean entry and not calculate a value
# So, this is the mean intake when the food is eaten

# No supplements (many NAs removed)
# If desired, remove columns that refer to supplements only. There are many NAs in supplements. 
# NAs can be problematic.
dat_noSup <- dat_long[, -c(grep("tns$", colnames(dat_long)))]
dat_mean_noSup <- aggregate(dat_noSup[,2:168], list(dat_noSup$subject_id), mean, na.rm = TRUE)
row.names(dat_mean_noSup) <- dat_mean_noSup[,"Group.1"]
dat_mean_noSup <- subset(dat_mean_noSup, select = -c(Group.1))
```

```{r Save averaged data}
# Save 
# csv
write.csv(dat_mean, file = "../../data/processed/asa24/ASA24_meanIntake_211111.csv")
write.csv(dat_mean_noSup, file = "../../data/processed/asa24/ASA24_noSupplements_meanIntake_211111.csv")
# rds
saveRDS(dat_mean, file = "../../data/processed/asa24/ASA24_meanIntake_211111.rds")
saveRDS(dat_mean_noSup, file = "../../data/processed/asa24/ASA24_noSupplements_meanIntake_211111.rds")
```

Many of these variables do not have a normal distribution. We will use this data to run parametric statistics so go ahead and transform key variables for future use.

```{r Transform}
dat_mean2 <- dat_mean

dat_mean2$T.pf_mps_totals_sqrRt <- dat_mean2$pf_meat_totals ^ .5
dat_mean2$T.pf_meat_totals_sqrRt <- dat_mean2$pf_meat_totals ^ .5
dat_mean2$T.pf_curedmeat_totals_sqrRt <- dat_mean2$pf_curedmeat_totals ^ .5
dat_mean2$T.pf_poult_totals_sqrRt <- dat_mean2$pf_poult_totals ^ .5
dat_mean2$T.pf_seafd_hi_totals_Binary <- ifelse(dat_mean2$pf_seafd_hi_totals > 0, "Consumes", "DoesNotConsume")
dat_mean2$T.pf_seafd_low_totals_Binary <- ifelse(dat_mean2$pf_seafd_low_totals > 0, "Consumes", "DoesNotConsume")
dat_mean2$T.pf_eggs_totals_sqrRt <- dat_mean2$pf_eggs_totals ^ .5
dat_mean2$T.pf_nutsds_totals_sqrRt <- dat_mean2$pf_nutsds_totals ^ .5
dat_mean2$T.pf_legumes_totals_sqrRt <- dat_mean2$pf_legumes_totals ^ .5
dat_mean2$T.pf_total_totals_natLog <- log(dat_mean2$pf_total_totals)
dat_mean2$T.pf_soy_totals_Binary <- ifelse(dat_mean2$pf_soy_totals > 0, "Consumes", "DoesNotConsume")
dat_mean2$T.v_drkgr_totals_sqrRt <- dat_mean2$v_drkgr_totals ^ .5
dat_mean2$T.v_redor_total_totals_sqrRt <- dat_mean2$v_redor_total_totals ^ .5
dat_mean2$T.v_redor_tomato_totals_sqrRt <- dat_mean2$v_redor_tomato_totals ^ .5
dat_mean2$T.v_redor_other_totals_sqrRt <- dat_mean2$v_redor_other_totals ^ .5
dat_mean2$T.v_starchy_total_totals_sqrRt <- dat_mean2$v_starchy_total_totals ^ .5
dat_mean2$T.v_starchy_potato_totals_sqrRt <- dat_mean2$v_starchy_potato_totals ^ .5
dat_mean2$T.v_starchy_other_totals_sqrRt <- dat_mean2$v_starchy_other_totals ^ .5
dat_mean2$T.v_other_totals_sqrRt <- dat_mean2$v_other_totals ^ .5
dat_mean2$T.v_total_totals_sqrRt <- dat_mean2$v_total_totals ^ .5
dat_mean2$T.v_legumes_total_sqrRt <- dat_mean2$v_legumes_total ^ .5
dat_mean2$T.g_whole_totals_sqrRt <- dat_mean2$g_whole_totals ^ .5
dat_mean2$T.g_refined_totals_sqrRt <- dat_mean2$g_refined_totals ^ .5
#g_total_totals <- #none
dat_mean2$T.d_total_totals_sqrRt <- dat_mean2$d_total_totals ^ .5
dat_mean2$T.d_milk_totals_sqrRt <- dat_mean2$d_milk_totals ^ .5
dat_mean2$T.d_cheese_totals_sqrRt <- dat_mean2$d_cheese_totals ^ .5
dat_mean2$T.d_yogurt_totals_Binary <- ifelse(dat_mean2$d_yogurt_totals > 0, "Consumes", "DoesNotConsume")
```

Save
```{r Save averaged data with transformations}
# rds
saveRDS(dat_mean2, file = "../../data/processed/asa24/ASA24_meanIntake_withTransformations_211111.rds")
```

Little exploration below. Pay no mind.
```{r}
# Count number of subjects with 2 vs 3 recalls
dat2_count <- dat2[complete.cases(dat2$lang_tnf),] # 384
dat3_count <- dat3[complete.cases(dat3$lang_tnf),] # 378
dat4_count <- dat4[complete.cases(dat4$lang_tnf),] # 347

# What is the distribution of meat intake across the recalls?
summary(dat2$pf_meat_totals)
summary(dat3$pf_meat_totals)
summary(dat4$pf_meat_totals)

summary(dat2$pf_curedmeat_totals)
summary(dat3$pf_curedmeat_totals)
summary(dat4$pf_curedmeat_totals)

t.test(dat2$pf_curedmeat_totals, dat4$pf_curedmeat_totals) # not different, good

# List of TMAO subjects

# List of subjects with at least 2 recalls and TMAO data

# List of subjects with at least 3 recalls and TMAO data
```

### 2021-11-12 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Update analysis workflows

---
title: "Describe Participants: Characteristics, Diet, and Cardiometabolic Risk"
author: "Kristen James"
date: "11/11/2021"
output: html_document
---
# About

**Project:** FL100

**About:** Objectives of the paper analyzed in this markdown include:

1) Who are the participants in the study? --> Demonstrate who is in the study by making tables and plots for the manuscript.
2) What do they eat? --> Describe food intake by TMAO quartile. Complete multiple linear regression analysis.
3) Are TMAO levels related to cardiometabolic risk? --> Complete multiple linear regression analysis.

**Inputs:** Inputs include:

1) Master phenotype data
2) ASA24 averaged data with transformations

**Outputs:** Outputs include:

1)

**Sources:**

```{r Libraries, warning=FALSE, message=FALSE}
library(plyr)
library(tableone)
```

```{r Load data}
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")

# Complete TMAO data only
phen <- phen[complete.cases(phen$tmao == TRUE),]

# Get rid of NA ethnicity category
phen$ethnicity <- droplevels(phen$ethnicity)

# Factor and recode sex
phen$sex <- as.factor(phen$sex)
phen$sex <- plyr::revalue(phen$sex, c("1" = "Male", "2" = "Female"))

# Make TMAO log variable
phen$tmao_log <- log(phen$tmao)
```

# Descriptive Characteristics

## Table 1. Descriptive characteristics of participants

Note, crp is in units of ng/mL. Report in mg/dL so multiply by 0.0001.
il6 is in units of pg/mL
tnfa is in units of pg/mL

```{r Convert crp units}
phen$crp_bd1.mgdL <- phen$crp_bd1 * 0.0001
```

```{r Participant characterisitics}
# TMAO table by age bin - females
factorVars <- c("age_cat")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.f <- CreateTableOne(vars = vars,
                                    strata = "age_cat",
                                    data = phen[phen$sex == "Female",],
                                    factorVars = factorVars)
table_tmao.f

# TMAO table by age bin - males
factorVars <- c("age_cat")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.m <- CreateTableOne(vars = vars,
                                    strata = "age_cat",
                                    data = phen[phen$sex == "Male",],
                                    factorVars = factorVars)
table_tmao.m

# TMAO table by sex
factorVars <- c("sex")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.all <- CreateTableOne(vars = vars,
                                    strata = "sex",
                                    data = phen,
                                    factorVars = factorVars)
table_tmao.all
```

Save the tables

```{r Save characteristic tables}
table_save.f <- print(table_tmao.f, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.f, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byAgeBin_211111.csv")

table_save.m <- print(table_tmao.m, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.m, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byAgeBin_211111.csv")

table_save.all <- print(table_tmao.all, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.all, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_bySex_211111.csv")
```

Is TMAO significantly different by ethnicity?
```{r Ethnicity}
ethnicity_aov <- (aov(tmao_log ~ ethnicity, phen))
summary(ethnicity_aov)
TukeyHSD(ethnicity_aov)

ethnicity_lm <- (lm(tmao_log ~ ethnicity + sex:age + cystatinc_bd1, phen))
summary(ethnicity_lm)
```

There are significant differences in TMAO levels by ethnic groups but ethnicity does not predict TMAO.

Back to tables... make similar tables but show TMAO quantiles

```{r Quantile TMAO}
tmao_quantile <- quantile(phen$tmao)
phen$tmao_quantile <- ifelse(phen$tmao <= tmao_quantile[2], "Quantile1",
                             ifelse(phen$tmao > tmao_quantile[2] & phen$tmao <= tmao_quantile[3], "Quantile2",
                                   ifelse(phen$tmao > tmao_quantile[3] & phen$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))

# Check it
table(phen$tmao_quantile)
phen$tmao_quantile <- as.factor(phen$tmao_quantile)
str(phen$tmao_quantile)
```

```{r Participant characterisitics by TMAO quantile}
# TMAO table by age bin - females
factorVars <- c("age_cat", "tmao_quantile")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.fq <- CreateTableOne(vars = vars,
                                    strata = "tmao_quantile",
                                    data = phen[phen$sex == "Female",],
                                    factorVars = factorVars)
table_tmao.fq

# TMAO table by age bin - males
factorVars <- c("age_cat", "tmao_quantile")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.mq <- CreateTableOne(vars = vars,
                                    strata = "tmao_quantile",
                                    data = phen[phen$sex == "Male",],
                                    factorVars = factorVars)
table_tmao.mq

# TMAO table by sex
factorVars <- c("tmao_quantile")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.allq <- CreateTableOne(vars = vars,
                                    strata = "tmao_quantile",
                                    data = phen,
                                    factorVars = factorVars)
table_tmao.allq
```

```{r Save characteristic quantile tables}
table_save.fq <- print(table_tmao.fq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byTMAOQuantile_211111.csv")

table_save.mq <- print(table_tmao.mq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byTMAOQuantile_211111.csv")

table_save.allq <- print(table_tmao.allq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOQuantile_211111.csv")
```

# Cardiometabolic Risk Analysis

```{r Cardiometabolic Multiple Linear Regression}
# Checked all variables for normality via shapiro.test() in console. TG, and glc ere the only ones that didn't pass.
phen$glc_bd1_ln <- log(phen$glc_bd1)
phen$tg_bd1_ln <- log(phen$tg_bd1)
phen$crp_bd1_ln <- log(phen$crp_bd1)
phen$tnfa_bd1_ln <- log(phen$tnfa_bd1)
phen$il6_bd1_ln <- log(phen$il6_bd1)

transformed_keep_cardio_tbl <- c("sex", "age","ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")

#~~~~~~~~
# TMAO  
#~~~~~~~~
# Make a smaller df from master (the old data)
phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log)) # change variable here
n <- ncol(cvd_biomark)

# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, phen_sub )) # include kcal

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [CVD row, estimate] corresponds to CVD B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [CVD row, Std.Error] corresponds to CVD Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [CVD row, T value] corresponds to CVD t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [CVD row, P] corresponds to CVD P value
my_lm361_lcms_cvd <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm361_lcms_cvd) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm361_lcms_cvd) <- colnames(cvd_biomark)
res <- as.data.frame(my_lm361_lcms_cvd)

# BH correction
res$p.adj <- stats::p.adjust(res$P, method = "BH", n = n)

# Save table
write.csv(res, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_211111.csv")
```

# Diet Analysis

Load the diet data.

```{r Load ASA24 averaged data}
food <- readRDS(file = "../../data/processed/asa24/ASA24_meanIntake_withTransformations_211111.rds")

# Combine food with phenotype data
full <- merge(phen, food, by = 0)
row.names(full) <- full[,"Row.names"]
full <- subset(full, select = -c(Row.names))
```

Make a list of foods and covariate variables to keep.

```{r Foods to keep, results='hide'}
transformed_keep_tbl <- c("sex", 
#"sex_numeric",
"kcal_tnfs", 
"age", 
"cystatinc_bd1", 
"ethnicity",
"tmao", 
"tmao_log", 
#"TMAO.0",
#"TMAO.30",
#"TMAO.180",
#"TMAO.360",
#"lnTMAO.0",
#"lnTMAO.30",
#"lnTMAO.180",
#"lnTMAO.360",
#"TMAO_0_30",
#"TMAO_0_180",
#"TMAO_0_360",
#"Ind.SO4",
#"p.Cresol.SO4",
#"TMAO_quart_num",
#"T.pf_mps_totals_sqrRt",
"T.pf_meat_totals_sqrRt",
"T.pf_curedmeat_totals_sqrRt",
"T.pf_poult_totals_sqrRt",
"T.pf_seafd_hi_totals_Binary",
"T.pf_seafd_low_totals_Binary",
"T.pf_eggs_totals_sqrRt",
"T.pf_nutsds_totals_sqrRt",
"T.pf_legumes_totals_sqrRt",
"T.pf_total_totals_natLog",
#"T.pf_soy_totals_Binary",
"T.v_drkgr_totals_sqrRt",
"T.v_redor_total_totals_sqrRt",
#"T.v_redor_tomato_totals_sqrRt",
#"T.v_redor_other_totals_sqrRt",
"T.v_starchy_total_totals_sqrRt",
#"T.v_starchy_potato_totals_sqrRt",
#"T.v_starchy_other_totals_sqrRt",
"T.v_other_totals_sqrRt",
"T.v_total_totals_sqrRt",
#"T.v_legumes_total_sqrRt",
"T.g_whole_totals_sqrRt",
"T.g_refined_totals_sqrRt",
"g_total_totals",
"T.d_total_totals_sqrRt",
"T.d_milk_totals_sqrRt",
"T.d_cheese_totals_sqrRt",
"T.d_yogurt_totals_Binary"
)
```

Now run the multiple linear regression.

```{r TMAO ~ food - LCMS BD1 - lm 356 ppl, results='hide'}
# Make a smaller df from master (the old data)
full_sub <- full[, colnames(full) %in% transformed_keep_tbl]
food <- subset(full_sub, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
n <- ncol(food)

# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(tmao_log ~ food[,x] + sex*age + cystatinc_bd1 + ethnicity + kcal_tnfs, full_sub )) # include kcal

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm356_lcms) <- colnames(food)
res_food <- as.data.frame(my_lm356_lcms)

# BH correction
res_food$p.adj <- stats::p.adjust(res_food$P, method = "BH", n = n)

# Save table
write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24_211111.csv")
```

### 2021-11-12 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update analysis workflows

---
title: "Stool Microbiome Analysis"
author: "Kristen James"
date: "11/5/2021"
output:
  html_document:
    toc: true
    toc_float: true
---

# About

**Project:** FL100

**About:** Preproccess the data by determining the best prevalence thresholds to use. Run "standard" microbiome analysis focusing on plasma TMAO. This includes alpha and beta diversity, ANCOM differential abundance, and DESeq2 differential abundance analyses.

**Inputs:** Inputs include:

1) TMAO phyloseq object
2) Master phenotype data

**Outputs:** Outputs include:

1) Plots per section [expand as analysis goes final]
2) Tables [expand as analysis goes final]

**Sources:**

1) [Example using Negative Binomial in Microbiome Differential Abundance Testing](https://bioconductor.org/packages/devel/bioc/vignettes/phyloseq/inst/doc/phyloseq-mixture-models.html)
2) [Workflow for Microbiome Data Analysis: from raw reads to community analyses](https://bioconductor.org/help/course-materials/2017/BioC2017/Day1/Workshops/Microbiome/MicrobiomeWorkflowII.html#different_ordination_projections)

# Set Up

```{r, warning=FALSE,message=FALSE}
library(DESeq2)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(vegan)
```

```{r Load phyloseq object}
load("../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData") #PSOtmao
```

# Preproccess

Table of read counts at the Phylum level. We will filter low abundant Phyla. We will also require that taxa are present in at least 20% of samples (at least 71 samples) and occur at least 1000 times. We will then agglomerate at the Genus level.

```{r Tables to Guide Supervised Filtering}
print("The number of different bacteria at each phyla:")
table(tax_table(PSOtmao)[, "Phylum"])

# Prevalence - how many samples the taxa is observed in
prevdf = apply(X = otu_table(PSOtmao),
               MARGIN = ifelse(taxa_are_rows(PSOtmao), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(PSOtmao), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao))

print("Table showing prevalence, total abundance, and taxonomy:")
prevdf[1:20,]

# Make table with mean prevalence, total prevalence
print("Table showing mean prevalencce and total prevalence per Phylum:")
plyr::ddply(prevdf, "Phylum", function(df_mb){cbind("Mean" = mean(df_mb$Prevalence),"Sum" = sum(df_mb$Prevalence))})

```
From the phylum table, we can see that p__Elusimicrobia, p__Spirochaetes, and p__Synergistetes have the same mean and sum, meaning they are only present in 1 sample. We are interested and confident in the more abundant taxa so manually filterr them out.

```{r Supervised Filtering}
# Phylum that only occur 1 time may be artifcats. We can remove them.
filterPhyla <- c(" p__Elusimicrobia", " p__Spirochaetes", " p__Synergistetes") # note, our taxa table has these awful hidden spaces!

# Subset
PSOtmao1 = subset_taxa(PSOtmao, !Phylum %in% filterPhyla)

# View
cat("The Phyla present after this step of subsetting includes:", "\n")
table(tax_table(PSOtmao1)[, "Phylum"])

# Subset to the remaining phyla
prevdf1 = subset(prevdf, Phylum %in% get_taxa_unique(PSOtmao1, "Phylum"))

# Look at phylum level distributions
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1),color=Phylum)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  # Include a guess for parameter
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +  
  xlab("Total Abundance") + 
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Phylum) + 
  theme(legend.position="none")

# Define prevalence threshold as 20% of total samples
prevalenceThreshold = 0.20 * nsamples(PSOtmao1)
cat("The 20% prevalence threshold is", prevalenceThreshold, "samples.", "\n")

# Execute prevalence filter, using `prune_taxa()` function
keepTaxa = rownames(prevdf1)[(prevdf1$Prevalence >= prevalenceThreshold)]
PSOtmao2 = prune_taxa(keepTaxa, PSOtmao1)

cat("Which taxa have less than 1000 counts accross all samples?", "\n")
which(!rowSums(otu_table(PSOtmao2)) > 1000) # 6 OTUS
# Remove these to see if that cleans ordination
PSOtmao3 <- prune_taxa(rowSums(otu_table(PSOtmao2)) >= 1000, PSOtmao2)

# Agglomerate taxa
PSOtmao4 = tax_glom(PSOtmao3, "Genus", NArm = TRUE)
```

Great. Do your due diligence and inspect each phyloseq object to see how the filtering and glomming steps changed the nubmer of taxonomy.

```{r Phyloseq object inspection}
# Inspect the effect of each filtering step on the phyloseq object
PSOtmao1
PSOtmao2
PSOtmao3
PSOtmao4
```

Look at Genus level distributions of the bugs.

```{r Genus level distributions}
# Prevalence - how many samples the taxa is observed in
prevdf = apply(X = otu_table(PSOtmao4),
               MARGIN = ifelse(taxa_are_rows(PSOtmao4), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})

prevdf = data.frame(Prevalence = prevdf,
                    TotalAbundance = taxa_sums(PSOtmao4), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
                    tax_table(PSOtmao4))

print("Table showing prevalence, total abundance, and taxonomy:")
prevdf[1:10,]

# Family
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao4),color=Family)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
  geom_point(size = 1, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Family) +
  theme(legend.position="none")

# Genus - have to blow it up to make it useful
ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao4),color=Genus)) +
  geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
  geom_point(size = 1, alpha = 0.7) +
  scale_x_log10() +
  xlab("Total Abundance") +
  ylab("Prevalence [Frac. Samples]") +
  facet_wrap(~Genus) +
  theme(legend.position="none")
```

# Look for Outliers Using PCoA

We want to identify outlier taxa that might drive the statistical analyses. One way to do this is to look for outliers in the PCoA plot.

```{r PCoA for outliers}
PCoA_BC <- ordinate(PSOtmao4, "PCoA", "bray")

# BC
# taxa is similar to species
taxa_plot <- plot_ordination(PSOtmao4, PCoA_BC, type = "taxa")
taxa_plot
# It looks like there is an outlier in the top left corner of the plot. 
# Identify which taxa that bacteria is from and remove it. 
tp_data <- taxa_plot$data
plot(tp_data$Axis.2) # yes an outlier!
# Who is it?
outlier <- tp_data[tp_data$Axis.2 == max(tp_data$Axis.2), ] 
cat("Which genus is the outlier?", outlier$Genus, "\n")

# Remove it
filterGenus <- c(" g__Prevotella") # note, our taxa table has these awful hidden spaces!

# Subset
PSOtmao5 = subset_taxa(PSOtmao4, !Genus %in% filterGenus)

# Replot:
PCoA_BC2 <- ordinate(PSOtmao5, "PCoA", "bray")

# BC Round 2
# taxa is similar to species
taxa_plot2 <- plot_ordination(PSOtmao5, PCoA_BC2, type = "taxa")
taxa_plot2

# samples is similar to sites
plot_ordination(PSOtmao5, PCoA_BC2, type = "samples", color = "age")

# Which taxa have less than 1000 counts?
which(!rowSums(otu_table(PSOtmao5)) > 1000) # none, good

```

View the phyloseq objects again. We see that the most dramatic effect was when we filtered for phyla present in 20% of samples (PSOtmao1 to 2).

```{r Compare PSOs again}
PSOtmao
PSOtmao1
PSOtmao2
PSOtmao3
PSOtmao4
PSOtmao5 
```

Add age and bmi categories to the "final" phyloseq object. This will come in handy when building the DESeq2 analysis and downstream plotting.

```{r Add meta data to phyloseq object}
# Add age_cat
sample_data(PSOtmao5)$age_cat <- ifelse(sample_data(PSOtmao5)$age < 34, 1,
                         ifelse(sample_data(PSOtmao5)$age >= 34 & sample_data(PSOtmao5)$age < 50, 2, 3))
sample_data(PSOtmao5)$age_cat <- as.factor(sample_data(PSOtmao5)$age_cat)
# Add bmi_cat
sample_data(PSOtmao5)$bmi_cat <- ifelse(sample_data(PSOtmao5)$bmi_final.x < 25, 1,
                         ifelse(sample_data(PSOtmao5)$bmi_final.x >= 25 & sample_data(PSOtmao5)$bmi_final.x < 30, 2, 3))
sample_data(PSOtmao5)$bmi_cat <- as.factor(sample_data(PSOtmao5)$bmi_cat)

# Make sure "below" is the reference factor
str(sample_data(PSOtmao5)$tmao_mdn)
sample_data(PSOtmao5)$tmao_mdn <- relevel(sample_data(PSOtmao5)$tmao_mdn, ref = "below")
```

# DESeq2

From the DESeq2 [documentation](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html), "Note: In order to benefit from the default settings of the package, you should put the variable of interest at the end of the formula and make sure the control level is the first level."

```{r DESeq2 analysis}
# Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
sd <- sample_data(PSOtmao5)
str(sd$sex)
str(sd$bmi_final.x)
# Factor sex, BMI okay
sample_data(PSOtmao5)$sex = factor(sample_data(PSOtmao5)$sex)

# DESeq analysis
mb <- phyloseq_to_deseq2(PSOtmao5, ~ sex:age + bmi_final.x + tmao_mdn)
#mb2 <- estimateSizeFactors(mb, type="poscount")
mb3 = DESeq(mb, test="Wald", fitType="parametric")

# Results
res = results(mb3, cooksCutoff = FALSE)
alpha = 0.05
sigtab = res[which(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao5)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)
```

```{r Plot DESeq2 results}
sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=3) + 
  xlab("Taxonomy") +
  ylab("Log2 Fold Change") + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

# Alpha Diversity Analysis

We need to recalculate the alpha diversity metrics now that we have removed and glommed the taxa data.

```{r Update alpha statistics}
#plot_richness(PSOtmao4, x="tmao_mdn", color="bmi_cat", measures=c("Chao1", "Shannon"))

# Recalculate alpha diversity now that some taxonomy have been filtered out
updated_alpha <- estimate_richness(PSOtmao4)
row.names(updated_alpha) <- gsub("X", "", row.names(updated_alpha))
```

First, load the master phenotype file so that we can have complete access to our covariate variables. Merge the master phenotyp file with the sample data from the phyloseq object.

```{r Alpha diversity load data}
# Load master pheno file
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")
phen <- subset(phen, select = -c(sex, age, age_cat, bmi_cat))

# Check bin structure and factor if needed
#str(phen$age_cat)
#phen$age_cat <- as.factor(phen$age_cat)
#str(phen$bmi_cat)phen$bmi_cat <- as.factor(phen$bmi_cat)

# Grab PSO sample data
df <- as(sample_data(PSOtmao5), "data.frame")

# Merge
df2 <- merge(phen, df, by = 0)
row.names(df2) <- df2[,"Row.names"]
df2 <- subset(df2, select =-c(Row.names))

# Merge updated alpha
df_alpha2 <- merge(updated_alpha, df2, by = 0)
row.names(df_alpha2) <- df_alpha2[,"Row.names"]
df_alpha2 <- subset(df_alpha2, select =-c(Row.names))

# Original alpha
summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df2))
# Updated alpha
summary(lm(df_alpha2$Shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
```

Check the distributions of the alpha diversity metrics and run the multiple linear regression models.

```{r Original Alpha diversity stats}
# Check distribution
shapiro.test(df2$shannon) # not normal, proceed with caution
shapiro.test(log(df2$shannon))
# Run linear regression
summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

# Check distribution
shapiro.test(df2$faith_pd) # normal
# Run linear regression
summary(lm(faith_pd ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

# Check distribution
shapiro.test(df2$pielou_e) # not normal, proceed with caution
# Run linear regression
summary(lm(pielou_e ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

# Check distribution
shapiro.test(df2$observed_otus) # normal
# Run linear regression
summary(lm(observed_otus ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
```

```{r Updated Alpha diversity stats}
#~~~~~~~~~~~
# Shannon
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$Shannon) # not normal, proceed with caution
shapiro.test((df_alpha2$shannon)^.5) # not better
# Run linear regression
summary(lm(Shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

#~~~~~~~~~~~
# Chao1
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$Chao1) # ehh not normal
# Run linear regression
summary(lm(Chao1 ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

#~~~~~~~~~~~
# InvSimpson
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$InvSimpson) # normal
# Run linear regression
summary(lm(InvSimpson ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))

#~~~~~~~~~~~
# Observed
#~~~~~~~~~~~
# Check distribution
shapiro.test(df_alpha2$Observed) # eh not normal
# Run linear regression
summary(lm(Observed ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
```

## Alpha Diversity Plots

Make plots that demonstrate the relationship between TMAO and the alpha-diversity metrics.

```{r Original Alpha Diversity Plots}
# Shannon
y_variable <- "Shannon"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=shannon)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Faith's Phylogenetic Diversity
y_variable <- "Faith's Phylogenetic Diversity"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=faith_pd)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Pielou's Evenness
y_variable <- "Pielou's Evenness"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=pielou_e)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Observed OTUs
y_variable <- "Observed OTUs"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=observed_otus)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

```

```{r Updated Alpha Diversity Plots}
# Shannon
y_variable <- "Shannon"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=Shannon)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Faith's Phylogenetic Diversity
y_variable <- "Chao1"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=Chao1)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Pielou's Evenness
y_variable <- "Inverse Simpson"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=InvSimpson)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

# Observed OTUs
y_variable <- "Observed OTUs"
x_variable <- "Natural Log of TMAO (uM)"
mdn_cut <- log(median(df_alpha2$tmao.x))
# Plot
ggplot(df_alpha2, aes(x=tmao_log, y=Observed)) +
  geom_point(aes(shape=bmi_cat, color = age_cat)) +
  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
  theme_minimal() + 
  ylab(y_variable) +
  xlab(x_variable) +
  labs(color = "Age bin", shape = "BMI bin") +
  scale_color_brewer(palette = "Dark2")

```

# Beta Diversity Analysis

```{r Beta diversity calculations}
# PCoA
PCoA_WU <- ordinate(PSOtmao5, "PCoA", "wunifrac")
PCoA_UnWU <- ordinate(PSOtmao5, "PCoA", "uunifrac")
PCoA_BC <- ordinate(PSOtmao5, "PCoA", "bray")

# Calculate distance matrix
dismax_wu = phyloseq::distance(PSOtmao5, method="wunifrac", type = "samples")
dismax_uwu = phyloseq::distance(PSOtmao5, method="unifrac", type = "samples")
dismax_bc = phyloseq::distance(PSOtmao5, method="bray", type = "samples")

```

```{r Beta diversity stats}
# Get sample_data
df <- as(sample_data(PSOtmao5), "data.frame")
df$tmao_mdn <- relevel(df$tmao_mdn, ref = "below")

#~~~~~~~~~~~
# Weighted
#~~~~~~~~~~~
# Beta dispersion
bdis <- betadisper(dismax_wu, df$tmao_mdn)
permutest(bdis) # NS, good
boxplot(bdis, las = 2)

# ADONIS (perMANOVA) analysis
adonis2(dismax_wu ~ tmao_log + sex:age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .025

#~~~~~~~~~~~
# Unweighted
#~~~~~~~~~~~
# Beta dispersion
bdis <- betadisper(dismax_uwu, df$tmao_mdn)
permutest(bdis) # NS, good
boxplot(bdis, las = 2)

# ADONIS (perMANOVA) analysis
adonis2(dismax_uwu ~ tmao_log + sex:age + bmi_final.y, data=df, method = "unifrac", perm=999) # .001

#~~~~~~~~~~~
# BrayCurtis
#~~~~~~~~~~~
# Beta dispersion
bdis <- betadisper(dismax_bc, df$tmao_mdn)
permutest(bdis) # NS, good
boxplot(bdis, las = 2)

# ADONIS (perMANOVA) analysis
adonis2(dismax_bc ~ tmao_log + sex:age + bmi_final.y, data=df, method = "bray", perm=999) # .017

```

# ANCOM

```{r}
source("../../Rscripts/ancom_v2.1.R")

# OTU table
MicroData <- otu_table(PSOtmao5)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao5)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df_alpha2 %>% dplyr::select("tmao_mdn")
MetaData2$Sample.ID <- rownames(MetaData2)

feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
# Function
prepro = feature_table_pre_process(feature_table = feature_table,
                                   meta_data = MetaData2,
                                   sample_var = sample_var,
                                   group_var = NULL,
                                   out_cut = out_cut,
                                   zero_cut = zero_cut,
                                   lib_cut = lib_cut,
                                   neg_lb = neg_lb)

# Step 2: ANCOM
feature_table = prepro$feature_table # preprocessing went from 98 to 60 bugs
meta_data = prepro$meta_data
struc_zero = prepro$structure_zeros

# Define variables for ANCOM
main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
adj_formula = NULL; rand_formula = NULL
# Run ANCCOM
t_start = Sys.time()
res = ANCOM(feature_table = feature_table, 
            meta_data = meta_data, 
            struc_zero = struc_zero, 
            main_var = main_var, 
            p_adj_method = p_adj_method, 
            alpha = alpha, 
            adj_formula = adj_formula, 
            rand_formula = rand_formula)
t_end = Sys.time()
t_run = t_end - t_start 
t_run # around 5s

# Results
res$out
res$fig$data

# Step 3: Volcano Plot
# Number of taxa except structural zeros
n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")

# Annotation data
dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")

fig = res$fig +  
  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
fig

# ORGANIZE PLOT

TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
rownames(TaxaData2) <- TaxaData2$Feature.ID

ancomRes <- res$out

test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)

# Make plot 
tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
asv_id <- unique(tempdata$taxa_id)

MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]

# Rename to the taxonomy 
#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
# Make a unique naming column that incorporates multiple levels of taxonomy
TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
rownames(TaxaData3) <- rownames(TaxaData2)
# Merge
MicroData5 <- merge(MicroData4, TaxaData3,
                    by=0, all.x = T, all.y = F)

row.names(MicroData5) <- MicroData5$OrderFamilyGenus

MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))

# Merge with the phenotype data 
MetaData3 <- MetaData2 %>% dplyr::select(tmao_mdn)

ggData <- merge(MetaData3, MicroData5, by =0)
row.names(ggData) <- ggData$Row.names
ggData <- ggData %>% dplyr::select(-Row.names)

# Gather the data
library(tidyr)
ggData2 <- gather(ggData,
                  key = "Taxa",
                  value = "Abundance",
                  -tmao_mdn)

ggData2$Abundance <- log(ggData2$Abundance)

# Plot it
p <- ggplot(data = ggData2,
            aes(x = Taxa, 
                y = Abundance,
                fill = tmao_mdn), 
            alpha = 0.1) +
  geom_boxplot(lwd=.25) +
  xlab(NULL) +
  ylab("Log of Taxa Abundance") +
  theme_bw() + # note order matters here - need bw to be before theme()
  theme(axis.text.y = element_text( hjust = 1),
        axis.text.x = element_text(angle = 60, hjust = 1),
        text=element_text(size=11))  +
  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
  #scale_fill_viridis(discrete = TRUE) +
  #coord_flip() +
  labs(fill = "TMAO \n Classification")

p
```

### 2021-11-16 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Update workflow.

@@
 
-ethnicity_lm <- (lm(tmao_log ~ ethnicity + sex:age + cystatinc_bd1, phen))
+ethnicity_lm <- (lm(tmao_log ~ ethnicity + sex*age + cystatinc_bd1, phen))
 summary(ethnicity_lm)
@@
 str(phen$tmao_quantile)
+
+# Checked all variables for normality via shapiro.test() in console. TG, and glc ere the only ones that didn't pass.
+phen$glc_bd1_ln <- log(phen$glc_bd1)
+phen$tg_bd1_ln <- log(phen$tg_bd1)
+phen$crp_bd1_ln <- log(phen$crp_bd1)
+phen$tnfa_bd1_ln <- log(phen$tnfa_bd1)
+phen$il6_bd1_ln <- log(phen$il6_bd1)
 ```
@@
 
+Note, we will remove ethnicity in these regressions because when we specifically asked if ethnicity was related to the cardiometabolic risk variables there were all non-significant relationships (uncomment the line to demonstrate for yourself).
+We will keep cystatin C in the regression using the same reasoning - many significant relationships.
+
 ```{r Cardiometabolic Multiple Linear Regression}
-# Checked all variables for normality via shapiro.test() in console. TG, and glc ere the only ones that didn't pass.
-phen$glc_bd1_ln <- log(phen$glc_bd1)
-phen$tg_bd1_ln <- log(phen$tg_bd1)
-phen$crp_bd1_ln <- log(phen$crp_bd1)
-phen$tnfa_bd1_ln <- log(phen$tnfa_bd1)
-phen$il6_bd1_ln <- log(phen$il6_bd1)
-
-transformed_keep_cardio_tbl <- c("sex", "age","ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
+# Transfored variable list
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
 
@@
 # Run linear model
-my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, phen_sub )) # include kcal
+my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex + age + cystatinc_bd1, phen_sub )) # include kcal
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ ethnicity, phen_sub ))
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ cystatinc_bd1, phen_sub ))
 
@@
 res$p.adj <- stats::p.adjust(res$P, method = "BH", n = n) 
+```
 
+Save table
+```{r Save TMAO linear regression table}
 # Save table
@@
 ```
+
+## Females 
+```{r Cardiometabolic Multiple Linear Regression - Females}
+# Transfored variable list
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
+
+#~~~~~~~~
+# TMAO  
+#~~~~~~~~
+# Make a smaller df from master (the old data)
+phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
+# Sex data frame
+phen_sub.f <- phen_sub[phen_sub$sex == "Female",]
+cvd_biomark <- subset(phen_sub.f, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log)) # change variable here
+n <- ncol(cvd_biomark)
+
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + age + cystatinc_bd1, phen_sub.f )) # take out sex
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ ethnicity, phen_sub ))
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ cystatinc_bd1, phen_sub ))
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [CVD row, estimate] corresponds to CVD B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [CVD row, Std.Error] corresponds to CVD Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [CVD row, T value] corresponds to CVD t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [CVD row, P] corresponds to CVD P value
+my_lm361_lcms_cvd.f <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm361_lcms_cvd.f) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm361_lcms_cvd.f) <- colnames(cvd_biomark)
+res.f <- as.data.frame(my_lm361_lcms_cvd.f)
+
+# BH correction
+res.f$p.adj <- stats::p.adjust(res.f$P, method = "BH", n = n) 
+```
+
+Save table
+```{r Save TMAO linear regression table - Females}
+# Save table
+write.csv(res.f, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_females_211112.csv")
+```
+
+No significant effects found in female only cohort.
+
+## Males 
+```{r Cardiometabolic Multiple Linear Regression - Males}
+# Transfored variable list
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
+
+#~~~~~~~~
+# TMAO  
+#~~~~~~~~
+# Make a smaller df from master (the old data)
+phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
+# Sex data frame
+phen_sub.m <- phen_sub[phen_sub$sex == "Male",]
+cvd_biomark <- subset(phen_sub.m, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log)) # change variable here
+n <- ncol(cvd_biomark)
+
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + age + cystatinc_bd1, phen_sub.m )) # take out sex
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ ethnicity, phen_sub ))
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ cystatinc_bd1, phen_sub ))
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [CVD row, estimate] corresponds to CVD B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [CVD row, Std.Error] corresponds to CVD Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [CVD row, T value] corresponds to CVD t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [CVD row, P] corresponds to CVD P value
+my_lm361_lcms_cvd.m <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm361_lcms_cvd.m) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm361_lcms_cvd.m) <- colnames(cvd_biomark)
+res.m <- as.data.frame(my_lm361_lcms_cvd.m)
+
+# BH correction
+res.m$p.adj <- stats::p.adjust(res.m$P, method = "BH", n = n) 
+```
+
+Save table
+```{r Save TMAO linear regression table - Males}
+# Save table
+write.csv(res.m, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_males_211112.csv")
+```
+
+No significant effects found in male only cohort.
 
@@
 Now run the multiple linear regression.
+Note, we will not include ethnicity in these regressions because although TMAO levels significantly differ between ethnic groups, there was no relationship between TMAO and ethnicity.
+Additionally, diet patterns differ by ethnicity so we could be over adjusting.
 
@@
 # Run linear model
-my_lm <- lapply(1:n, function(x) lm(tmao_log ~ food[,x] + sex*age + cystatinc_bd1 + ethnicity + kcal_tnfs, full_sub )) # include kcal
+my_lm <- lapply(1:n, function(x) lm(tmao_log ~ food[,x] + sex*age + cystatinc_bd1 + kcal_tnfs, full_sub )) # include kcal
 
@@
 res_food$p.adj <- stats::p.adjust(res_food$P, method = "BH", n = n) 
+```
 
+Save
+```{r save food linear regression table}
 # Save table
@@
 ```
+
+Make TMAO food intake by quantile table
+```{r Food Intake by TMAO quantile}
+# TMAO table by age bin - females
+factorVars <- c("age_cat", "tmao_quantile")
+vars <- c("pf_meat_totals", 
+          "pf_curedmeat_totals",
+          "pf_poult_totals",
+          "pf_seafd_hi_totals",
+          "pf_seafd_low_totals",
+          "pf_eggs_totals",
+          "pf_nutsds_totals",
+          "pf_legumes_totals",
+          "pf_total_totals",
+          #"pf_soy_totals_Binary",
+          "v_drkgr_totals",
+          "v_redor_total_totals",
+          #"v_redor_tomato_totals",
+          #"v_redor_other_totals",
+          "v_starchy_total_totals",
+          #"v_starchy_potato_totals",
+          #"v_starchy_other_totals",
+          "v_other_totals",
+          "v_total_totals",
+          #"v_legumes_total",
+          "g_whole_totals",
+          "g_refined_totals",
+          "g_total_totals",
+          "d_total_totals",
+          "d_milk_totals",
+          "d_cheese_totals",
+          "d_yogurt_totals",
+          "hei_asa24_totalscore",
+          "hei_asa24_totalveg",
+          "hei_asa24_sfat",
+          "hei_asa24_",
+          "age",
+          "ethnicity" ,
+          "bmi_final",
+          "tmao",
+          "choline",
+          "betaine",
+          "carnitine")
+table_food.fq <- CreateTableOne(vars = vars,
+                                    strata = "tmao_quantile",
+                                    data = full[full$sex == "Female",],
+                                    factorVars = factorVars)
+table_food.fq
+
+# TMAO table by age bin - males
+table_food.mq <- CreateTableOne(vars = vars,
+                                    strata = "tmao_quantile",
+                                    data = full[full$sex == "Male",],
+                                    factorVars = factorVars)
+table_food.mq
+
+# TMAO table by sex
+table_food.allq <- CreateTableOne(vars = vars,
+                                    strata = "tmao_quantile",
+                                    data = full,
+                                    factorVars = factorVars)
+table_food.allq
+```
+
+```{r Save food quantile tables}
+table_save.fq <- print(table_food.fq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Food_Female_byTMAOQuantile_211111.csv")
+
+table_save.mq <- print(table_food.mq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Food_Male_byTMAOQuantile_211111.csv")
+
+table_save.allq <- print(table_food.allq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOQuantile_211111.csv")
+```
+
+# Miscellaneous for paper
+
+Demonstrate the sex:age interaction for females.
+```{r sex*age interaction}
+summary(lm(tmao_log ~ sex*age + ethnicity + cystatinc_bd1, phen))
+```
+
+Describe meat intake in the cohort
+
+```{r Meat intake}
+# Beef, veal, pork, lamb, and game meat; excludes organ meat and cured meat (oz. eq.)
+summary(full$pf_meat_totals)
+
+# Frankfurters, sausages, corned beef, and luncheon meat that are made from beef, pork, or poultry (oz. eq.)
+summary(full$pf_curedmeat_totals)
+
+# Total meat, poultry, organ meat, cured meat, seafood, eggs, soy, and nuts and seeds; excludes legumes (oz. eq.)
+summary(full$pf_total_totals)
+
+# Total of meat, poultry, seafood, organ meat, and cured meat (oz. eq.)
+summary(full$pf_mps_total_totals)
+
+# Kcal
+summary(full$kcal_tnfs)
+```
+
+```{r}
+full$low_mps_meat <- ifelse(full$pf_mps_total_totals <= median(full$pf_mps_total_totals), "low", "high")
+full$low_mps_meat <- ifelse(full$pf_mps_total_totals <= 9.091, "low", "high")
+summary(lm(tmao_log ~ low_mps_meat + sex*age + cystatinc_bd1 + kcal_tnfs, full))
+```

### 2021-11-16 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update workflow.

@@
 load("../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData") #PSOtmao
+```
+
+```{r Add meta data to phyloseq object}
+# Add age_cat
+sample_data(PSOtmao)$age_cat <- ifelse(sample_data(PSOtmao)$age < 34, 1,
+                         ifelse(sample_data(PSOtmao)$age >= 34 & sample_data(PSOtmao)$age < 50, 2, 3))
+sample_data(PSOtmao)$age_cat <- as.factor(sample_data(PSOtmao)$age_cat)
+# Add bmi_cat
+sample_data(PSOtmao)$bmi_cat <- ifelse(sample_data(PSOtmao)$bmi_final.x < 25, 1,
+                         ifelse(sample_data(PSOtmao)$bmi_final.x >= 25 & sample_data(PSOtmao)$bmi_final.x < 30, 2, 3))
+sample_data(PSOtmao)$bmi_cat <- as.factor(sample_data(PSOtmao)$bmi_cat)
+
+# Make sure "below" is the reference factor
+str(sample_data(PSOtmao)$tmao_mdn)
+sample_data(PSOtmao)$tmao_mdn <- relevel(sample_data(PSOtmao)$tmao_mdn, ref = "below")
+
+# Add TMAO quantiles
+tmao_quantile <- quantile(sample_data(PSOtmao)$tmao)
+sample_data(PSOtmao)$tmao_quantile <- ifelse(sample_data(PSOtmao)$tmao <= tmao_quantile[2], "Quantile1",
+                             ifelse(sample_data(PSOtmao)$tmao > tmao_quantile[2] & sample_data(PSOtmao)$tmao <= tmao_quantile[3], "Quantile2",
+                                   ifelse(sample_data(PSOtmao)$tmao > tmao_quantile[3] & sample_data(PSOtmao)$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))
+
+# Factor it
+sample_data(PSOtmao)$tmao_quantile = factor(sample_data(PSOtmao)$tmao_quantile)
 ```
@@
 # Agglomerate taxa
-PSOtmao4 = tax_glom(PSOtmao3, "Genus", NArm = TRUE)
+PSOtmao3G = tax_glom(PSOtmao3, "Genus", NArm = TRUE)
 ```
@@
 PSOtmao3
-PSOtmao4
+PSOtmao3G
 ```
@@
 # Prevalence - how many samples the taxa is observed in
-prevdf = apply(X = otu_table(PSOtmao4),
-               MARGIN = ifelse(taxa_are_rows(PSOtmao4), yes = 1, no = 2),
+prevdf = apply(X = otu_table(PSOtmao3G),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao3G), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})
@@
 prevdf = data.frame(Prevalence = prevdf,
-                    TotalAbundance = taxa_sums(PSOtmao4), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
-                    tax_table(PSOtmao4))
+                    TotalAbundance = taxa_sums(PSOtmao3G), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao3G))
 
@@
 # Family
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao4),color=Family)) +
+ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao3G),color=Family)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
@@
 # Genus - have to blow it up to make it useful
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao4),color=Genus)) +
+ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao3G),color=Genus)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
@@
 ```{r PCoA for outliers}
-PCoA_BC <- ordinate(PSOtmao4, "PCoA", "bray")
+PCoA_BC <- ordinate(PSOtmao3G, "PCoA", "bray")
 
@@
 # taxa is similar to species
-taxa_plot <- plot_ordination(PSOtmao4, PCoA_BC, type = "taxa")
+taxa_plot <- plot_ordination(PSOtmao3G, PCoA_BC, type = "taxa")
 taxa_plot
@@
 # Subset 
-PSOtmao5 = subset_taxa(PSOtmao4, !Genus %in% filterGenus)
+PSOtmao4G = subset_taxa(PSOtmao3G, !Genus %in% filterGenus)
 
 # Replot:
-PCoA_BC2 <- ordinate(PSOtmao5, "PCoA", "bray")
+PCoA_BC2 <- ordinate(PSOtmao4G, "PCoA", "bray")
 
@@
 # taxa is similar to species
-taxa_plot2 <- plot_ordination(PSOtmao5, PCoA_BC2, type = "taxa")
+taxa_plot2 <- plot_ordination(PSOtmao4G, PCoA_BC2, type = "taxa")
 taxa_plot2
@@
 # samples is similar to sites
-plot_ordination(PSOtmao5, PCoA_BC2, type = "samples", color = "age")
+plot_ordination(PSOtmao4G, PCoA_BC2, type = "samples", color = "age")
 
 # Which taxa have less than 1000 counts?
-which(!rowSums(otu_table(PSOtmao5)) > 1000) # none, good
+which(!rowSums(otu_table(PSOtmao4G)) > 1000) # none, good
 
@@
 PSOtmao3
-PSOtmao4
-PSOtmao5 
+PSOtmao3G
+PSOtmao4G
 ```
 
-Add age and bmi categories to the "final" phyloseq object. This will come in handy when building the DESeq2 analysis and downstream plotting.
+Save
 
-```{r Add meta data to phyloseq object}
-# Add age_cat
-sample_data(PSOtmao5)$age_cat <- ifelse(sample_data(PSOtmao5)$age < 34, 1,
-                         ifelse(sample_data(PSOtmao5)$age >= 34 & sample_data(PSOtmao5)$age < 50, 2, 3))
-sample_data(PSOtmao5)$age_cat <- as.factor(sample_data(PSOtmao5)$age_cat)
-# Add bmi_cat
-sample_data(PSOtmao5)$bmi_cat <- ifelse(sample_data(PSOtmao5)$bmi_final.x < 25, 1,
-                         ifelse(sample_data(PSOtmao5)$bmi_final.x >= 25 & sample_data(PSOtmao5)$bmi_final.x < 30, 2, 3))
-sample_data(PSOtmao5)$bmi_cat <- as.factor(sample_data(PSOtmao5)$bmi_cat)
+```{r save PSOtmao5}
+save(PSOtmao4G, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao5_211111.RData")
+```
 
-# Make sure "below" is the reference factor
-str(sample_data(PSOtmao5)$tmao_mdn)
-sample_data(PSOtmao5)$tmao_mdn <- relevel(sample_data(PSOtmao5)$tmao_mdn, ref = "below")
+# Phylogenetic Abundance plots
+
+```{r Phylo Abundance Plot by TMAO Quantile}
+# Agglomerate taxa
+PSOtmao3F = tax_glom(PSOtmao3, "Family", NArm = TRUE)
+PSOtmao3F.top10 <- microbiome::aggregate_top_taxa(PSOtmao3, "Family", top = 10)
+
+ps3F.10.family.comp <- microbiome::transform(PSOtmao3F.top10, transform="compositional")
+
+plot.composition.relAbun <- plot_composition(ps3F.10.family.comp,
+                                             sample.sort = "Description",
+                                             x.label = "tmao_quantile",
+                                             group_by = "tmao_quantile",
+                                             average_by = "tmao_quantile") +
+  theme_minimal() +
+  scale_fill_brewer("Family", palette = "Paired") + 
+  ggtitle("Relative Abundance at Family Level")
+
+plot.composition.relAbun
+
 ```
@@
 # Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
-sd <- sample_data(PSOtmao5)
+sd <- sample_data(PSOtmao4G)
 str(sd$sex)
@@
 # Factor sex, BMI okay
-sample_data(PSOtmao5)$sex = factor(sample_data(PSOtmao5)$sex)
+sample_data(PSOtmao4G)$sex = factor(sample_data(PSOtmao4G)$sex)
 
 # DESeq analysis
-mb <- phyloseq_to_deseq2(PSOtmao5, ~ sex:age + bmi_final.x + tmao_mdn)
+mb <- phyloseq_to_deseq2(PSOtmao4G, ~ sex:age + bmi_final.x + tmao_mdn)
 #mb2 <- estimateSizeFactors(mb, type="poscount")
@@
 sigtab = res[which(res$padj < alpha), ]
-sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao5)[rownames(sigtab), ], "matrix"))
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao4G)[rownames(sigtab), ], "matrix"))
 sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
@@
 We need to recalculate the alpha diversity metrics now that we have removed and glommed the taxa data.
+Note, by recalculating these metrics we don't want to use measures that count singletons (Chao1, ACE) because we purposefully removed those occurences!
+By redoing the metrics we are asking, "At the Genus level, is within diversity (richness, evenness, or a combination depending on the statistical test) different?"
 
@@
 # Recalculate alpha diversity now that some taxonomy have been filtered out
-updated_alpha <- estimate_richness(PSOtmao4)
+updated_alpha <- estimate_richness(PSOtmao4G)
 row.names(updated_alpha) <- gsub("X", "", row.names(updated_alpha))
@@
 # Grab PSO sample data
-df <- as(sample_data(PSOtmao5), "data.frame")
+df <- as(sample_data(PSOtmao4G), "data.frame")
 
@@
 # Original alpha
-summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df2))
+summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1, df2))
 # Updated alpha
-summary(lm(df_alpha2$Shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
+summary(lm(df_alpha2$Shannon ~ tmao_log + sex*age + cystatinc_bd1, df_alpha2))
 ```
@@
 #~~~~~~~~~~~
-# Chao1
+# Fisher
 #~~~~~~~~~~~
 # Check distribution
-shapiro.test(df_alpha2$Chao1) # ehh not normal
+shapiro.test(df_alpha2$Fisher) # 
 # Run linear regression
-summary(lm(Chao1 ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
+summary(lm(Fisher ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
 
@@
 # PCoA
-PCoA_WU <- ordinate(PSOtmao5, "PCoA", "wunifrac")
-PCoA_UnWU <- ordinate(PSOtmao5, "PCoA", "uunifrac")
-PCoA_BC <- ordinate(PSOtmao5, "PCoA", "bray")
+PCoA_WU <- ordinate(PSOtmao4G, "PCoA", "wunifrac")
+PCoA_UnWU <- ordinate(PSOtmao4G, "PCoA", "uunifrac")
+PCoA_BC <- ordinate(PSOtmao4G, "PCoA", "bray")
 
 # Calculate distance matrix
-dismax_wu = phyloseq::distance(PSOtmao5, method="wunifrac", type = "samples")
-dismax_uwu = phyloseq::distance(PSOtmao5, method="unifrac", type = "samples")
-dismax_bc = phyloseq::distance(PSOtmao5, method="bray", type = "samples")
+dismax_wu = phyloseq::distance(PSOtmao4G, method="wunifrac", type = "samples")
+dismax_uwu = phyloseq::distance(PSOtmao4G, method="unifrac", type = "samples")
+dismax_bc = phyloseq::distance(PSOtmao4G, method="bray", type = "samples")
 
@@
 
-```{r}
+```{r ANCOM - genus level}
 source("../../Rscripts/ancom_v2.1.R")
@@
 # OTU table
-MicroData <- otu_table(PSOtmao5)
+MicroData <- otu_table(PSOtmao4G)
 MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
@@
 # OTU data
-otu_data <- otu_table(PSOtmao5)
+otu_data <- otu_table(PSOtmao4G)
 otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
@@
 # Subset
-MetaData2 <- df_alpha2 %>% dplyr::select("tmao_mdn")
+MetaData2 <- df_alpha2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
 MetaData2$Sample.ID <- rownames(MetaData2)
@@
 # Step 2: ANCOM
-feature_table = prepro$feature_table # preprocessing went from 98 to 60 bugs
+feature_table = prepro$feature_table 
 meta_data = prepro$meta_data
@@
 main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
-adj_formula = NULL; rand_formula = NULL
+adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
 # Run ANCCOM

### 2021-11-17 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update microbiome analysis.

@@
 
-```{r save PSOtmao5}
-save(PSOtmao4G, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao5_211111.RData")
+```{r save PSOtmao4G}
+save(PSOtmao4G, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao4G_211111.RData")
 ```
@@
 
-plot.composition.relAbun <- plot_composition(ps3F.10.family.comp,
+plot.composition.relAbun <- microbiome::plot_composition(ps3F.10.family.comp,
                                              sample.sort = "Description",
@@
   geom_point(size=3) + 
-  xlab("Taxonomy") +
-  ylab("Log2 Fold Change") + 
+  ylab("Taxonomy") +
+  xlab("Log2 Fold Change") + 
   scale_color_brewer(palette = "Dark2") +
@@
 # Get sample_data
-df <- as(sample_data(PSOtmao5), "data.frame")
+df <- as(sample_data(PSOtmao4G), "data.frame")
 df$tmao_mdn <- relevel(df$tmao_mdn, ref = "below")
@@
 
+## ANCOM2 no-covariate analysis
+
 ```{r ANCOM - genus level}
+source("../../Rscripts/ancom_v2.1.R")
+
+# OTU table
+MicroData <- otu_table(PSOtmao4G)
+MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
+# Add 1
+#MicroData2 <- MicroData +1
+#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
+#dim(MicroData2) # 103 x 98
+
+# Load the phenotype data
+#------------------------
+# Perform ANCOM analysis
+# Step 1: Data preprocessing
+# OTU data
+otu_data <- otu_table(PSOtmao4G)
+otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
+otu_id <- rownames(otu_data)
+
+# Metadata 
+#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
+#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
+# Subset
+MetaData2 <- df_alpha2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
+MetaData2$Sample.ID <- rownames(MetaData2)
+
+feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
+out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
+# Function
+prepro = feature_table_pre_process(feature_table = feature_table,
+                                   meta_data = MetaData2,
+                                   sample_var = sample_var,
+                                   group_var = NULL,
+                                   out_cut = out_cut,
+                                   zero_cut = zero_cut,
+                                   lib_cut = lib_cut,
+                                   neg_lb = neg_lb)
+
+# Step 2: ANCOM
+feature_table = prepro$feature_table 
+meta_data = prepro$meta_data
+struc_zero = prepro$structure_zeros
+
+# Define variables for ANCOM
+main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
+adj_formula = NULL ; rand_formula = NULL
+# Run ANCCOM
+t_start = Sys.time()
+res = ANCOM(feature_table = feature_table, 
+            meta_data = meta_data, 
+            struc_zero = struc_zero, 
+            main_var = main_var, 
+            p_adj_method = p_adj_method, 
+            alpha = alpha, 
+            adj_formula = adj_formula, 
+            rand_formula = rand_formula)
+t_end = Sys.time()
+t_run = t_end - t_start 
+t_run # around 5s
+
+# Results
+res$out
+res$fig$data
+
+# Step 3: Volcano Plot
+# Number of taxa except structural zeros
+n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
+# Cutoff values for declaring differentially abundant taxa
+cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
+names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")
+
+# Annotation data
+dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")
+
+fig = res$fig +  
+  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
+  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
+            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
+fig
+
+# ORGANIZE PLOT
+
+TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
+rownames(TaxaData2) <- TaxaData2$Feature.ID
+
+ancomRes <- res$out
+
+test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)
+
+# Make plot 
+tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
+asv_id <- unique(tempdata$taxa_id)
+
+MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]
+
+# Rename to the taxonomy 
+#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
+TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
+# Make a unique naming column that incorporates multiple levels of taxonomy
+TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
+rownames(TaxaData3) <- rownames(TaxaData2)
+# Merge
+MicroData5 <- merge(MicroData4, TaxaData3,
+                    by=0, all.x = T, all.y = F)
+
+row.names(MicroData5) <- MicroData5$OrderFamilyGenus
+
+MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
+MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))
+
+# Merge with the phenotype data 
+MetaData3 <- MetaData2 %>% dplyr::select(tmao_mdn)
+
+ggData <- merge(MetaData3, MicroData5, by =0)
+row.names(ggData) <- ggData$Row.names
+ggData <- ggData %>% dplyr::select(-Row.names)
+
+# Gather the data
+library(tidyr)
+ggData2 <- gather(ggData,
+                  key = "Taxa",
+                  value = "Abundance",
+                  -tmao_mdn)
+
+ggData2$Abundance <- log(ggData2$Abundance)
+
+# Plot it
+p <- ggplot(data = ggData2,
+            aes(x = Taxa, 
+                y = Abundance,
+                fill = tmao_mdn), 
+            alpha = 0.1) +
+  geom_boxplot(lwd=.25) +
+  xlab(NULL) +
+  ylab("Log of Taxa Abundance") +
+  theme_bw() + # note order matters here - need bw to be before theme()
+  theme(axis.text.y = element_text( hjust = 1),
+        axis.text.x = element_text(angle = 60, hjust = 1),
+        text=element_text(size=11))  +
+  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
+  #scale_fill_viridis(discrete = TRUE) +
+  #coord_flip() +
+  labs(fill = "TMAO \n Classification")
+
+p
+```
+
+## ANCOM2 covariate analysis
+ANCOM2 lets you include covariates into the analysis. If we add the same covariates as in the DESeq2 analysis, do we get similar results?
+
+```{r ANCOM - genus level with covariates}
 source("../../Rscripts/ancom_v2.1.R")

### 2021-11-18 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update the filtering per Dr. K's advice.

@@
 
-1) TMAO phyloseq object
+1) TMAO phyloseq object.
+Recall, the data from Dr.
+Kable was filtered to taxa with 2% abundance, which removes technical noise.
+Therefore, the remaining data is believable.
+Further filtering we do will be driven by our interest in more prevalent taxa.
+Furthermore, we should not recalculate the alpha diversity statistics after our filtering because that would incorrectly describe the microbiome environment.
 2) Master phenotype data
@@
 sample_data(PSOtmao)$tmao_quantile = factor(sample_data(PSOtmao)$tmao_quantile)
+
+# Add TMAO tertiles
+tmao_tertile <- quantile(sample_data(PSOtmao)$tmao, c(0:3/3))
+sample_data(PSOtmao)$tmao_tertile <- ifelse(sample_data(PSOtmao)$tmao <= tmao_tertile[2], "Tertile1",
+                             ifelse(sample_data(PSOtmao)$tmao > tmao_tertile[2] & sample_data(PSOtmao)$tmao <= tmao_tertile[3], "Tertile2", "Tertile3"))
+# Factor it
+sample_data(PSOtmao)$tmao_tertile = factor(sample_data(PSOtmao)$tmao_tertile)
+# Lookss good
+table(sample_data(PSOtmao)$tmao_tertile)
 ```
@@
 We will filter low abundant Phyla.
-We will also require that taxa are present in at least 20% of samples (at least 71 samples) and occur at least 1000 times.
+We will also require that taxa are present in at least 10% of samples (at least 36 samples).
 We will then agglomerate at the Genus level.
 
-```{r Tables to Guide Supervised Filtering}
+```{r Getting Familiar with Taxa Distributions}
 print("The number of different bacteria at each phyla:")
@@
 From the phylum table, we can see that p__Elusimicrobia, p__Spirochaetes, and p__Synergistetes have the same mean and sum, meaning they are only present in 1 sample.
-We are interested and confident in the more abundant taxa so manually filterr them out.
+We are interested and confident in the more abundant taxa so manually filter them out.
+Per Dr.
+Kable's expert advice, she already filtered to 2% prevalence to remove technical error.
+We will increase the filtering to 5%, a common threshold in the literature.
 
-```{r Supervised Filtering}
-# Phylum that only occur 1 time may be artifcats. We can remove them.
-filterPhyla <- c(" p__Elusimicrobia", " p__Spirochaetes", " p__Synergistetes") # note, our taxa table has these awful hidden spaces!
+```{r Supervised Filtering 5%}
+# Define prevalence threshold as 5% of total samples
+prevalenceThreshold <- 0.05 * nsamples(PSOtmao)
+cat("The 10% prevalence threshold is", prevalenceThreshold, "samples.", "\n")
 
-# Subset
-PSOtmao1 = subset_taxa(PSOtmao, !Phylum %in% filterPhyla)
+# Execute prevalence filter, using `prune_taxa()` function
+keepTaxa = rownames(prevdf)[(prevdf$Prevalence >= prevalenceThreshold)]
+PSOtmao1 = prune_taxa(keepTaxa, PSOtmao)
 
-# View
-cat("The Phyla present after this step of subsetting includes:", "\n")
+print("The number of different bacteria at each phyla:")
 table(tax_table(PSOtmao1)[, "Phylum"])
 
-# Subset to the remaining phyla
-prevdf1 = subset(prevdf, Phylum %in% get_taxa_unique(PSOtmao1, "Phylum"))
-
 # Look at phylum level distributions
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1),color=Phylum)) +
+ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1),color=Phylum)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) +  # Include a guess for parameter
@@
 
-# Define prevalence threshold as 20% of total samples
-prevalenceThreshold = 0.20 * nsamples(PSOtmao1)
-cat("The 20% prevalence threshold is", prevalenceThreshold, "samples.", "\n")
-
-# Execute prevalence filter, using `prune_taxa()` function
-keepTaxa = rownames(prevdf1)[(prevdf1$Prevalence >= prevalenceThreshold)]
-PSOtmao2 = prune_taxa(keepTaxa, PSOtmao1)
-
-cat("Which taxa have less than 1000 counts accross all samples?", "\n")
-which(!rowSums(otu_table(PSOtmao2)) > 1000) # 6 OTUS
-# Remove these to see if that cleans ordination
-PSOtmao3 <- prune_taxa(rowSums(otu_table(PSOtmao2)) >= 1000, PSOtmao2)
-
 # Agglomerate taxa
-PSOtmao3G = tax_glom(PSOtmao3, "Genus", NArm = TRUE)
+PSOtmao1G = tax_glom(PSOtmao1, "Genus", NArm = TRUE)
 ```
@@
 # Inspect the effect of each filtering step on the phyloseq object
+PSOtmao
 PSOtmao1
-PSOtmao2
-PSOtmao3
-PSOtmao3G
+PSOtmao1G
 ```
@@
 # Prevalence - how many samples the taxa is observed in
-prevdf = apply(X = otu_table(PSOtmao3G),
-               MARGIN = ifelse(taxa_are_rows(PSOtmao3G), yes = 1, no = 2),
+prevdf = apply(X = otu_table(PSOtmao1G),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1G), yes = 1, no = 2),
                FUN = function(x){sum(x > 0)})
@@
 prevdf = data.frame(Prevalence = prevdf,
-                    TotalAbundance = taxa_sums(PSOtmao3G), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
-                    tax_table(PSOtmao3G))
+                    TotalAbundance = taxa_sums(PSOtmao1G), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1G))
 
@@
 # Family
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao3G),color=Family)) +
+ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Family)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
@@
 # Genus - have to blow it up to make it useful
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao3G),color=Genus)) +
+ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Genus)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
@@
 ```{r PCoA for outliers}
-PCoA_BC <- ordinate(PSOtmao3G, "PCoA", "bray")
+PCoA_BC <- ordinate(PSOtmao1G, "PCoA", "bray")
 
@@
 # taxa is similar to species
-taxa_plot <- plot_ordination(PSOtmao3G, PCoA_BC, type = "taxa")
+taxa_plot <- plot_ordination(PSOtmao1G, PCoA_BC, type = "taxa")
 taxa_plot
@@
 cat("Which genus is the outlier?", outlier$Genus, "\n")
-
-# Remove it
-filterGenus <- c(" g__Prevotella") # note, our taxa table has these awful hidden spaces!
-
-# Subset 
-PSOtmao4G = subset_taxa(PSOtmao3G, !Genus %in% filterGenus)
-
-# Replot:
-PCoA_BC2 <- ordinate(PSOtmao4G, "PCoA", "bray")
-
-# BC Round 2
-# taxa is similar to species
-taxa_plot2 <- plot_ordination(PSOtmao4G, PCoA_BC2, type = "taxa")
-taxa_plot2
-
-# samples is similar to sites
-plot_ordination(PSOtmao4G, PCoA_BC2, type = "samples", color = "age")
-
-# Which taxa have less than 1000 counts?
-which(!rowSums(otu_table(PSOtmao4G)) > 1000) # none, good
-
 ```
 
-View the phyloseq objects again. We see that the most dramatic effect was when we filtered for phyla present in 20% of samples (PSOtmao1 to 2).
-
-```{r Compare PSOs again}
-PSOtmao
-PSOtmao1
-PSOtmao2
-PSOtmao3
-PSOtmao3G
-PSOtmao4G
-```
+Super interesting!
+The "outlier" is prevotella, which is interesting in the microbiome community as a marker of individuals who consume/live/have non-industrialized diets/environments/guts.
+We won't remove this bug afterall.
 
@@
 ```{r save PSOtmao4G}
-save(PSOtmao4G, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao4G_211111.RData")
+save(PSOtmao1, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao1_211118.RData")
+save(PSOtmao1G, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao1G_211118.RData")
 ```
@@
 
+## Family Plot of TMAO Quantile
+
 ```{r Phylo Abundance Plot by TMAO Quantile}
 # Agglomerate taxa
-PSOtmao3F = tax_glom(PSOtmao3, "Family", NArm = TRUE)
-PSOtmao3F.top10 <- microbiome::aggregate_top_taxa(PSOtmao3, "Family", top = 10)
+PSOtmao1F = tax_glom(PSOtmao1, "Family", NArm = TRUE)
+PSOtmao1F.top10 <- microbiome::aggregate_top_taxa(PSOtmao1F, "Family", top = 10)
 
-ps3F.10.family.comp <- microbiome::transform(PSOtmao3F.top10, transform="compositional")
+ps1F.10.family.comp <- microbiome::transform(PSOtmao1F.top10, transform="compositional")
 
-plot.composition.relAbun <- microbiome::plot_composition(ps3F.10.family.comp,
+plot.composition.relAbun <- microbiome::plot_composition(ps1F.10.family.comp,
                                              sample.sort = "Description",
@@
 plot.composition.relAbun
+```
 
+## Family Plot of TMAO Tertile
+
+```{r Tertiles}
+# Agglomerate taxa
+PSOtmao1F = tax_glom(PSOtmao1, "Family", NArm = TRUE)
+PSOtmao1F.top10 <- microbiome::aggregate_top_taxa(PSOtmao1F, "Family", top = 10)
+ps1F.10.family.comp <- microbiome::transform(PSOtmao1F.top10, transform="compositional")
+
+plot.composition.relAbun <- microbiome::plot_composition(ps1F.10.family.comp,
+                                             sample.sort = "Description",
+                                             x.label = "tmao_tertile",
+                                             group_by = "tmao_tertile",
+                                             average_by = "tmao_tertile") +
+  theme_minimal() +
+  scale_fill_brewer("Family", palette = "Paired") + 
+  ggtitle("Relative Abundance at Family Level")
+
+plot.composition.relAbun
+
+# Save
+jpeg("../../Outputs/plots/microbiome/RelAbund_Fam_TMAOTertiles_2111.jpg", width = 4, height = 4, res = 300)
+plot.composition.relAbun
+dev.off()
 ```

### 2021-11-23 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Updates to workflow post meeting with collab. DESeq update and use original alpha div.

@@
 2) [Workflow for Microbiome Data Analysis: from raw reads to community analyses](https://bioconductor.org/help/course-materials/2017/BioC2017/Day1/Workshops/Microbiome/MicrobiomeWorkflowII.html#different_ordination_projections)
+3) [Example of what descriptive microbiome stats to report in the results](https://pubmed.ncbi.nlm.nih.gov/28407784/)
 
@@
 
-```{r DESeq2 analysis}
+From Dr. Kable's example,
+
+> # Test the impact of fiber cluster, accounting for covariates, using the log ratio test
+dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
+>
+> # This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero.
+dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
+>
+> # The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
+resultsNames(dds.ratio.batcho)
+res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))
+
+## Factored by TMAO Median 
+```{r DESeq2 analysis - TMAO median}
 # Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
-sd <- sample_data(PSOtmao4G)
+sd <- sample_data(PSOtmao1G)
 str(sd$sex)
@@
 # Factor sex, BMI okay
-sample_data(PSOtmao4G)$sex = factor(sample_data(PSOtmao4G)$sex)
+sample_data(PSOtmao1G)$sex = factor(sample_data(PSOtmao1G)$sex)
 
+#~~~~~~~~~~~~~~~
 # DESeq analysis
-mb <- phyloseq_to_deseq2(PSOtmao4G, ~ sex:age + bmi_final.x + tmao_mdn)
-#mb2 <- estimateSizeFactors(mb, type="poscount")
-mb3 = DESeq(mb, test="Wald", fitType="parametric")
-
-# Results
-res = results(mb3, cooksCutoff = FALSE)
+# TMAO median
+#~~~~~~~~~~~~~~~
+mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_mdn)
+mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
+res.ratio <- results(mb.ratio)
 alpha = 0.05
-sigtab = res[which(res$padj < alpha), ]
-sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao4G)[rownames(sigtab), ], "matrix"))
+alpha.big = 0.5
+sigtab = res.ratio[which(res.ratio$padj < alpha.big), ]
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
 sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
 head(sigtab)
+
 ```
 
-```{r Plot DESeq2 results}
+No results when we look at TMAO median and control for age, sex, and BMI.
+
+## Factored by TMAO Tertile
+
+```{r DESeq2 analysis - TMAO tertile}
+
+# DESeq analysis - tmao_tertile
+mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_tertile)
+mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
+res.ratio <- results(mb.ratio, contrast = c("tmao_tertile", "Tertile1", "Tertile3")) # Look at most different tertiles
+alpha = 0.05
+sigtab = res.ratio[which(res.ratio$padj < alpha), ]
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
+sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
+head(sigtab)
+
+#~~~~~~~~~~~~~~~
+# Plot Results
+#~~~~~~~~~~~~~~~
+resOTU <- rownames(sigtab)
+otuData <- as.data.frame(otu_table(PSOtmao1G))
+otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
+#otuData_desRes <- as.data.frame(t(otuData_desRes))
+# Get sample data groupings
+sampleData <- as.data.frame(sample_data(PSOtmao1G))
+sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
+# Get tax table
+taxaData <- as.data.frame(tax_table(PSOtmao1G))
+taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
+taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
+
+# Merge
+ggData <- merge(taxaData_res, otuData_desRes, by = 0)
+rownames(ggData) <-  ggData$OrderFamilyGenus
+ggData <- ggData[,-(1:9)]
+ggData <- as.data.frame(t(ggData))
+ggData <- merge(ggData, sampleData, by = 0)
+rownames(ggData) <- ggData[,"Row.names"]
+ggData <- subset(ggData, select = -c(Row.names))
+
+# Loop Plot
+nres <- length(resOTU)
+for (i in 1:nres) { 
+    print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+            geom_boxplot()+
+            ylab(colnames(ggData)[i])) 
+}
+
+#~~~~~~~~~~~~~~~
+# Plot Results Fold Change
+#~~~~~~~~~~~~~~~
+sigtabgen = subset(sigtab, !is.na(Genus))
+# Phylum order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
+# Genus order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
+ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
+  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
+  geom_point(size=3) + 
+  ylab("Taxonomy") +
+  xlab("Log2 Fold Change") + 
+  scale_color_brewer(palette = "Dark2") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+```
+
+## Factored by TMAO Quantile
+
+```{r DESeq2 analysis - TMAO quantile}
+# DESeq analysis - tmao_quantile
+mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_quantile)
+mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
+res.ratio <- results(mb.ratio, contrast = c("tmao_quantile", "Quantile1", "Quantile4"))
+alpha = 0.05
+sigtab = res.ratio[which(res.ratio$padj < alpha), ]
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
+sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
+head(sigtab)
+
+#~~~~~~~~~~~~~~~
+# Plot Results
+#~~~~~~~~~~~~~~~
+resOTU <- rownames(sigtab)
+otuData <- as.data.frame(otu_table(PSOtmao1G))
+otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
+#otuData_desRes <- as.data.frame(t(otuData_desRes))
+# Get sample data groupings
+sampleData <- as.data.frame(sample_data(PSOtmao1G))
+sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
+# Get tax table
+taxaData <- as.data.frame(tax_table(PSOtmao1G))
+taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
+taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
+
+# Merge
+ggData <- merge(taxaData_res, otuData_desRes, by = 0)
+rownames(ggData) <-  ggData$OrderFamilyGenus
+ggData <- ggData[,-(1:9)]
+ggData <- as.data.frame(t(ggData))
+ggData <- merge(ggData, sampleData, by = 0)
+rownames(ggData) <- ggData[,"Row.names"]
+ggData <- subset(ggData, select = -c(Row.names))
+
+# Loop Plot
+nres <- length(resOTU)
+for (i in 1:nres) { 
+    print(ggplot(ggData, aes(x = tmao_quantile, y = log(ggData[,i]))) +
+            geom_boxplot()+
+            ylab(colnames(ggData)[i])) 
+}
+
+#~~~~~~~~~~~~~~~
+# Plot Results Fold Change
+#~~~~~~~~~~~~~~~
 sigtabgen = subset(sigtab, !is.na(Genus))
@@
 
-We need to recalculate the alpha diversity metrics now that we have removed and glommed the taxa data.
-Note, by recalculating these metrics we don't want to use measures that count singletons (Chao1, ACE) because we purposefully removed those occurences!
-By redoing the metrics we are asking, "At the Genus level, is within diversity (richness, evenness, or a combination depending on the statistical test) different?"
+We do not need to redo the alpha diversity calculations because the filtering we did was not to remove technical variation but instead to focus our analysis on the more common players (also so we can be more statistically confident using less scarce data).
+Note, if you did want to calculate alpha diversity, the command estimate_richness() can accomplish that for you.
 
-```{r Update alpha statistics}
-#plot_richness(PSOtmao4, x="tmao_mdn", color="bmi_cat", measures=c("Chao1", "Shannon"))
+Check the distributions of the alpha diversity metrics and run the multiple linear regression models with covariates.
 
-# Recalculate alpha diversity now that some taxonomy have been filtered out
-updated_alpha <- estimate_richness(PSOtmao4G)
-row.names(updated_alpha) <- gsub("X", "", row.names(updated_alpha))
-```
-
-First, load the master phenotype file so that we can have complete access to our covariate variables. Merge the master phenotyp file with the sample data from the phyloseq object.
-
-```{r Alpha diversity load data}
+```{r Original Alpha diversity stats}
 # Load master pheno file
@@
 
-# Check bin structure and factor if needed
-#str(phen$age_cat)
-#phen$age_cat <- as.factor(phen$age_cat)
-#str(phen$bmi_cat)phen$bmi_cat <- as.factor(phen$bmi_cat)
-
 # Grab PSO sample data
-df <- as(sample_data(PSOtmao4G), "data.frame")
-
+df <- as(sample_data(PSOtmao1G), "data.frame")
 # Merge
@@
 
-# Merge updated alpha
-df_alpha2 <- merge(updated_alpha, df2, by = 0)
-row.names(df_alpha2) <- df_alpha2[,"Row.names"]
-df_alpha2 <- subset(df_alpha2, select =-c(Row.names))
+#~~~~~~~~~~~~~~~~
+# Compute
+#~~~~~~~~~~~~~~~~
 
-# Original alpha
-summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1, df2))
-# Updated alpha
-summary(lm(df_alpha2$Shannon ~ tmao_log + sex*age + cystatinc_bd1, df_alpha2))
-```
-
-Check the distributions of the alpha diversity metrics and run the multiple linear regression models.
-
-```{r Original Alpha diversity stats}
 # Check distribution
@@
 # Run linear regression
-summary(lm(shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
+summary(lm(tmao_log ~ shannon + sex*age + cystatinc_bd1, df2))
 
@@
 # Run linear regression
-summary(lm(faith_pd ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
+summary(lm(tmao_log ~ faith_pd + sex*age + cystatinc_bd1, df2))
 
@@
 # Run linear regression
-summary(lm(pielou_e ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
+summary(lm(tmao_log ~ pielou_e + sex*age + cystatinc_bd1, df2))
 
@@
 # Run linear regression
-summary(lm(observed_otus ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
-```
-
-```{r Updated Alpha diversity stats}
-#~~~~~~~~~~~
-# Shannon
-#~~~~~~~~~~~
-# Check distribution
-shapiro.test(df_alpha2$Shannon) # not normal, proceed with caution
-shapiro.test((df_alpha2$shannon)^.5) # not better
-# Run linear regression
-summary(lm(Shannon ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
-
-#~~~~~~~~~~~
-# Fisher
-#~~~~~~~~~~~
-# Check distribution
-shapiro.test(df_alpha2$Fisher) # 
-# Run linear regression
-summary(lm(Fisher ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
-
-#~~~~~~~~~~~
-# InvSimpson
-#~~~~~~~~~~~
-# Check distribution
-shapiro.test(df_alpha2$InvSimpson) # normal
-# Run linear regression
-summary(lm(InvSimpson ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
-
-#~~~~~~~~~~~
-# Observed
-#~~~~~~~~~~~
-# Check distribution
-shapiro.test(df_alpha2$Observed) # eh not normal
-# Run linear regression
-summary(lm(Observed ~ tmao_log + sex*age + cystatinc_bd1 + ethnicity, df_alpha2))
+summary(lm(tmao_log ~ observed_otus + sex*age + cystatinc_bd1, df2))
 ```
@@
 x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
+mdn_cut <- log(median(df2$tmao.x))
 # Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=shannon)) +
+ggplot(df2, aes(x=tmao_log, y=shannon)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
 x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
+mdn_cut <- log(median(df2$tmao.x))
 # Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=faith_pd)) +
+ggplot(df2, aes(x=tmao_log, y=faith_pd)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
 x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
+mdn_cut <- log(median(df2$tmao.x))
 # Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=pielou_e)) +
+ggplot(df2, aes(x=tmao_log, y=pielou_e)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
 x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
+mdn_cut <- log(median(df2$tmao.x))
 # Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=observed_otus)) +
-  geom_point(aes(shape=bmi_cat, color = age_cat)) +
-  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
-  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
-  theme_minimal() + 
-  ylab(y_variable) +
-  xlab(x_variable) +
-  labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
-
-```
-
-```{r Updated Alpha Diversity Plots}
-# Shannon
-y_variable <- "Shannon"
-x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
-# Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=Shannon)) +
-  geom_point(aes(shape=bmi_cat, color = age_cat)) +
-  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
-  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
-  theme_minimal() + 
-  ylab(y_variable) +
-  xlab(x_variable) +
-  labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
-
-# Faith's Phylogenetic Diversity
-y_variable <- "Chao1"
-x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
-# Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=Chao1)) +
-  geom_point(aes(shape=bmi_cat, color = age_cat)) +
-  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
-  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
-  theme_minimal() + 
-  ylab(y_variable) +
-  xlab(x_variable) +
-  labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
-
-# Pielou's Evenness
-y_variable <- "Inverse Simpson"
-x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
-# Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=InvSimpson)) +
-  geom_point(aes(shape=bmi_cat, color = age_cat)) +
-  geom_vline(xintercept = mdn_cut, linetype = "dashed") +
-  geom_smooth(method = "lm", se = FALSE, fullrange = F) +
-  theme_minimal() + 
-  ylab(y_variable) +
-  xlab(x_variable) +
-  labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
-
-# Observed OTUs
-y_variable <- "Observed OTUs"
-x_variable <- "Natural Log of TMAO (uM)"
-mdn_cut <- log(median(df_alpha2$tmao.x))
-# Plot
-ggplot(df_alpha2, aes(x=tmao_log, y=Observed)) +
+ggplot(df2, aes(x=tmao_log, y=observed_otus)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
 # OTU table
-MicroData <- otu_table(PSOtmao4G)
+MicroData <- otu_table(PSOtmao1G)
 MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
@@
 # OTU data
-otu_data <- otu_table(PSOtmao4G)
+otu_data <- otu_table(PSOtmao1G)
 otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
@@
 # Subset
-MetaData2 <- df_alpha2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
+MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
 MetaData2$Sample.ID <- rownames(MetaData2)
@@
 ## ANCOM2 covariate analysis
+
 ANCOM2 lets you include covariates into the analysis. If we add the same covariates as in the DESeq2 analysis, do we get similar results?
 
-```{r ANCOM - genus level with covariates}
+```{r ANCOM - genus level with covariates median}
 source("../../Rscripts/ancom_v2.1.R")
@@
 ```
+
+Repeat looking at tertile 1 and 3. This section isn't currently working but I'm not sure why.
+
+```{r ANCOM - genus level with covariates tertile}
+source("../../Rscripts/ancom_v2.1.R")
+
+# OTU table
+MicroData <- otu_table(PSOtmao1G)
+MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
+# Add 1
+#MicroData2 <- MicroData +1
+#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
+#dim(MicroData2) # 103 x 98
+
+# Load the phenotype data
+#------------------------
+# Perform ANCOM analysis
+# Step 1: Data preprocessing
+# OTU data
+otu_data <- otu_table(PSOtmao1G)
+otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
+otu_id <- rownames(otu_data)
+
+# Metadata 
+#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
+#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
+# Subset
+MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
+MetaData2$Sample.ID <- rownames(MetaData2)
+
+feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
+out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
+# Function
+prepro = feature_table_pre_process(feature_table = feature_table,
+                                   meta_data = MetaData2,
+                                   sample_var = sample_var,
+                                   group_var = NULL,
+                                   out_cut = out_cut,
+                                   zero_cut = zero_cut,
+                                   lib_cut = lib_cut,
+                                   neg_lb = neg_lb)
+
+# Step 2: ANCOM
+feature_table = prepro$feature_table 
+meta_data = prepro$meta_data
+struc_zero = prepro$structure_zeros
+
+# Step 2.1: Tertile 1 and 3
+tert1_3 <- df2[df2$tmao_tertile == "Tertile1" | df2$tmao_tertile == "Tertile3",]
+feature_table = feature_table[,colnames(feature_table) %in% rownames(tert1_3)]
+meta_data = meta_data[rownames(meta_data) %in% rownames(tert1_3),]
+
+# Define variables for ANCOM
+main_var = "tmao_tertile"; p_adj_method = "BH"; alpha = 0.05
+adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
+# Run ANCCOM
+t_start = Sys.time()
+res = ANCOM(feature_table = feature_table, 
+            meta_data = meta_data, 
+            struc_zero = struc_zero, 
+            main_var = main_var, 
+            p_adj_method = p_adj_method, 
+            alpha = alpha, 
+            adj_formula = adj_formula, 
+            rand_formula = rand_formula)
+t_end = Sys.time()
+t_run = t_end - t_start 
+t_run # around 5s
+
+# Results
+res$out
+res$fig$data
+
+# Step 3: Volcano Plot
+# Number of taxa except structural zeros
+n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
+# Cutoff values for declaring differentially abundant taxa
+cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
+names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")
+
+# Annotation data
+dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")
+
+fig = res$fig +  
+  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
+  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
+            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
+fig
+
+# ORGANIZE PLOT
+
+TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
+rownames(TaxaData2) <- TaxaData2$Feature.ID
+
+ancomRes <- res$out
+
+test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)
+
+# Make plot 
+tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
+asv_id <- unique(tempdata$taxa_id)
+
+MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]
+
+# Rename to the taxonomy 
+#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
+TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
+# Make a unique naming column that incorporates multiple levels of taxonomy
+TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
+rownames(TaxaData3) <- rownames(TaxaData2)
+# Merge
+MicroData5 <- merge(MicroData4, TaxaData3,
+                    by=0, all.x = T, all.y = F)
+
+row.names(MicroData5) <- MicroData5$OrderFamilyGenus
+
+MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
+MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))
+
+# Merge with the phenotype data 
+MetaData3 <- MetaData2 %>% dplyr::select(tmao_tertile)
+
+ggData <- merge(MetaData3, MicroData5, by =0)
+row.names(ggData) <- ggData$Row.names
+ggData <- ggData %>% dplyr::select(-Row.names)
+
+# Gather the data
+library(tidyr)
+ggData2 <- gather(ggData,
+                  key = "Taxa",
+                  value = "Abundance",
+                  -tmao_tertile)
+
+ggData2$Abundance <- log(ggData2$Abundance)
+
+# Plot it
+p <- ggplot(data = ggData2,
+            aes(x = Taxa, 
+                y = Abundance,
+                fill = tmao_tertile), 
+            alpha = 0.1) +
+  geom_boxplot(lwd=.25) +
+  xlab(NULL) +
+  ylab("Log of Taxa Abundance") +
+  theme_bw() + # note order matters here - need bw to be before theme()
+  theme(axis.text.y = element_text( hjust = 1),
+        axis.text.x = element_text(angle = 60, hjust = 1),
+        text=element_text(size=11))  +
+  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
+  #scale_fill_viridis(discrete = TRUE) +
+  #coord_flip() +
+  labs(fill = "TMAO \n Classification")
+
+p
+```

### 2021-11-25 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update analysis.

@@
 
+# Factor sex, BMI okay
+sample_data(PSOtmao)$sex = factor(sample_data(PSOtmao)$sex)
+
 # Make sure "below" is the reference factor
@@
 
-```{r Supervised Filtering 5%}
+```{r Supervised Filtering 0.05}
 # Define prevalence threshold as 5% of total samples
@@
 
-Great. Do your due diligence and inspect each phyloseq object to see how the filtering and glomming steps changed the nubmer of taxonomy.
+Great. Do your due diligence and inspect each phyloseq object to see how the filtering and glomming steps changed the number of taxonomy.
 
@@
 # Family
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Family)) +
+ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Family)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
@@
 # Genus - have to blow it up to make it useful
-ggplot(prevdf1, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Genus)) +
+ggplot(prevdf, aes(TotalAbundance, Prevalence / nsamples(PSOtmao1G),color=Genus)) +
   geom_hline(yintercept = 0.05, alpha = 0.5, linetype = 2) + # Include a guess for parameter
@@
 
-> # Test the impact of fiber cluster, accounting for covariates, using the log ratio test
-dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
+> DESeq2 Set Up
+> Test the impact of fiber cluster, accounting for covariates, using the log ratio test
 >
-> # This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero.
-dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
+> dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
+> This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero.
 >
-> # The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
-resultsNames(dds.ratio.batcho)
-res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))
+> dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
+> 
+> The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
+> 
+> resultsNames(dds.ratio.batcho)
+> res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))
 
 ## Factored by TMAO Median
+
 ```{r DESeq2 analysis - TMAO median}
-# Grab the sample data to easily check the structure of your covariates, age, sex, and BMI
-sd <- sample_data(PSOtmao1G)
-str(sd$sex)
-str(sd$bmi_final.x)
-# Factor sex, BMI okay
-sample_data(PSOtmao1G)$sex = factor(sample_data(PSOtmao1G)$sex)
-
 #~~~~~~~~~~~~~~~
@@
 ```{r DESeq2 analysis - TMAO tertile}
-
 # DESeq analysis - tmao_tertile
@@
 ```
+Save
+
+```{r save deseq2 tmao tertile results}
+write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv")
+```
 
@@
 # PCoA
-PCoA_WU <- ordinate(PSOtmao4G, "PCoA", "wunifrac")
-PCoA_UnWU <- ordinate(PSOtmao4G, "PCoA", "uunifrac")
-PCoA_BC <- ordinate(PSOtmao4G, "PCoA", "bray")
+PCoA_WU <- ordinate(PSOtmao1G, "PCoA", "wunifrac")
+PCoA_UnWU <- ordinate(PSOtmao1G, "PCoA", "uunifrac")
+PCoA_BC <- ordinate(PSOtmao1G, "PCoA", "bray")
 
 # Calculate distance matrix
-dismax_wu = phyloseq::distance(PSOtmao4G, method="wunifrac", type = "samples")
-dismax_uwu = phyloseq::distance(PSOtmao4G, method="unifrac", type = "samples")
-dismax_bc = phyloseq::distance(PSOtmao4G, method="bray", type = "samples")
-
+dismax_wu = phyloseq::distance(PSOtmao1G, method="wunifrac", type = "samples")
+dismax_uwu = phyloseq::distance(PSOtmao1G, method="unifrac", type = "samples")
+dismax_bc = phyloseq::distance(PSOtmao1G, method="bray", type = "samples")
+
+```
+
+Save
+
+```{r save beta matrices}
+save(PCoA_WU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU.RData")
+save(PCoA_UnWU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU.RData")
+save(PCoA_BC, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC.RData")
+
+save(dismax_wu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU.RData")
+save(dismax_uwu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU.RData")
+save(dismax_bc, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC.RData")
 ```
@@
 # Get sample_data
-df <- as(sample_data(PSOtmao4G), "data.frame")
+df <- as(sample_data(PSOtmao1G), "data.frame")
 df$tmao_mdn <- relevel(df$tmao_mdn, ref = "below")
@@
 # ADONIS (perMANOVA) analysis
-adonis2(dismax_wu ~ tmao_log + sex:age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .025
+adonis2(dismax_wu ~ tmao_log + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .025
+#adonis2(tmao_log ~ dismax_wu + sex*age + cystatinc_bd1, data=df, method = "wunifrac", perm=999)
 
@@
 # ADONIS (perMANOVA) analysis
-adonis2(dismax_uwu ~ tmao_log + sex:age + bmi_final.y, data=df, method = "unifrac", perm=999) # .001
+adonis2(dismax_uwu ~ tmao_log + sex + age + bmi_final.y, data=df, method = "unifrac", perm=999) # .001
+#adonis2(tmao_log ~ dismax_uwu + sex*age + cystatinc_bd1, data=df, method = "unifrac", perm=999)
 
@@
 # ADONIS (perMANOVA) analysis
-adonis2(dismax_bc ~ tmao_log + sex:age + bmi_final.y, data=df, method = "bray", perm=999) # .017
+adonis2(dismax_bc ~ tmao_log + sex + age + bmi_final.y, data=df, method = "bray", perm=999) # .017
+#adonis2(tmao_log ~ dismax_bc + sex*age + cystatinc_bd1, data=df, method = "bray", perm=999)
 
@@
 ```
-
-Repeat looking at tertile 1 and 3. This section isn't currently working but I'm not sure why.
-
-```{r ANCOM - genus level with covariates tertile}
-source("../../Rscripts/ancom_v2.1.R")
-
-# OTU table
-MicroData <- otu_table(PSOtmao1G)
-MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
-# Add 1
-#MicroData2 <- MicroData +1
-#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
-#dim(MicroData2) # 103 x 98
-
-# Load the phenotype data
-#------------------------
-# Perform ANCOM analysis
-# Step 1: Data preprocessing
-# OTU data
-otu_data <- otu_table(PSOtmao1G)
-otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
-otu_id <- rownames(otu_data)
-
-# Metadata 
-#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
-#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
-# Subset
-MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
-MetaData2$Sample.ID <- rownames(MetaData2)
-
-feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
-out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
-# Function
-prepro = feature_table_pre_process(feature_table = feature_table,
-                                   meta_data = MetaData2,
-                                   sample_var = sample_var,
-                                   group_var = NULL,
-                                   out_cut = out_cut,
-                                   zero_cut = zero_cut,
-                                   lib_cut = lib_cut,
-                                   neg_lb = neg_lb)
-
-# Step 2: ANCOM
-feature_table = prepro$feature_table 
-meta_data = prepro$meta_data
-struc_zero = prepro$structure_zeros
-
-# Step 2.1: Tertile 1 and 3
-tert1_3 <- df2[df2$tmao_tertile == "Tertile1" | df2$tmao_tertile == "Tertile3",]
-feature_table = feature_table[,colnames(feature_table) %in% rownames(tert1_3)]
-meta_data = meta_data[rownames(meta_data) %in% rownames(tert1_3),]
-
-# Define variables for ANCOM
-main_var = "tmao_tertile"; p_adj_method = "BH"; alpha = 0.05
-adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
-# Run ANCCOM
-t_start = Sys.time()
-res = ANCOM(feature_table = feature_table, 
-            meta_data = meta_data, 
-            struc_zero = struc_zero, 
-            main_var = main_var, 
-            p_adj_method = p_adj_method, 
-            alpha = alpha, 
-            adj_formula = adj_formula, 
-            rand_formula = rand_formula)
-t_end = Sys.time()
-t_run = t_end - t_start 
-t_run # around 5s
-
-# Results
-res$out
-res$fig$data
-
-# Step 3: Volcano Plot
-# Number of taxa except structural zeros
-n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
-# Cutoff values for declaring differentially abundant taxa
-cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
-names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")
-
-# Annotation data
-dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")
-
-fig = res$fig +  
-  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
-  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
-            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
-fig
-
-# ORGANIZE PLOT
-
-TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
-rownames(TaxaData2) <- TaxaData2$Feature.ID
-
-ancomRes <- res$out
-
-test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)
-
-# Make plot 
-tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
-asv_id <- unique(tempdata$taxa_id)
-
-MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]
-
-# Rename to the taxonomy 
-#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
-TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
-# Make a unique naming column that incorporates multiple levels of taxonomy
-TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
-rownames(TaxaData3) <- rownames(TaxaData2)
-# Merge
-MicroData5 <- merge(MicroData4, TaxaData3,
-                    by=0, all.x = T, all.y = F)
-
-row.names(MicroData5) <- MicroData5$OrderFamilyGenus
-
-MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
-MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))
-
-# Merge with the phenotype data 
-MetaData3 <- MetaData2 %>% dplyr::select(tmao_tertile)
-
-ggData <- merge(MetaData3, MicroData5, by =0)
-row.names(ggData) <- ggData$Row.names
-ggData <- ggData %>% dplyr::select(-Row.names)
-
-# Gather the data
-library(tidyr)
-ggData2 <- gather(ggData,
-                  key = "Taxa",
-                  value = "Abundance",
-                  -tmao_tertile)
-
-ggData2$Abundance <- log(ggData2$Abundance)
-
-# Plot it
-p <- ggplot(data = ggData2,
-            aes(x = Taxa, 
-                y = Abundance,
-                fill = tmao_tertile), 
-            alpha = 0.1) +
-  geom_boxplot(lwd=.25) +
-  xlab(NULL) +
-  ylab("Log of Taxa Abundance") +
-  theme_bw() + # note order matters here - need bw to be before theme()
-  theme(axis.text.y = element_text( hjust = 1),
-        axis.text.x = element_text(angle = 60, hjust = 1),
-        text=element_text(size=11))  +
-  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
-  #scale_fill_viridis(discrete = TRUE) +
-  #coord_flip() +
-  labs(fill = "TMAO \n Classification")
-
-p
-```
+# Stats for paper
+
+Take a TMAO centric view point. So for the following questions, consider the whole group and the TMAO quantiles/tertiles.
+
+```{r Desc.Stats set up}
+# Get sample data groupings
+sampleData <- as.data.frame(sample_data(PSOtmao1))
+sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
+
+# Make lists of participants in each group
+tertile1 <- rownames(sampleData[sampleData$tmao_tertile == "Tertile1",])
+tertile2 <- rownames(sampleData[sampleData$tmao_tertile == "Tertile2",])
+tertile3 <- rownames(sampleData[sampleData$tmao_tertile == "Tertile3",])
+
+# Get OTU table
+otuTbl_PSO1 <- as.data.frame(otu_table(PSOtmao1))
+otuTbl_PSO1 <- as.data.frame(t(otuTbl_PSO1)) # subjects to rows
+
+# Get tax table
+taxaData <- as.data.frame(tax_table(PSOtmao1))
+taxaData$PCOFG <- paste0(taxaData$Phylum, taxaData$Class, taxaData$Order, taxaData$Family, taxaData$Genus)
+
+```
+
+How many OTUs (not glommed) in analysis?
+
+```{r How many}
+#~~~~~~~~~~
+# Tertile 1
+#~~~~~~~~~~
+# What is the average number of OTUs per participant in tertile 1?
+taxa_occurrences_T1 <- sapply(otuTbl_PSO1[row.names(otuTbl_PSO1) %in% tertile1,], function(x) sum(x>0))
+summary(taxa_occurrences_T1)
+
+otuTbl_PSO1_smpCol <- as.data.frame(t(otuTbl_PSO1))
+ppl_occurrences_T1 <- sapply(otuTbl_PSO1_smpCol[colnames(otuTbl_PSO1_smpCol) %in% tertile1,], function(x) sum(x>0))
+summary(ppl_occurrences_T1)
+
+#~~~~~~~~~~
+# Tertile 2
+#~~~~~~~~~~
+# What is the average number of OTUs per participant in tertile 1?
+taxa_occurrences_T2 <- sapply(otuTbl_PSO1[row.names(otuTbl_PSO1) %in% tertile2,], function(x) sum(x>0))
+summary(taxa_occurrences_T2)
+
+otuTbl_PSO1_smpCol <- as.data.frame(t(otuTbl_PSO1))
+ppl_occurrences_T2 <- sapply(otuTbl_PSO1_smpCol[colnames(otuTbl_PSO1_smpCol) %in% tertile2,], function(x) sum(x>0))
+summary(ppl_occurrences_T2)
+
+#~~~~~~~~~~
+# Tertile 3
+#~~~~~~~~~~
+# What is the average number of OTUs per participant in tertile 1?
+taxa_occurrences_T3 <- sapply(otuTbl_PSO1[row.names(otuTbl_PSO1) %in% tertile3,], function(x) sum(x>0))
+summary(taxa_occurrences_T3)
+
+otuTbl_PSO1_smpCol <- as.data.frame(t(otuTbl_PSO1))
+ppl_occurrences_T3 <- sapply(otuTbl_PSO1_smpCol[colnames(otuTbl_PSO1_smpCol) %in% tertile3,], function(x) sum(x>0))
+summary(ppl_occurrences_T3)
+```
+
+How many OTUs are present in 5%, 25%, 50% of samples?
+
+```{r}
+
+```
+
+What is the firmicutes to bacteroidetes ratio per tertile?
+How common are Firmicutes? Mean, SD, range?
+How common are Bacteroidetes? Mean, SD, range?
+
+```{r Firmicutes to bacteroidetes ratio per tertile}
+PSOtmao1_Phylum <- tax_glom(PSOtmao1, "Phylum", NArm = TRUE)
+PSOtmao1_Phylum_tertile1 <- subset_samples(PSOtmao1_Phylum, tmao_tertile == "Tertile1")
+PSOtmao1_Phylum_tertile2 <- subset_samples(PSOtmao1_Phylum, tmao_tertile == "Tertile2")
+PSOtmao1_Phylum_tertile3 <- subset_samples(PSOtmao1_Phylum, tmao_tertile == "Tertile3")
+
+#~~~~~~~~~~
+# Tertile 1
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalence_Phy_tert1 = apply(X = otu_table(PSOtmao1_Phylum_tertile1),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1_Phylum_tertile1), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalence_Phy_tert1 = data.frame(Prevalence = prevalence_Phy_tert1,
+                    TotalAbundance = taxa_sums(PSOtmao1_Phylum_tertile1), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1_Phylum_tertile1))
+
+# Firmicutes:Bacteroidetes Abundance Ratio
+firm_bacter_t1 <- prevalence_Phy_tert1$TotalAbundance[prevalence_Phy_tert1$Phylum == " p__Firmicutes"] / prevalence_Phy_tert1$TotalAbundance[prevalence_Phy_tert1$Phylum == " p__Bacteroidetes"]
+
+#~~~~~~~~~~
+# Tertile 2
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalence_Phy_tert2 = apply(X = otu_table(PSOtmao1_Phylum_tertile2),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1_Phylum_tertile2), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalence_Phy_tert2 = data.frame(Prevalence = prevalence_Phy_tert2,
+                    TotalAbundance = taxa_sums(PSOtmao1_Phylum_tertile2), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1_Phylum_tertile2))
+
+firm_bacter_t2 <- prevalence_Phy_tert2$TotalAbundance[prevalence_Phy_tert2$Phylum == " p__Firmicutes"] / prevalence_Phy_tert2$TotalAbundance[prevalence_Phy_tert2$Phylum == " p__Bacteroidetes"]
+
+#~~~~~~~~~~
+# Tertile 3
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalence_Phy_tert3 = apply(X = otu_table(PSOtmao1_Phylum_tertile3),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1_Phylum_tertile3), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalence_Phy_tert3 = data.frame(Prevalence = prevalence_Phy_tert3,
+                    TotalAbundance = taxa_sums(PSOtmao1_Phylum_tertile3), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1_Phylum_tertile3))
+
+firm_bacter_t3 <- prevalence_Phy_tert3$TotalAbundance[prevalence_Phy_tert3$Phylum == " p__Firmicutes"] / prevalence_Phy_tert3$TotalAbundance[prevalence_Phy_tert3$Phylum == " p__Bacteroidetes"]
+
+# Compare ratios
+cat("Firmicutes:Bacteroidetes ratio tertile 1:", firm_bacter_t1)
+cat("Firmicutes:Bacteroidetes ratio tertile 2:", firm_bacter_t2)
+cat("Firmicutes:Bacteroidetes ratio tertile 3:", firm_bacter_t3)
+
+```
+
+What is the % abundance of a given taxa? Get summary statistics per tertile.
+
+Range percentage of taxa per person = n occurrences of bug/total occurrences of all bugs * 100
+
+```{r Firmicute summary}
+#~~~~~~~~~~
+# Tertile 1
+#~~~~~~~~~~
+trans <- transform_sample_counts(PSOtmao1_Phylum_tertile1, function(x) x / sum(x))
+melted <- psmelt(trans)
+subdf <- melted[melted$Phylum == " p__Firmicutes", ]
+cat("Summmary of Firmicute abundance in tertile 1:")
+summary(subdf$Abundance)
+
+#~~~~~~~~~~
+# Tertile 2
+#~~~~~~~~~~
+trans <- transform_sample_counts(PSOtmao1_Phylum_tertile2, function(x) x / sum(x))
+melted <- psmelt(trans)
+subdf <- melted[melted$Phylum == " p__Firmicutes", ]
+cat("Summmary of Firmicute abundance in tertile 2:")
+summary(subdf$Abundance)
+
+#~~~~~~~~~~
+# Tertile 3
+#~~~~~~~~~~
+trans <- transform_sample_counts(PSOtmao1_Phylum_tertile3, function(x) x / sum(x))
+melted <- psmelt(trans)
+subdf <- melted[melted$Phylum == " p__Firmicutes", ]
+cat("Summmary of Firmicute abundance in tertile 3:")
+summary(subdf$Abundance)
+
+```
+
+```{r Bacteroidetes summary}
+#~~~~~~~~~~
+# Tertile 1
+#~~~~~~~~~~
+trans <- transform_sample_counts(PSOtmao1_Phylum_tertile1, function(x) x / sum(x))
+melted <- psmelt(trans)
+subdf <- melted[melted$Phylum == " p__Bacteroidetes", ]
+cat("Summmary of Bacteroidetes abundance in tertile 1:")
+summary(subdf$Abundance)
+
+#~~~~~~~~~~
+# Tertile 2
+#~~~~~~~~~~
+trans <- transform_sample_counts(PSOtmao1_Phylum_tertile2, function(x) x / sum(x))
+melted <- psmelt(trans)
+subdf <- melted[melted$Phylum == " p__Bacteroidetes", ]
+cat("Summmary of Bacteroidetes abundance in tertile 2:")
+summary(subdf$Abundance)
+
+#~~~~~~~~~~
+# Tertile 3
+#~~~~~~~~~~
+trans <- transform_sample_counts(PSOtmao1_Phylum_tertile3, function(x) x / sum(x))
+melted <- psmelt(trans)
+subdf <- melted[melted$Phylum == " p__Bacteroidetes", ]
+cat("Summmary of Bacteroidetes abundance in tertile 3:")
+summary(subdf$Abundance)
+
+```
+
+Genus: What are the most common genus?
+Use the PSOtmao1G, subset to low-tertile, make prevalence data frame, sort in descending order. Repeat for other tertiles.
+
+```{r Genus Prevalence by TMAO Tertile}
+# Subset by tertiles 
+PSOtmao1_tertile1 <- subset_samples(PSOtmao1, tmao_tertile == "Tertile1")
+PSOtmao1_tertile2 <- subset_samples(PSOtmao1, tmao_tertile == "Tertile2")
+PSOtmao1_tertile3 <- subset_samples(PSOtmao1, tmao_tertile == "Tertile3")
+
+# How many taxa are there per tertile?
+cat("n taxa in tertile 1:", ntaxa(PSOtmao1_tertile1))
+cat("n taxa in tertile 2:", ntaxa(PSOtmao1_tertile2))
+cat("n taxa in tertile 3:", ntaxa(PSOtmao1_tertile3))
+
+# Now subset at the genus level to identify the most common genus
+# Subset by tertiles (Genus)
+PSOtmao1G_tertile1 <- subset_samples(PSOtmao1G, tmao_tertile == "Tertile1")
+PSOtmao1G_tertile2 <- subset_samples(PSOtmao1G, tmao_tertile == "Tertile2")
+PSOtmao1G_tertile3 <- subset_samples(PSOtmao1G, tmao_tertile == "Tertile3")
+
+#~~~~~~~~~~
+# Tertile 1
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalence_tert1 = apply(X = otu_table(PSOtmao1G_tertile1),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1G_tertile1), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalence_tert1 = data.frame(Prevalence = prevalence_tert1,
+                    TotalAbundance = taxa_sums(PSOtmao1G_tertile1), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1G_tertile1))
+
+#~~~~~~~~~~
+# Tertile 2
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalence_tert2 = apply(X = otu_table(PSOtmao1G_tertile2),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1G_tertile2), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalence_tert2 = data.frame(Prevalence = prevalence_tert2,
+                    TotalAbundance = taxa_sums(PSOtmao1G_tertile2), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1G_tertile2))
+#~~~~~~~~~~
+# Tertile 3
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalence_tert3 = apply(X = otu_table(PSOtmao1G_tertile3),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1G_tertile3), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalence_tert3 = data.frame(Prevalence = prevalence_tert3,
+                    TotalAbundance = taxa_sums(PSOtmao1G_tertile3), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1G_tertile3))
+```
+
+View and sort top 10
+```{r View abundant genus top 10}
+cat("Most abundant genus in T1:")
+head(prevalence_tert1[order(prevalence_tert1$TotalAbundance, decreasing= TRUE),], n = 10)
+cat("Most abundant genus in T2:")
+head(prevalence_tert2[order(prevalence_tert2$TotalAbundance, decreasing= TRUE),], n = 10)
+cat("Most abundant genus in T3:")
+head(prevalence_tert3[order(prevalence_tert3$TotalAbundance, decreasing= TRUE),], n = 10)
+
+```
+
+Save
+
+```{r Save tertile prevalence csv}
+write.csv(prevalence_tert1, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile1.csv")
+write.csv(prevalence_tert2, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile2.csv")
+write.csv(prevalence_tert3, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile3.csv")
+```
+
+# Correlate Fecal Microbiome with Phenos
+
+Now that we identified microbiome characteristics that are relevant to TMAO, ask if they
+
+1) Improve the variance explained between TMAO and a given cardiometabolic marker
+2) Correlate to cardiometabolic phenos
+
+```{r Get the microbiome data ready to combine with pheno data}
+da <- read.csv("../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv", row.names = 1)
+resOTU <- rownames(da)
+otuData <- as.data.frame(otu_table(PSOtmao1G))
+otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
+otuData_desRes <- as.data.frame(t(otuData_desRes))
+# rename for ease downstream
+names(otuData_desRes)[names(otuData_desRes) == "bb1b75f41ff9c9db1d1de41e8388eb52"] <- "f__Bacteroidaceae_g__Bacteroides"
+names(otuData_desRes)[names(otuData_desRes) == "b180a2455b236c103cf0bfe620095736"] <- "f__Peptostreptococcaceae_g__"
+names(otuData_desRes)[names(otuData_desRes) == "e1a2800b24cdf9779b28dc897cddb12a"] <- "f__Christensenellaceae_g__"
+names(otuData_desRes)[names(otuData_desRes) == "8f5ceded1cd7e86b8271d3fab24d322a"] <- "f__Lachnospiraceae_g__Butyrivibrio"
+```
+
+Next chunk is borrowed from "Characteristics_Diet_Cardiometabolic"
+
+```{r Load phenotype data}
+phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")
+
+# Complete TMAO data only
+phen <- phen[complete.cases(phen$tmao == TRUE),]
+
+# Get rid of NA ethnicity category
+phen$ethnicity <- droplevels(phen$ethnicity)
+
+# Factor and recode sex
+phen$sex <- as.factor(phen$sex)
+phen$sex <- plyr::revalue(phen$sex, c("1" = "Male", "2" = "Female"))
+
+# Make TMAO log variable
+phen$tmao_log <- log(phen$tmao)
+
+# Quantile
+tmao_quantile <- quantile(phen$tmao)
+phen$tmao_quantile <- ifelse(phen$tmao <= tmao_quantile[2], "Quantile1",
+                             ifelse(phen$tmao > tmao_quantile[2] & phen$tmao <= tmao_quantile[3], "Quantile2",
+                                   ifelse(phen$tmao > tmao_quantile[3] & phen$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))
+
+# Check it
+table(phen$tmao_quantile)
+phen$tmao_quantile <- as.factor(phen$tmao_quantile)
+str(phen$tmao_quantile)
+
+# Tertile
+tmao_tertile <- quantile(phen$tmao, c(0:3/3))
+phen$tmao_tertile <- ifelse(phen$tmao <= tmao_tertile[2], "Tertile1",
+                             ifelse(phen$tmao > tmao_tertile[2] & phen$tmao <= tmao_tertile[3], "Tertile2", "Tertile3"))
+# Factor it
+phen$tmao_tertile <- factor(phen$tmao_tertile)
+table(phen$tmao_tertile)
+
+# Checked all variables for normality via shapiro.test() in console. TG, and glc ere the only ones that didn't pass.
+phen$glc_bd1_ln <- log(phen$glc_bd1)
+phen$tg_bd1_ln <- log(phen$tg_bd1)
+phen$crp_bd1_ln <- log(phen$crp_bd1)
+phen$tnfa_bd1_ln <- log(phen$tnfa_bd1)
+phen$il6_bd1_ln <- log(phen$il6_bd1)
+```
+
+Merge phenotype and microbiome data for linear regression analysis
+```{r}
+phen <- merge(phen, otuData_desRes, by=0)
+rownames(phen) <- phen[,"Row.names"]
+phen <- subset(phen, select = -c(Row.names))
+```
+
+Run the regression!
+
+f__Bacteroidaceae_g__Bacteroides + f__Peptostreptococcaceae_g__ + f__Christensenellaceae_g__ + f__Lachnospiraceae_g__Butyrivibrio
+
+```{r Cardiometabolic + Microbiome Multiple Linear Regression}
+# Transfored variable list
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log", "f__Bacteroidaceae_g__Bacteroides", "f__Peptostreptococcaceae_g__", "f__Christensenellaceae_g__", "f__Lachnospiraceae_g__Butyrivibrio")
+
+#~~~~~~~~
+# TMAO  
+#~~~~~~~~
+# Make a smaller df from master (the old data)
+phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
+cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log, f__Bacteroidaceae_g__Bacteroides, f__Peptostreptococcaceae_g__, f__Christensenellaceae_g__, f__Lachnospiraceae_g__Butyrivibrio)) # change variable here
+n <- ncol(cvd_biomark)
+
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex + age + cystatinc_bd1, phen_sub )) # include kcal
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ ethnicity, phen_sub ))
+#my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ cystatinc_bd1, phen_sub ))
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [CVD row, estimate] corresponds to CVD B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [CVD row, Std.Error] corresponds to CVD Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [CVD row, T value] corresponds to CVD t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [CVD row, P] corresponds to CVD P value
+my_lm361_lcms_cvd <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm361_lcms_cvd) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm361_lcms_cvd) <- colnames(cvd_biomark)
+res <- as.data.frame(my_lm361_lcms_cvd)
+
+# BH correction
+res$p.adj <- stats::p.adjust(res$P, method = "BH", n = n)
+
+#~~~~~~~~
+# TMAO + Bug
+#~~~~~~~~
+# Make a smaller df from master (the old data)
+phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
+cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log, f__Bacteroidaceae_g__Bacteroides, f__Peptostreptococcaceae_g__, f__Christensenellaceae_g__, f__Lachnospiraceae_g__Butyrivibrio)) # change variable here
+n <- ncol(cvd_biomark)
+
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex + age + cystatinc_bd1 + f__Bacteroidaceae_g__Bacteroides + f__Peptostreptococcaceae_g__ + f__Christensenellaceae_g__ + f__Lachnospiraceae_g__Butyrivibrio, phen_sub )) # add bugs
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [CVD row, estimate] corresponds to CVD B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [CVD row, Std.Error] corresponds to CVD Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [CVD row, T value] corresponds to CVD t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [CVD row, P] corresponds to CVD P value
+my_lm361_lcms_cvd_des <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm361_lcms_cvd_des) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm361_lcms_cvd_des) <- colnames(cvd_biomark)
+res_des <- as.data.frame(my_lm361_lcms_cvd_des)
+
+# BH correction
+res_des$p.adj <- stats::p.adjust(res_des$P, method = "BH", n = n) 
+```

### 2021-11-26 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Update workflow for paper

@@
 ```{r Quantile TMAO}
+# Quantile
 tmao_quantile <- quantile(phen$tmao)
@@
 str(phen$tmao_quantile)
+
+# Tertile
+tmao_tertile <- quantile(phen$tmao, c(0:3/3))
+phen$tmao_tertile <- ifelse(phen$tmao <= tmao_tertile[2], "Tertile1",
+                             ifelse(phen$tmao > tmao_tertile[2] & phen$tmao <= tmao_tertile[3], "Tertile2", "Tertile3"))
+# Factor it
+phen$tmao_tertile <- factor(phen$tmao_tertile)
+table(phen$tmao_tertile)
+
+# Print tertiles for manuscript
+tmao_tertile
 
@@
 
+Save
+
 ```{r Save characteristic quantile tables}
@@
 write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOQuantile_211111.csv")
+```
+
+Make similar tables but for TMAO tertile
+
+```{r Participant characterisitics by TMAO tertile}
+# TMAO table by tmao_tertile - females
+factorVars <- c("age_cat", "tmao_tertile")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+table_tmao.ft <- CreateTableOne(vars = vars,
+                                    strata = "tmao_tertile",
+                                    data = phen[phen$sex == "Female",],
+                                    factorVars = factorVars)
+table_tmao.ft
+
+# TMAO table by tmao_tertile - males
+factorVars <- c("age_cat", "tmao_tertile")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+table_tmao.mt <- CreateTableOne(vars = vars,
+                                    strata = "tmao_tertile",
+                                    data = phen[phen$sex == "Male",], # note 6103 and 7098 have high il6
+                                    factorVars = factorVars)
+table_tmao.mt
+
+# TMAO table by tmao_tertile - all, add sex
+factorVars <- c("tmao_tertile")
+vars <- c("age", "sex", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+table_tmao.allt <- CreateTableOne(vars = vars,
+                                    strata = "tmao_tertile",
+                                    data = phen,
+                                    factorVars = factorVars)
+table_tmao.allt
+
+# Do BH correction
+allt_ContTable <- attr(table_tmao.allt$ContTable, "pValues")
+n <- length(allt_ContTable$pNormal)
+# BH correction
+allt_ContTable$padj <- stats::p.adjust(allt_ContTable$pNormal, method = "BH", n = n) 
+allt_ContTable
+```
+
+Save
+
+```{r Save characteristic tertile tables}
+table_save.ft <- print(table_tmao.ft, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.ft, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byTMAOTertile.csv")
+
+table_save.mt <- print(table_tmao.mt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.mt, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byTMAOTertile.csv")
+
+table_save.allt <- print(table_tmao.allt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.allt, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOTertile.csv")
+
+# BH corrections for the all tertile table
+write.csv(allt_ContTable, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOTertile_Padj.csv")
 ```
@@
 
+TMAO tertile 
+Make TMAO food intake by tertile table
+```{r Food Intake by TMAO tertile}
+# TMAO table by age bin - females
+factorVars <- c("age_cat", "tmao_quantile")
+vars <- c("pf_meat_totals", 
+          "pf_curedmeat_totals",
+          "pf_poult_totals",
+          "pf_seafd_hi_totals",
+          "pf_seafd_low_totals",
+          "pf_eggs_totals",
+          "pf_nutsds_totals",
+          "pf_legumes_totals",
+          "pf_total_totals",
+          #"pf_soy_totals_Binary",
+          "v_drkgr_totals",
+          "v_redor_total_totals",
+          #"v_redor_tomato_totals",
+          #"v_redor_other_totals",
+          "v_starchy_total_totals",
+          #"v_starchy_potato_totals",
+          #"v_starchy_other_totals",
+          "v_other_totals",
+          "v_total_totals",
+          #"v_legumes_total",
+          "g_whole_totals",
+          "g_refined_totals",
+          "g_total_totals",
+          "d_total_totals",
+          "d_milk_totals",
+          "d_cheese_totals",
+          "d_yogurt_totals",
+          "hei_asa24_totalscore",
+          "hei_asa24_totalveg",
+          "hei_asa24_sfat",
+          "hei_asa24_",
+          "age",
+          "ethnicity" ,
+          "bmi_final",
+          "tmao",
+          "choline",
+          "betaine",
+          "carnitine")
+table_food.ft <- CreateTableOne(vars = vars,
+                                    strata = "tmao_tertile",
+                                    data = full[full$sex == "Female",],
+                                    factorVars = factorVars)
+table_food.ft
+
+# TMAO table by age bin - males
+table_food.mt <- CreateTableOne(vars = vars,
+                                    strata = "tmao_tertile",
+                                    data = full[full$sex == "Male",],
+                                    factorVars = factorVars)
+table_food.mt
+
+# TMAO table by sex
+table_food.allt <- CreateTableOne(vars = vars,
+                                    strata = "tmao_tertile",
+                                    data = full,
+                                    factorVars = factorVars)
+table_food.allt
+
+# Do BH correction
+food_allt_ContTable <- attr(table_food.allt$ContTable, "pValues")
+n <- length(food_allt_ContTable$pNormal)
+# BH correction
+food_allt_ContTable$padj <- stats::p.adjust(food_allt_ContTable$pNormal, method = "BH", n = n) 
+food_allt_ContTable
+```
+
+```{r Save food tertile tables}
+table_save.ft <- print(table_food.ft, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.ft, file = "../../Outputs/tables/ParticipantChar/Food_Female_byTMAOTertile.csv")
+
+table_save.mt <- print(table_food.mt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.mt, file = "../../Outputs/tables/ParticipantChar/Food_Male_byTMAOTertile.csv")
+
+table_save.allt <- print(table_food.allt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
+write.csv(table_save.allt, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOTertile.csv")
+
+write.csv(food_allt_ContTable, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOTertile_PadjTable.csv")
+```
+
 # Miscellaneous for paper
@@
 summary(full$pf_meat_totals)
+sd(full$pf_meat_totals)
 
@@
 summary(full$pf_curedmeat_totals)
+sd(full$pf_curedmeat_totals)
+
+# Chicken, turkey, Cornish hens, duck, goose, quail, and pheasant (game birds); excludes organ meat and cured meat (oz. eq.)
+summary(full$pf_poult_totals)
+sd(full$pf_poult_totals)
 
@@
 summary(full$pf_total_totals)
+sd(full$pf_total_totals)
 
@@
 summary(full$pf_mps_total_totals)
+sd(full$pf_mps_total_totals)
+
+# Eggs (chicken, duck, goose, quail) and egg substitutes (oz. eq.)
+summary(full$pf_eggs_totals)
+sd(full$pf_eggs_totals)
+
+# HEI total score
+summary(full$hei_asa24_totalscore)
+sd(full$hei_asa24_totalscore, na.rm = TRUE)
 
@@
 summary(full$kcal_tnfs)
+sd(full$kcal_tnfs)
 ```

### 2021-12-01 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Add dietary choline to analysis

@@
 phen$tmao_log <- log(phen$tmao)
+
 ```
@@
 full <- subset(full, select = -c(Row.names))
+
+# Make ASA24 choline log variable
+full$choln_tnfs_log <- log(full$choln_tnfs)
 ```
@@
 "T.d_cheese_totals_sqrRt",
-"T.d_yogurt_totals_Binary"
+"T.d_yogurt_totals_Binary",
+"choln_tnfs_log"
 )
@@
           "hei_asa24_",
+          "choln_tnfs",
           "age",
@@
 
+"waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln",
+
+BS section below...
 ```{r}
@@
 full$low_mps_meat <- ifelse(full$pf_mps_total_totals <= 9.091, "low", "high")
-summary(lm(tmao_log ~ low_mps_meat + sex*age + cystatinc_bd1 + kcal_tnfs, full))
+summary(lm(glc_bd1_ln ~  tmao_log + sex*age + cystatinc_bd1, full[full$low_mps_meat == "high",]))
+summary(lm(glc_bd1_ln ~  tmao_log + sex*age + cystatinc_bd1, full[full$low_mps_meat == "high",]))
 ```

### 2021-12-01 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Look at christensenellaceae genus with BMI

@@
 Merge phenotype and microbiome data for linear regression analysis
-```{r}
+```{r Merge pheno and microbiome tertile DESeq results}
 phen <- merge(phen, otuData_desRes, by=0)
@@
 ```
+
+The family christensenellaceae is widely inversely associated with BMI. A review is [here](https://bmcbiol.biomedcentral.com/articles/10.1186/s12915-019-0699-4).
+
+```{r Christensenellaceae genus and BMI}
+summary(lm(f__Christensenellaceae_g__ ~ bmi_final + sex + age, phen))
+```

### 2021-12-02 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Add code to save plots for manuscript

@@
 
-```{r, warning=FALSE,message=FALSE}
+```{r, warning=FALSE, message=FALSE}
 library(DESeq2)
@@
 PSOtmao1F.top10 <- microbiome::aggregate_top_taxa(PSOtmao1F, "Family", top = 10)
-ps1F.10.family.comp <- microbiome::transform(PSOtmao1F.top10, transform="compositional")
+ps1F.10.family.comp <- microbiome::transform(PSOtmao1F.top10, transform="compositional") #compositional
 
@@
   theme_minimal() +
+  upstartr::scale_y_percent() + 
   scale_fill_brewer("Family", palette = "Paired") + 
-  ggtitle("Relative Abundance at Family Level")
+  labs(x= "TMAO Tertiles", y = "Relative Abundance (%)", 
+       title = "Relative Abundance of Top 10 Prevalent Families")
 
@@
 # Save
-jpeg("../../Outputs/plots/microbiome/RelAbund_Fam_TMAOTertiles_2111.jpg", width = 4, height = 4, res = 300)
+tiff("../../Outputs/plots/microbiome/RelAbund_Fam_TMAOTertiles_2111.tiff", width = 4, height = 4, units = 'in', res = 300)
 plot.composition.relAbun
@@
     print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+            geom_boxplot()+
+            ylab(colnames(ggData)[i])) 
+}
+
+#~~~~~~~~~~~~~~~
+# Plot Results Fold Change
+#~~~~~~~~~~~~~~~
+sigtabgen = subset(sigtab, !is.na(Genus))
+# Phylum order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
+# Genus order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
+DESeqPlot <- ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
+  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
+  geom_point(size=3) + 
+  ylab("Taxonomy") +
+  xlab("Log2 Fold Change") + 
+  scale_color_brewer(palette = "Paired") +
+  theme_minimal() +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+
+DESeqPlot
+
+# Save
+tiff("../../Outputs/plots/microbiome/DESeq2_Genus_TMAOTertile1v3_2112.tiff", width = 8, height = 3, units = 'in', res = 300)
+DESeqPlot
+dev.off()
+```
+Save
+
+```{r save deseq2 tmao tertile results}
+write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv")
+```
+
+## Factored by TMAO Quantile
+
+```{r DESeq2 analysis - TMAO quantile}
+# DESeq analysis - tmao_quantile
+mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_quantile)
+mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
+res.ratio <- results(mb.ratio, contrast = c("tmao_quantile", "Quantile1", "Quantile4"))
+alpha = 0.05
+sigtab = res.ratio[which(res.ratio$padj < alpha), ]
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
+sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
+head(sigtab)
+
+#~~~~~~~~~~~~~~~
+# Plot Results
+#~~~~~~~~~~~~~~~
+resOTU <- rownames(sigtab)
+otuData <- as.data.frame(otu_table(PSOtmao1G))
+otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
+#otuData_desRes <- as.data.frame(t(otuData_desRes))
+# Get sample data groupings
+sampleData <- as.data.frame(sample_data(PSOtmao1G))
+sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
+# Get tax table
+taxaData <- as.data.frame(tax_table(PSOtmao1G))
+taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
+taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
+
+# Merge
+ggData <- merge(taxaData_res, otuData_desRes, by = 0)
+rownames(ggData) <-  ggData$OrderFamilyGenus
+ggData <- ggData[,-(1:9)]
+ggData <- as.data.frame(t(ggData))
+ggData <- merge(ggData, sampleData, by = 0)
+rownames(ggData) <- ggData[,"Row.names"]
+ggData <- subset(ggData, select = -c(Row.names))
+
+# Loop Plot
+nres <- length(resOTU)
+for (i in 1:nres) { 
+    print(ggplot(ggData, aes(x = tmao_quantile, y = log(ggData[,i]))) +
             geom_boxplot()+
@@
 ```
-Save
-
-```{r save deseq2 tmao tertile results}
-write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv")
-```
-
-## Factored by TMAO Quantile
-
-```{r DESeq2 analysis - TMAO quantile}
-# DESeq analysis - tmao_quantile
-mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_quantile)
-mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
-res.ratio <- results(mb.ratio, contrast = c("tmao_quantile", "Quantile1", "Quantile4"))
-alpha = 0.05
-sigtab = res.ratio[which(res.ratio$padj < alpha), ]
-sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
-sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
-head(sigtab)
-
-#~~~~~~~~~~~~~~~
-# Plot Results
-#~~~~~~~~~~~~~~~
-resOTU <- rownames(sigtab)
-otuData <- as.data.frame(otu_table(PSOtmao1G))
-otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
-#otuData_desRes <- as.data.frame(t(otuData_desRes))
-# Get sample data groupings
-sampleData <- as.data.frame(sample_data(PSOtmao1G))
-sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
-# Get tax table
-taxaData <- as.data.frame(tax_table(PSOtmao1G))
-taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
-taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
-
-# Merge
-ggData <- merge(taxaData_res, otuData_desRes, by = 0)
-rownames(ggData) <-  ggData$OrderFamilyGenus
-ggData <- ggData[,-(1:9)]
-ggData <- as.data.frame(t(ggData))
-ggData <- merge(ggData, sampleData, by = 0)
-rownames(ggData) <- ggData[,"Row.names"]
-ggData <- subset(ggData, select = -c(Row.names))
-
-# Loop Plot
-nres <- length(resOTU)
-for (i in 1:nres) { 
-    print(ggplot(ggData, aes(x = tmao_quantile, y = log(ggData[,i]))) +
-            geom_boxplot()+
-            ylab(colnames(ggData)[i])) 
-}
-
-#~~~~~~~~~~~~~~~
-# Plot Results Fold Change
-#~~~~~~~~~~~~~~~
-sigtabgen = subset(sigtab, !is.na(Genus))
-# Phylum order
-x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
-x = sort(x, TRUE)
-sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
-# Genus order
-x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
-x = sort(x, TRUE)
-sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
-ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
-  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
-  geom_point(size=3) + 
-  ylab("Taxonomy") +
-  xlab("Log2 Fold Change") + 
-  scale_color_brewer(palette = "Dark2") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
-```
 
@@
 # Plot
-ggplot(df2, aes(x=tmao_log, y=shannon)) +
+shannon_plot <- ggplot(df2, aes(x=tmao_log, y=shannon)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
   labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
+  scale_color_brewer(palette = "Dark2") +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+shannon_plot
 
@@
 # Plot
-ggplot(df2, aes(x=tmao_log, y=faith_pd)) +
+faith_plot <- ggplot(df2, aes(x=tmao_log, y=faith_pd)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
   labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
+  scale_color_brewer(palette = "Dark2") +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+faith_plot
 
@@
 # Plot
-ggplot(df2, aes(x=tmao_log, y=pielou_e)) +
+pielou_plot <- ggplot(df2, aes(x=tmao_log, y=pielou_e)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
   labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
+  scale_color_brewer(palette = "Dark2") +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+pielou_plot
 
@@
 # Plot
-ggplot(df2, aes(x=tmao_log, y=observed_otus)) +
+obsOTU_plot <- ggplot(df2, aes(x=tmao_log, y=observed_otus)) +
   geom_point(aes(shape=bmi_cat, color = age_cat)) +
@@
   labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2")
-
+  scale_color_brewer(palette = "Dark2") +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+obsOTU_plot
+```
+
+Save
+
+```{r save alpha diversity plots}
+# Save
+tiff("../../Outputs/plots/microbiome/AlphaDiversity_Shannon_TMAO_2112.tiff", width = 6, height = 4, units = 'in', res = 300)
+shannon_plot
+dev.off()
 ```
@@
 
+```
+
+## Beta Diversity Plots
+
+```{r beta diversity plots}
+# Weighted
+betaplot_WU <- plot_ordination(PSOtmao1G, PCoA_WU, 
+                    color="tmao_mdn",
+                    shape = NULL,
+                    axes = 1:2) +
+  scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
+  stat_ellipse(linetype = 1) +
+  labs(color = "TMAO Median Threshold") +
+  theme_minimal() + 
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+
+betaplot_WU
+
+# Unweighted
+betaplot_UWU <- plot_ordination(PSOtmao1G, PCoA_UnWU, 
+                    color="tmao_mdn",
+                    shape = NULL,
+                    axes = 1:2) +
+  scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
+  stat_ellipse(linetype = 1) +
+  labs(color = "TMAO Median Threshold") +
+  theme_minimal() + 
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+
+betaplot_UWU
+
+# Bray Curtis
+betaplot_bc <- plot_ordination(PSOtmao1G, PCoA_BC, 
+                    color="tmao_mdn",
+                    shape = NULL,
+                    axes = 1:2) +
+  scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
+  stat_ellipse(linetype = 1) +
+  labs(color = "TMAO Median Threshold") +
+  theme_minimal() + 
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+
+betaplot_bc
+
+```
+
+Save
+
+```{r save beta diversity plots}
+# Save
+tiff("../../Outputs/plots/microbiome/BetaDiversity_PCoAWU_TMAOmdn_2112.tiff", width = 6, height = 4, units = 'in', res = 300)
+betaplot_WU
+dev.off()
 ```

### 2021-12-08 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Small adjustments to tables. Added HEI. Added food intake by ethnicity analysis.

@@
 factorVars <- c("age_cat")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.f <- CreateTableOne(vars = vars,
@@
 factorVars <- c("age_cat")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.m <- CreateTableOne(vars = vars,
@@
 factorVars <- c("sex")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.all <- CreateTableOne(vars = vars,
@@
 table_save.f <- print(table_tmao.f, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.f, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byAgeBin_211111.csv")
+write.csv(table_save.f, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byAgeBin_211203.csv")
 
 table_save.m <- print(table_tmao.m, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.m, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byAgeBin_211111.csv")
+write.csv(table_save.m, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byAgeBin_211203.csv")
 
 table_save.all <- print(table_tmao.all, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.all, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_bySex_211111.csv")
+write.csv(table_save.all, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_bySex_211203.csv")
 ```
@@
 phen$il6_bd1_ln <- log(phen$il6_bd1)
+phen$insulin_bd1_ln <- log(phen$insulin_bd1)
 ```
@@
 factorVars <- c("age_cat", "tmao_quantile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.fq <- CreateTableOne(vars = vars,
@@
 factorVars <- c("age_cat", "tmao_quantile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.mq <- CreateTableOne(vars = vars,
@@
 factorVars <- c("tmao_quantile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.allq <- CreateTableOne(vars = vars,
@@
 table_save.fq <- print(table_tmao.fq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byTMAOQuantile_211111.csv")
+write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byTMAOQuantile_211203.csv")
 
 table_save.mq <- print(table_tmao.mq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byTMAOQuantile_211111.csv")
+write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byTMAOQuantile_211203.csv")
 
 table_save.allq <- print(table_tmao.allq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOQuantile_211111.csv")
+write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOQuantile_211203.csv")
 ```
@@
 factorVars <- c("age_cat", "tmao_tertile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.ft <- CreateTableOne(vars = vars,
@@
 factorVars <- c("age_cat", "tmao_tertile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.mt <- CreateTableOne(vars = vars,
@@
 # TMAO table by tmao_tertile - all, add sex
-factorVars <- c("tmao_tertile")
-vars <- c("age", "sex", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
+factorVars <- c("tmao_tertile", "sex")
+vars <- c("age", "sex", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
 table_tmao.allt <- CreateTableOne(vars = vars,
@@
 
-# Do BH correction
+# Do BH correction for continuous variables
 allt_ContTable <- attr(table_tmao.allt$ContTable, "pValues")
@@
 allt_ContTable
+
+# Do BH correction for categorical variables
+allt_CatTable <- attr(table_tmao.allt$CatTable, "pValues")
+n <- 23
+# BH correction
+allt_CatTable$padj <- stats::p.adjust(allt_CatTable$pApprox, method = "BH", n = n) 
+allt_CatTable
 ```
@@
 # Transfored variable list
-transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln", "insulin_bd1_ln", "tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
 
@@
 # Save table
-write.csv(res, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_211111.csv")
+write.csv(res, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD.csv")
 ```
@@
 # Transfored variable list
-transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln",  "insulin_bd1_ln", "tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
 
@@
 # Save table
-write.csv(res.f, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_females_211112.csv")
+write.csv(res.f, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_females.csv")
 ```
@@
 # Transfored variable list
-transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln", "insulin_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log")
 
@@
 # Save table
-write.csv(res.m, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_males_211112.csv")
+write.csv(res.m, "../../Outputs/tables/CardiometabolicMarkers/MultipleLinReg_TMAO_LCMS_n361_CVD_males.csv")
 ```
@@
 "T.d_yogurt_totals_Binary",
-"choln_tnfs_log"
+"choln_tnfs_log",
+"hei_asa24_totalscore"
 )
@@
 # Save table
-write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24_211111.csv")
+write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24.csv")
 ```
@@
           "hei_asa24_sfat",
-          "hei_asa24_",
           "choln_tnfs",
@@
 table_save.fq <- print(table_food.fq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Food_Female_byTMAOQuantile_211111.csv")
+write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Food_Female_byTMAOQuantile.csv")
 
 table_save.mq <- print(table_food.mq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Food_Male_byTMAOQuantile_211111.csv")
+write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Food_Male_byTMAOQuantile.csv")
 
 table_save.allq <- print(table_food.allq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOQuantile_211111.csv")
+write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOQuantile.csv")
 ```
@@
 
+Median TMAO in the cohort?
+
+```{r TMAO median for paper}
+cat("Summary of fasting plasma TMAO for the full cohort:")
+summary(full$tmao)
+```
+
+What is the average choline intake for males and females?
+```{r Choline intake summary}
+cat("Summary of choline intake in foods and supplements for the full cohort:")
+summary(full$choln_tnfs)
+sd(full$choln_tnfs)
+
+cat("Summary of choline intake in foods and supplements for the female cohort:")
+summary(subset(full, sex == "Female")$choln_tnfs)
+sd(subset(full, sex == "Female")$choln_tnfs)
+
+cat("Summary of choline intake in foods and supplements for the male cohort:")
+summary(subset(full, sex == "Male")$choln_tnfs)
+sd(subset(full, sex == "Male")$choln_tnfs)
+```
+
+Check to see if food intake is different by ethnicity?
+
+```{r food ~ ethnicity - LCMS BD1 - aov 353 ppl}
+# Make a smaller df from master (the old data)
+full_sub <- full[, colnames(full) %in% transformed_keep_tbl]
+
+# Get rid of NAs
+full_sub_na <- full_sub[is.na(full_sub$ethnicity) == FALSE,]
+
+full_sub_na2 <- full_sub_na %>% dplyr::select(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log, T.pf_seafd_hi_totals_Binary, T.pf_seafd_low_totals_Binary, T.d_yogurt_totals_Binary, everything())
+
+# Loop
+for(i in 11:ncol(full_sub_na2)){
+  column <- names(full_sub_na2[i])
+  aovres <- aov(full_sub_na2[,i] ~ ethnicity, data = full_sub_na2)
+  result <- summary(aovres)
+  tuk <- TukeyHSD(aovres)
+  print(column)
+  print(result)
+  print(tuk)
+}
+
+```
+```
+
 "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln",

### 2021-12-08 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Load PCoA so don't have to rerun everytime. Tweaked graphs. Saved tax table for manuscript.

@@
 table(sample_data(PSOtmao)$tmao_tertile)
+```
+
+Save taxonomy table for manuscript.
+
+```{r Save the taxonomy table for manuscript}
+tax_manuscript <- tax_table(PSOtmao)
+write.csv(tax_manuscript, "../../Outputs/tables/Microbiome/PSOtmao_taxonomyTable.csv")
 ```
@@
   xlab(x_variable) +
-  labs(color = "Age bin", shape = "BMI bin") +
-  scale_color_brewer(palette = "Dark2") +
+  scale_shape_discrete(name="BMI Bin (kg/m^2)", labels=c("18.5-24", "25-29", "30>")) +
+  scale_color_brewer(palette = "Dark2",name="Age Bin (yr)", labels=c("18-33", "34-49", "50-65")) +
   theme(axis.text = element_text(size = 12), # x and y axis text size
@@
 
+Load
+
+```{r Load bdiversity calculations}
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC.RData")
+
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC.RData")
+
+```
+
 ```{r Beta diversity stats}
@@
                     axes = 1:2) +
-  scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
+  #scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
   stat_ellipse(linetype = 1) +

### 2021-12-08 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Exploratory analysis to determine differentiating b-diversity factor visible in plots

@@
 
+Play around with the coloring to try to figure out what is discriminating these groups!
+
+```{r What defines the PCoA groups exploration}
+colnames(sample_data(PSOtmao1G))
+loopPheno <- c("pa_ee_total_mdn","endo_ln_rhi_mdn", "endo_al75_mdn", "betaine_mdn", "carnitine_mdn", "tmao_mdn",  "choline_mdn", "creatinine_mdn", "phosphocholine_mdn", "hei_asa24_totalveg_mdn", "hei_asa24_totalproteinfoods_mdn", "hei_asa24_sodium_mdn", "hei_asa24_refinedgrains_mdn", "hei_asa24_sfat_mdn",  "hei_asa24_addsug_mdn", "hei_asa24_wholegrain_mdn", "hei_asa24_totalscore_mdn", "tmao_log_mdn", "TMAO_5", "TMAO_6", "TMAO_7", "sex", "age_cat", "bmi_cat", "tmao_quantile", "tmao_tertile")
+
+# PCoA_WU, PCoA_UnWU, PCoA_BC plug in to view
+for(i in 1:length(loopPheno)){
+print(plot_ordination(PSOtmao1G, PCoA_WU, 
+                    color = loopPheno[i],
+                    shape = NULL,
+                    axes = 1:2) +
+  #scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
+  stat_ellipse(linetype = 1) +
+  labs(color = loopPheno[i]) +
+  theme_minimal() + 
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5)))
+}
+
+# From the plots, Weighted UniFrac and endo_al75 and betaine look different
+bdis <- betadisper(dismax_wu, df$betaine_mdn)
+permutest(bdis) # NS, good
+boxplot(bdis, las = 2)
+
+# ADONIS (perMANOVA) analysis
+adonis2(dismax_wu ~ betaine + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # different
+
+```
+
 # ANCOM

### 2022-01-05 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Investigate tnf-a (WIP)

@@
 
+Save for supplemental material
+```{r Save linear regression results with bacteria}
+write.csv(res_des, "../../Outputs/tables/Microbiome/LinReg_cardiometabolic_withDESeq2results.csv")
+saveRDS(res_des, "../../Outputs/tables/Microbiome/LinReg_cardiometabolic_withDESeq2results.Rds")
+```
+
+Is TNF-a related to the gut microbiome directly?
+## Factored by tnfa Tertile
+
+THIS CHUNK WORK IN PROGRESS
+
+```{r DESeq2 analysis - tnf-a tertile}
+# Quick check
+summary(lm(tnfa_bd1_ln ~ f__Bacteroidaceae_g__Bacteroides, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Peptostreptococcaceae_g__, phen_sub)) # Sig, minor inverse relationship
+summary(lm(tnfa_bd1_ln ~ f__Christensenellaceae_g__, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Lachnospiraceae_g__Butyrivibrio, phen_sub)) # NS
+# With covariates
+summary(lm(tnfa_bd1_ln ~ f__Bacteroidaceae_g__Bacteroides + sex + age + bmi_final, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Peptostreptococcaceae_g__ + sex + age + bmi_final, phen_sub)) # Sig, minor inverse relationship
+summary(lm(tnfa_bd1_ln ~ f__Christensenellaceae_g__ + sex + age + bmi_final, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Lachnospiraceae_g__Butyrivibrio + sex + age + bmi_final, phen_sub)) # NS
+
+# Now see about differential abundance with TNF-a (THIS PART WORK IN PROGRESS)
+# Add TNF-a to PSOtmao1G
+phen_tnfa <- subset(phen, select=c(tnfa_bd1))
+sample_data(PSOtmao1G)$tnfa_bd1 <- merge(sample_data(PSOtmao1G), phen_tnfa, by = 0)
+
+# Add TNF-a tertile
+tnfa_tertile <- quantile(sample_data(PSOtmao1G)$tnfa_bd1, c(0:3/3), na.rm = TRUE)
+sample_data(PSOtmao)$tmao_tertile <- ifelse(sample_data(PSOtmao)$tmao <= tmao_tertile[2], "Tertile1",
+                             ifelse(sample_data(PSOtmao)$tmao > tmao_tertile[2] & sample_data(PSOtmao)$tmao <= tmao_tertile[3], "Tertile2", "Tertile3"))
+# Factor it
+sample_data(PSOtmao)$tmao_tertile = factor(sample_data(PSOtmao)$tmao_tertile)
+# Lookss good
+table(sample_data(PSOtmao)$tmao_tertile)
+
+# DESeq analysis - tmao_tertile
+mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_tertile)
+mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
+res.ratio <- results(mb.ratio, contrast = c("tmao_tertile", "Tertile1", "Tertile3")) # Look at most different tertiles
+alpha = 0.05
+sigtab = res.ratio[which(res.ratio$padj < alpha), ]
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
+sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
+head(sigtab)
+
+#~~~~~~~~~~~~~~~
+# Plot Results
+#~~~~~~~~~~~~~~~
+resOTU <- rownames(sigtab)
+otuData <- as.data.frame(otu_table(PSOtmao1G))
+otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
+#otuData_desRes <- as.data.frame(t(otuData_desRes))
+# Get sample data groupings
+sampleData <- as.data.frame(sample_data(PSOtmao1G))
+sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
+# Get tax table
+taxaData <- as.data.frame(tax_table(PSOtmao1G))
+taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
+taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
+
+# Merge
+ggData <- merge(taxaData_res, otuData_desRes, by = 0)
+rownames(ggData) <-  ggData$OrderFamilyGenus
+ggData <- ggData[,-(1:9)]
+ggData <- as.data.frame(t(ggData))
+ggData <- merge(ggData, sampleData, by = 0)
+rownames(ggData) <- ggData[,"Row.names"]
+ggData <- subset(ggData, select = -c(Row.names))
+
+# Loop Plot
+nres <- length(resOTU)
+for (i in 1:nres) { 
+    print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+            geom_boxplot()+
+            ylab(colnames(ggData)[i])) 
+}
+
+#~~~~~~~~~~~~~~~
+# Plot Results Fold Change
+#~~~~~~~~~~~~~~~
+sigtabgen = subset(sigtab, !is.na(Genus))
+# Phylum order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
+# Genus order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
+DESeqPlot <- ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
+  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
+  geom_point(size=3) + 
+  ylab("Taxonomy") +
+  xlab("Log2 Fold Change") + 
+  scale_color_brewer(palette = "Paired") +
+  theme_minimal() +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+
+DESeqPlot
+
+# Save
+tiff("../../Outputs/plots/microbiome/DESeq2_Genus_tnfa_TMAOTertile1v3_2112.tiff", width = 8, height = 3, units = 'in', res = 300)
+DESeqPlot
+dev.off()
+```
+
 The family christensenellaceae is widely inversely associated with BMI. A review is [here](https://bmcbiol.biomedcentral.com/articles/10.1186/s12915-019-0699-4).

### 2022-01-05 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Add dietary inflammation index score

@@
 
+Create a table of participant characteristics by age category in males, females, and by sex.
+
 ```{r Participant characterisitics}
@@
 Is TMAO significantly different by ethnicity?
+
 ```{r Ethnicity}
@@
 
-There are significant differences in TMAO levels by ethnic groups but ethnicity does not predict TMAO.
+There are significant differences in TMAO levels by ethnic groups but ethnicity does not predict TMAO when you control for the covariates sex*age, and cystatin C.
 
-Back to tables... make similar tables but show TMAO quantiles
+Create categorical TMAO groups by quantile and tertile. These will be useful when we are making tables and conducting categorical analyses related to TMAO.
 
-```{r Quantile TMAO}
+Also, transform select variables so that they comply to the normal distribution (they have been tested by the Shapiro Wilk's test in previous analyses).
+
+```{r Quantile and Tertile TMAO}
 # Quantile
@@
 ```
+
+TMAO Quantile Tables
 
@@
 Save table
+
 ```{r Save TMAO linear regression table}
@@
 ## Females
+
 ```{r Cardiometabolic Multiple Linear Regression - Females}
@@
 Save table
+
 ```{r Save TMAO linear regression table - Females}
@@
 ## Males
+
 ```{r Cardiometabolic Multiple Linear Regression - Males}
@@
 full$choln_tnfs_log <- log(full$choln_tnfs)
+```
+
+Load the Dietary Inflammation Index Score
+
+```{r Dietary Inflammation Index}
+# Load the diet scores from Levi and grep for the dietary inflammation index scores only
+inflm <- read.csv("../../data/raw/from_LeviEvans/HEI scores_for LE_6142021.csv", row.names = 1)
+inflm <- inflm[,grep("dii", colnames(inflm))]
+
+# Combine with the full data from above
+full <- merge(full, inflm, by = 0)
+row.names(full) <- full[,"Row.names"]
+full <- subset(full, select = -c(Row.names))
 ```
@@
 Save
+
 ```{r save food linear regression table}
@@
 
+Check the relationship between the dietary inflammation index and fasting plasma TMAO.
+
+```{r TMAO ~ dii - LCMS BD1 - lm 356 ppl, results='hide'}
+transformed_keep_tbl2 <- c("sex", 
+#"sex_numeric",
+"kcal_tnfs", 
+"age", 
+"cystatinc_bd1", 
+"ethnicity",
+"tmao", 
+"tmao_log", 
+"dii_alcohol",
+"dii_b12",
+"dii_b6",
+"dii_betacarotene",
+"dii_caffeine",
+"dii_carb",
+"dii_chol",
+"dii_kcal",
+"dii_totalfat",
+"dii_fiber",
+"dii_folicacid",
+"dii_iron",
+"dii_magnesium",
+"dii_mufa",
+"dii_niacin",
+"dii_omega3",
+"dii_omega6",
+"dii_protein",
+"dii_pufa",
+"dii_riboflavin",
+"dii_satfat",
+"dii_selenium",
+"dii_thiamin",
+"dii_transfat",
+"dii_vita",
+"dii_vitc",
+"dii_vitd",
+"dii_vite",
+"dii_zinc",
+"dii_tea" ,
+"dii_flav3ol",
+"dii_flavone",
+"dii_flavonol",
+"dii_flavonon",
+"dii_anthocy",     
+"dii_isoflav",
+"dii_total_score")
+
+# Make a smaller df from master (the old data)
+full_sub <- full[, colnames(full) %in% transformed_keep_tbl2]
+dii <- subset(full_sub, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
+n <- ncol(dii)
+
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(tmao_log ~ dii[,x] + sex*age + cystatinc_bd1 + kcal_tnfs, full_sub))
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
+my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm356_lcms) <- colnames(dii)
+res_dii <- as.data.frame(my_lm356_lcms)
+
+# BH correction
+res_dii$p.adj <- stats::p.adjust(res_dii$P, method = "BH", n = n) 
+```
+
+There are no significant relationships after adjusting for multiple corrections in the full cohort.
+Prior to multiple testing correction, there is a significant finding with Vitamin C but that is not clearly supported by the literature.
+
+Save for your records.
+
+```{r save food linear regression table}
+# Save table
+write.csv(res_dii, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_DII.csv")
+```
+
+Double check by sex
+
+```{r TMAO ~ dii - LCMS BD1 male and female - lm 356 ppl, results='hide'}
+transformed_keep_tbl2 <- c("sex", 
+#"sex_numeric",
+"kcal_tnfs", 
+"age", 
+"cystatinc_bd1", 
+"ethnicity",
+"tmao", 
+"tmao_log", 
+"dii_alcohol",
+"dii_b12",
+"dii_b6",
+"dii_betacarotene",
+"dii_caffeine",
+"dii_carb",
+"dii_chol",
+"dii_kcal",
+"dii_totalfat",
+"dii_fiber",
+"dii_folicacid",
+"dii_iron",
+"dii_magnesium",
+"dii_mufa",
+"dii_niacin",
+"dii_omega3",
+"dii_omega6",
+"dii_protein",
+"dii_pufa",
+"dii_riboflavin",
+"dii_satfat",
+"dii_selenium",
+"dii_thiamin",
+"dii_transfat",
+"dii_vita",
+"dii_vitc",
+"dii_vitd",
+"dii_vite",
+"dii_zinc",
+"dii_tea" ,
+"dii_flav3ol",
+"dii_flavone",
+"dii_flavonol",
+"dii_flavonon",
+"dii_anthocy",     
+"dii_isoflav",
+"dii_total_score")
+
+# Investigate by sex
+#~~~~~~~~~~~~~
+# FEMALES 
+#~~~~~~~~~~~~~
+# Make a smaller df from master (the old data)
+full_sub_F <- full[, colnames(full) %in% transformed_keep_tbl2]
+full_sub_F <- full_sub_F[full_sub_F$sex == "Female",]
+dii_F <- subset(full_sub_F, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
+n <- ncol(dii_F)
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(tmao_log ~ dii_F[,x] + age + cystatinc_bd1 + kcal_tnfs, full_sub_F ))
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
+my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm356_lcms) <- colnames(dii_F)
+res_dii_F <- as.data.frame(my_lm356_lcms)
+
+# BH correction
+res_dii_F$p.adj <- stats::p.adjust(res_dii_F$P, method = "BH", n = n)
+
+#~~~~~~~~~~~~~
+# MALES 
+#~~~~~~~~~~~~~
+# Make a smaller df from master (the old data)
+full_sub_M <- full[, colnames(full) %in% transformed_keep_tbl2]
+full_sub_M <- full_sub_M[full_sub_M$sex == "Male",]
+dii_M <- subset(full_sub_M, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
+n <- ncol(dii_M)
+# LCMS method
+# Run linear model
+my_lm <- lapply(1:n, function(x) lm(tmao_log ~ dii_M[,x] + age + cystatinc_bd1 + kcal_tnfs, full_sub_M ))
+
+# Organize results
+my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
+my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
+my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
+my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
+my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
+my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
+colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
+row.names(my_lm356_lcms) <- colnames(dii_M)
+res_dii_M <- as.data.frame(my_lm356_lcms)
+
+# BH correction
+res_dii_M$p.adj <- stats::p.adjust(res_dii_M$P, method = "BH", n = n) 
+```
+
+Still no significant results after BH correction. Prior to BH correction, there is a suggested relationship with folic acid and TMAO in males.
+
 Make TMAO food intake by quantile table
+
 ```{r Food Intake by TMAO quantile}

### 2022-01-06 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Calculate total mps for male and females

@@
 
+# Total of meat, poultry, seafood, organ meat, and cured meat (oz. eq.) - MALES
+summary(full$pf_mps_total_totals[full$sex == "Male"])
+sd(full$pf_mps_total_totals[full$sex == "Male"])
+# Total of meat, poultry, seafood, organ meat, and cured meat (oz. eq.) - FEMALES
+summary(full$pf_mps_total_totals[full$sex == "Female"])
+sd(full$pf_mps_total_totals[full$sex == "Female"])
+
 # Eggs (chicken, duck, goose, quail) and egg substitutes (oz. eq.)

### 2022-01-19 Rmarkdown/preprocessing/Phyloseq_Preprocessing_211104.Rmd
Update taxonomy table via MK's suggestions and create an updated PSO named PSO_tmao_NA

@@
 3) Phyloseq object with TMAO releated phenotypes [PSOtmao]
+
+**Updates:**
+On 01-19-2022 I updated the chunks, "Original Data from Mary" and "Make Phyloseq object- specific TMAO variables".
+When MK was reading my manuscript, she noticed that I have some entries in the taxonomy table that say NA and some that say "g__" or "f__".
+While these may have different meanings, she suggests simply changing all cases to NA.
+(Another solution, you could manually differentaite "g__" to "g__ASV1" so that the glomming isn't wrong).
+I added code (from MK) to consistently change the NAs.
+Then, I created a new phyloseq object, PSO_tmao_NA.
+These will be added to the project branch named "update-taxonomy-table" in this GitHub project.
 
@@
 head(taxonTbl$data) # this reveals that the taxonomy table has classifications to species level (in few cases) (click the right arrow to see). 
-# Make taxonomy table human readable
-taxTbl<-taxonTbl$data %>% as_tibble() %>% tidyr::separate(Taxon, sep=";", c("Kingdom","Phylum","Class","Order","Family","Genus","Species"))
+# Make taxonomy table human readable (Nazmul's way and the original way I created the phyloseq objects)
+#taxTbl<-taxonTbl$data %>% as_tibble() %>% tidyr::separate(Taxon, sep=";", c("Kingdom","Phylum","Class","Order","Family","Genus","Species"))
 # View
-taxTbl
+#taxTbl
+
+# Update on 01-19-2022
+# From Mary Kable Jan.14.2022 via email - she suggests we want all unnamed taxa to be NA. For example, an entry that reads "g__" should just say NA. 
+tax <- as.data.frame(taxonTbl$data) %>% dplyr::select(-Confidence) %>%
+  column_to_rownames("Feature.ID") %>%
+  tidyr::separate(Taxon, c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";") %>%
+  as.matrix()
+
+# Trim tax to include only SVs in SV table
+taxt <- tax[c(rownames(tax)%in%rownames(countData)),]
+
+# Replace all empty levels with NA for the sake of consistancy
+# This prevents inaccurate glomming later
+taxt[,grep("Species", colnames(taxt))][grep("s__$", taxt[,grep("Species", colnames(taxt))])] <- NA
+taxt[,grep("Genus", colnames(taxt))][grep("g__$", taxt[,grep("Genus", colnames(taxt))])] <- NA
+taxt[,grep("Family", colnames(taxt))][grep("f__$", taxt[,grep("Family", colnames(taxt))])] <- NA
+taxt[,grep("Order", colnames(taxt))][grep("o__$", taxt[,grep("Order", colnames(taxt))])] <- NA
+taxt[,grep("Class", colnames(taxt))][grep("c__$", taxt[,grep("Class", colnames(taxt))])] <- NA
+taxt[,grep("Phylum", colnames(taxt))][grep("p__$", taxt[,grep("Phylum", colnames(taxt))])] <- NA
 ```
@@
 saveRDS(taxTbl, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
+
+# Save the human readable taxonomy table with consistent NAs per MK's suggestion - 01/19/2022
+write.csv(taxt, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained_consistentNAs.csv")
+saveRDS(taxt, "../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained_consistentNAs.rds")
 ```
@@
 # Taxonomy table
-taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
+#taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds") 
+taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained_consistentNAs.rds") 
 # Phylogenetic Tree, rooted
@@
 # Create phyloseq object
-PSOtmao <- phyloseq(
+PSOtmao_NA <- phyloseq(
   otu_table(ASVr, taxa_are_rows = T), 
   phy_tree(treeR$data), 
-  tax_table(as.data.frame(taxtable) %>% dplyr::select(-Confidence) %>% column_to_rownames("Feature.ID") %>% as.matrix()), #moving the taxonomy to the way phyloseq wants it
+  tax_table(as.data.frame(taxtable) %>% 
+              #dplyr::select(-Confidence) %>% 
+              #column_to_rownames("Feature.ID") %>% 
+              as.matrix()), #moving the taxonomy to the way phyloseq wants it
   sample_data(metadata_a)
 )
-print(PSOtmao)
+print(PSOtmao_NA)
 
 # Save the phyloseq object
-save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData")
+#save(PSOtmao, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData")
+save(PSOtmao_NA, file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_220119.RData")
 ```

### 2022-01-21 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Update the workflow with PSOtmao_NA. Add family level analysis.

--- 2021-12-08 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
@@
 
+**Updates:**
+On 01-19-2022 I reran many of the analyses with the upadted phyloseq object PSOtmao_NA to see if my results were consistent.
+I saved this markdown to the update-taxonomy-table branch of the GitHub FL100_TMAO project.
+Background - When MK was reading my manuscript, she noticed that I have some entries in the taxonomy table that say NA and some that say "g__" or "f__".
+While these may have different meanings, she suggests simply changing all cases to NA.
+(Another solution, you could manually differentaite "g__" to "g__ASV1" so that the glomming isn't wrong).
+
 # Set Up
@@
 ```{r Load phyloseq object}
-load("../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData") #PSOtmao
+#load("../../data/processed/microbiome/phyloseq_objects/PSOtmao_211108.RData") #PSOtmao
+load(file = "../../data/processed/microbiome/phyloseq_objects/PSOtmao_220119.RData") #PSOtmao_NA
+PSOtmao <- PSOtmao_NA
 ```
@@
 # Save
-tiff("../../Outputs/plots/microbiome/RelAbund_Fam_TMAOTertiles_2111.tiff", width = 4, height = 4, units = 'in', res = 300)
+tiff("../../Outputs/plots/microbiome/RelAbund_Fam_TMAOTertiles_2201.tiff", width = 5, height = 4, units = 'in', res = 300)
 plot.composition.relAbun
@@
 alpha.big = 0.5
-sigtab = res.ratio[which(res.ratio$padj < alpha.big), ]
+sigtab = res.ratio[which(res.ratio$padj < alpha), ]
 sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
@@
 
-No results when we look at TMAO median and control for age, sex, and BMI.
+results when we look at TMAO median and control for age, sex, and BMI.
 
@@
 # Save
-tiff("../../Outputs/plots/microbiome/DESeq2_Genus_TMAOTertile1v3_2112.tiff", width = 8, height = 3, units = 'in', res = 300)
+tiff("../../Outputs/plots/microbiome/DESeq2_Genus_TMAOTertile1v3_2201.tiff", width = 8, height = 3, units = 'in', res = 300)
 DESeqPlot
@@
 ```{r save deseq2 tmao tertile results}
-write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv")
+#write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile_2112.csv")
+write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile_220119.csv")
 ```
@@
 ```{r save beta matrices}
-save(PCoA_WU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU.RData")
-save(PCoA_UnWU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU.RData")
-save(PCoA_BC, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC.RData")
-
-save(dismax_wu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU.RData")
-save(dismax_uwu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU.RData")
-save(dismax_bc, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC.RData")
+# Outdated
+#save(PCoA_WU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU.RData")
+#save(PCoA_UnWU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU.RData")
+#save(PCoA_BC, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC.RData")
+
+#save(dismax_wu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU.RData")
+#save(dismax_uwu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU.RData")
+#save(dismax_bc, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC.RData")
+
+# With updated taxonomy table and from PSOtmaoNA
+save(PCoA_WU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU_220119.RData")
+save(PCoA_UnWU, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU_220119.RData")
+save(PCoA_BC, file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC_220119.RData")
+
+save(dismax_wu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU_220119.RData")
+save(dismax_uwu, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU_220119.RData")
+save(dismax_bc, file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC_220119.RData")
 ```
@@
 ```{r Load bdiversity calculations}
-load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU.RData")
-load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU.RData")
-load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC.RData")
-
-load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU.RData")
-load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU.RData")
-load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_WU_220119.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_UnWU_220119.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_PCoA_BC_220119.RData")
+
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_WU_220119.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_UWU_220119.RData")
+load(file = "../../data/processed/microbiome/diversity/PSOtmao1G_dismax_BC_220119.RData")
 
@@
 # ADONIS (perMANOVA) analysis
-adonis2(dismax_wu ~ tmao_log + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .025
+adonis2(dismax_wu ~ tmao_log + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # .076
 #adonis2(tmao_log ~ dismax_wu + sex*age + cystatinc_bd1, data=df, method = "wunifrac", perm=999)
@@
 bdis <- betadisper(dismax_bc, df$tmao_mdn)
-permutest(bdis) # NS, good
+permutest(bdis) # .04
 boxplot(bdis, las = 2)
@@
                     axes = 1:2) +
-  #scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
+  scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
   stat_ellipse(linetype = 1) +
@@
 # Save
-tiff("../../Outputs/plots/microbiome/BetaDiversity_PCoAWU_TMAOmdn_2112.tiff", width = 6, height = 4, units = 'in', res = 300)
+#tiff("../../Outputs/plots/microbiome/BetaDiversity_PCoAWU_TMAOmdn_2112.tiff", width = 6, height = 4, units = 'in', res = 300)
+tiff("../../Outputs/plots/microbiome/BetaDiversity_PCoA_WU_TMAOmdn_2201.tiff", width = 6, height = 4, units = 'in', res = 300)
 betaplot_WU
+dev.off()
+
+# unweighted
+#tiff("../../Outputs/plots/microbiome/BetaDiversity_PCoAWU_TMAOmdn_2112.tiff", width = 6, height = 4, units = 'in', res = 300)
+tiff("../../Outputs/plots/microbiome/BetaDiversity_PCoA_UWU_TMAOmdn_2201.tiff", width = 6, height = 4, units = 'in', res = 300)
+betaplot_UWU
 dev.off()
@@
 
-```{r Save tertile prevalence csv}
-write.csv(prevalence_tert1, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile1.csv")
-write.csv(prevalence_tert2, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile2.csv")
-write.csv(prevalence_tert3, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile3.csv")
+```{r Save Genus tertile prevalence csv}
+write.csv(prevalence_tert1, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile1_2201.csv")
+write.csv(prevalence_tert2, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile2_2201.csv")
+write.csv(prevalence_tert3, "../../Outputs/tables/Microbiome/Prevalence_Genus_Tertile3_2201.csv")
+```
+
+# Family Level Analysis
+
+Repeat TMAO analyses at the Family Level
+
+```{r Family Level}
+# Agglomerate taxa
+PSOtmao1F = tax_glom(PSOtmao1, "Family", NArm = TRUE)
+
+# Now subset at the genus level to identify the most common genus
+# Subset by tertiles (Genus)
+PSOtmao1F_tertile1 <- subset_samples(PSOtmao1F, tmao_tertile == "Tertile1")
+PSOtmao1F_tertile2 <- subset_samples(PSOtmao1F, tmao_tertile == "Tertile2")
+PSOtmao1F_tertile3 <- subset_samples(PSOtmao1F, tmao_tertile == "Tertile3")
+
+#~~~~~~~~~~
+# Tertile 1
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalenceF_tert1 = apply(X = otu_table(PSOtmao1F_tertile1),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1F_tertile1), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalenceF_tert1 = data.frame(Prevalence = prevalenceF_tert1,
+                    TotalAbundance = taxa_sums(PSOtmao1F_tertile1), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1F_tertile1))
+
+#~~~~~~~~~~
+# Tertile 2
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalenceF_tert2 = apply(X = otu_table(PSOtmao1F_tertile2),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1F_tertile2), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalenceF_tert2 = data.frame(Prevalence = prevalenceF_tert2,
+                    TotalAbundance = taxa_sums(PSOtmao1F_tertile2), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1F_tertile2))
+#~~~~~~~~~~
+# Tertile 3
+#~~~~~~~~~~
+# Prevalence - how many samples the taxa is observed in
+prevalenceF_tert3 = apply(X = otu_table(PSOtmao1F_tertile3),
+               MARGIN = ifelse(taxa_are_rows(PSOtmao1F_tertile3), yes = 1, no = 2),
+               FUN = function(x){sum(x > 0)})
+
+prevalenceF_tert3 = data.frame(Prevalence = prevalenceF_tert3,
+                    TotalAbundance = taxa_sums(PSOtmao1F_tertile3), # the total number of occurrences (even if only observed 50 times in 1 sample, the number would be 50)
+                    tax_table(PSOtmao1F_tertile3))
+```
+
+```{r Save Family tertile prevalence csv}
+write.csv(prevalenceF_tert1, "../../Outputs/tables/Microbiome/Prevalence_Family_Tertile1_2201.csv")
+write.csv(prevalenceF_tert2, "../../Outputs/tables/Microbiome/Prevalence_Family_Tertile2_2201.csv")
+write.csv(prevalenceF_tert3, "../../Outputs/tables/Microbiome/Prevalence_Family_Tertile3_2201.csv")
+```
+
+```{r DESeq2 analysis Family - TMAO tertile}
+# DESeq analysis - tmao_tertile
+mb <- phyloseq_to_deseq2(PSOtmao1F, ~ sex + age + bmi_final.x + tmao_tertile)
+mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
+res.ratio <- results(mb.ratio, contrast = c("tmao_tertile", "Tertile1", "Tertile3")) # Look at most different tertiles
+alpha = 0.05
+sigtab = res.ratio[which(res.ratio$padj < alpha), ]
+sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1F)[rownames(sigtab), ], "matrix"))
+sigtab$order_family <- paste0(sigtab$Order, sigtab$Family)
+head(sigtab)
+
+#~~~~~~~~~~~~~~~
+# Plot Results
+#~~~~~~~~~~~~~~~
+resOTU <- rownames(sigtab)
+otuData <- as.data.frame(otu_table(PSOtmao1F))
+otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
+#otuData_desRes <- as.data.frame(t(otuData_desRes))
+# Get sample data groupings
+sampleData <- as.data.frame(sample_data(PSOtmao1F))
+sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
+# Get tax table
+taxaData <- as.data.frame(tax_table(PSOtmao1F))
+taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
+taxaData_res$OrderFamily <- paste0(taxaData_res$Order, taxaData_res$Family)
+
+# Merge
+ggData <- merge(taxaData_res, otuData_desRes, by = 0)
+rownames(ggData) <-  ggData$OrderFamily
+ggData <- ggData[,-(1:9)]
+ggData <- as.data.frame(t(ggData))
+ggData <- merge(ggData, sampleData, by = 0)
+rownames(ggData) <- ggData[,"Row.names"]
+ggData <- subset(ggData, select = -c(Row.names))
+
+# Loop Plot
+nres <- length(resOTU)
+for (i in 1:nres) { 
+    print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+            geom_boxplot(aes(color=tmao_tertile)) +
+            geom_point() +
+            theme_minimal() +
+            ylab(paste0("Log of", colnames(ggData[i]))) +
+            xlab("TMAO Tertile") +
+            scale_color_brewer(palette = "Dark2") +
+            labs(color = "TMAO Tertile")
+    )
+}
+
+# Loop Plot
+nres <- length(resOTU)
+for (i in 1:nres) { 
+    temp_plot <- ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+            geom_boxplot(aes(color=tmao_tertile)) +
+            geom_point() +
+            theme_minimal() +
+            ylab(paste0("Log of", colnames(ggData[i]))) +
+            xlab("TMAO Tertile") +
+            scale_color_brewer(palette = "Dark2") +
+            labs(color = "TMAO Tertile")
+    tiff(paste0("../../Outputs/plots/microbiome/DESeq2_Family_TMAOTertile_Boxplot_", i,"_2201.tiff"), width = 4, height = 4, units = 'in', res = 300)
+    print(temp_plot)
+    dev.off()
+}
+
+#~~~~~~~~~~~~~~~
+# Plot Results Fold Change
+#~~~~~~~~~~~~~~~
+sigtabgen = subset(sigtab, !is.na(Family))
+# Phylum order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
+# Genus order
+x = tapply(sigtabgen$log2FoldChange, sigtabgen$Family, function(x) max(x))
+x = sort(x, TRUE)
+sigtabgen$Family = factor(as.character(sigtabgen$Family), levels=names(x))
+DESeqPlot <- ggplot(sigtabgen, aes(y=order_family, x=log2FoldChange, color=Phylum)) + 
+  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
+  geom_point(size=3) + 
+  ylab("Taxonomy") +
+  xlab("Log2 Fold Change") + 
+  scale_color_brewer(palette = "Paired") +
+  theme_minimal() +
+  theme(axis.text = element_text(size = 12), # x and y axis text size
+        axis.title = element_text(size = 12), # x and y label text size
+        legend.text = element_text(size = 12), # legend text size
+        legend.title = element_text(size = 12), # label legend text size
+        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
+
+DESeqPlot
+
+# Save
+tiff("../../Outputs/plots/microbiome/DESeq2_Family_TMAOTertile1v3_2201.tiff", width = 8, height = 3, units = 'in', res = 300)
+DESeqPlot
+dev.off()
+```
+Save
+
+```{r save deseq2 Family tmao tertile results}
+#write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile_2112.csv")
+write.csv(sigtabgen, "../../Outputs/tables/Microbiome/DESeq2_PSOtmao1F_TMAOTertile_220120.csv")
 ```
@@
 
-```{r Get the microbiome data ready to combine with pheno data}
-da <- read.csv("../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile.csv", row.names = 1)
+```{r Get the genus microbiome data ready to combine with pheno data}
+da <- read.csv("../../Outputs/tables/Microbiome/DESeq2_PSOtmao1G_TMAOTertile_220119.csv", row.names = 1)
+taxaName_famgen <- da[,"family_genus", drop = F]
+taxaName_famgen
 resOTU <- rownames(da)
@@
 # rename for ease downstream
-names(otuData_desRes)[names(otuData_desRes) == "bb1b75f41ff9c9db1d1de41e8388eb52"] <- "f__Bacteroidaceae_g__Bacteroides"
-names(otuData_desRes)[names(otuData_desRes) == "b180a2455b236c103cf0bfe620095736"] <- "f__Peptostreptococcaceae_g__"
-names(otuData_desRes)[names(otuData_desRes) == "e1a2800b24cdf9779b28dc897cddb12a"] <- "f__Christensenellaceae_g__"
+#names(otuData_desRes)[names(otuData_desRes) == "bb1b75f41ff9c9db1d1de41e8388eb52"] <- "f__Bacteroidaceae_g__Bacteroides"
+#names(otuData_desRes)[names(otuData_desRes) == "b180a2455b236c103cf0bfe620095736"] <- "f__Peptostreptococcaceae_g__"
+#names(otuData_desRes)[names(otuData_desRes) == "e1a2800b24cdf9779b28dc897cddb12a"] <- "f__Christensenellaceae_g__"
+#names(otuData_desRes)[names(otuData_desRes) == "8f5ceded1cd7e86b8271d3fab24d322a"] <- "f__Lachnospiraceae_g__Butyrivibrio"
+
+# Update PSOtmao_NA results
+names(otuData_desRes)[names(otuData_desRes) == "a0798db0c868e9ad6089502956d68d87"] <- "f__Erysipelotrichaceae_g__Catenibacterium"
+names(otuData_desRes)[names(otuData_desRes) == "c5859b7f5d14b8c145fbc3ad583c70e8"] <- "f__Erysipelotrichaceae_g__Coprobacillus"
+names(otuData_desRes)[names(otuData_desRes) == "5c4ca852b40641b3eb0ad23e69bb6583"] <- "f__Lachnospiraceae_g__Roseburia"
 names(otuData_desRes)[names(otuData_desRes) == "8f5ceded1cd7e86b8271d3fab24d322a"] <- "f__Lachnospiraceae_g__Butyrivibrio"
@@
 
+Old results:
 f__Bacteroidaceae_g__Bacteroides + f__Peptostreptococcaceae_g__ + f__Christensenellaceae_g__ + f__Lachnospiraceae_g__Butyrivibrio
+
+New results:
+f__Erysipelotrichaceae_g__Catenibacterium + f__Erysipelotrichaceae_g__Coprobacillus + f__Lachnospiraceae_g__Roseburia + f__Lachnospiraceae_g__Butyrivibrio
 
@@
 # Transfored variable list
-transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log", "f__Bacteroidaceae_g__Bacteroides", "f__Peptostreptococcaceae_g__", "f__Christensenellaceae_g__", "f__Lachnospiraceae_g__Butyrivibrio")
+transformed_keep_cardio_tbl <- c("sex", "age","bmi_final", "ethnicity", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln", "cystatinc_bd1", "choline", "betaine", "carnitine","tmao_log", "f__Erysipelotrichaceae_g__Catenibacterium", "f__Erysipelotrichaceae_g__Coprobacillus", "f__Lachnospiraceae_g__Roseburia", "f__Lachnospiraceae_g__Butyrivibrio")
 
@@
 phen_sub <- phen[, colnames(phen) %in% transformed_keep_cardio_tbl]
-cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log, f__Bacteroidaceae_g__Bacteroides, f__Peptostreptococcaceae_g__, f__Christensenellaceae_g__, f__Lachnospiraceae_g__Butyrivibrio)) # change variable here
+#cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log, f__Bacteroidaceae_g__Bacteroides, f__Peptostreptococcaceae_g__, f__Christensenellaceae_g__, f__Lachnospiraceae_g__Butyrivibrio)) # change variable here
+cvd_biomark <- subset(phen_sub, select = -c(sex, age, ethnicity, cystatinc_bd1, tmao_log, f__Erysipelotrichaceae_g__Catenibacterium, f__Erysipelotrichaceae_g__Coprobacillus, f__Lachnospiraceae_g__Roseburia, f__Lachnospiraceae_g__Butyrivibrio)) # change variable here
 n <- ncol(cvd_biomark)
@@
 # Run linear model
-my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex + age + cystatinc_bd1 + f__Bacteroidaceae_g__Bacteroides + f__Peptostreptococcaceae_g__ + f__Christensenellaceae_g__ + f__Lachnospiraceae_g__Butyrivibrio, phen_sub )) # add bugs
+my_lm <- lapply(1:n, function(x) lm(cvd_biomark[,x] ~ tmao_log + sex + age + cystatinc_bd1 + f__Erysipelotrichaceae_g__Catenibacterium + f__Erysipelotrichaceae_g__Coprobacillus + f__Lachnospiraceae_g__Roseburia + f__Lachnospiraceae_g__Butyrivibrio, phen_sub )) # add bugs
 
@@
 
+Save for supplemental material
+```{r Save linear regression results with bacteria}
+write.csv(res_des, "../../Outputs/tables/Microbiome/LinReg_cardiometabolic_withDESeq2results_2201.csv")
+saveRDS(res_des, "../../Outputs/tables/Microbiome/LinReg_cardiometabolic_withDESeq2results_2201.Rds")
+```
+
+Is TNF-a related to the gut microbiome directly?
+## Factored by tnfa Tertile
+
+THIS CHUNK WORK IN PROGRESS
+
+```{r DESeq2 analysis - tnf-a tertile}
+# Quick check
+summary(lm(tnfa_bd1_ln ~ f__Erysipelotrichaceae_g__Catenibacterium, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Erysipelotrichaceae_g__Coprobacillus, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Lachnospiraceae_g__Roseburia, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Lachnospiraceae_g__Butyrivibrio, phen_sub)) # NS
+# With covariates
+summary(lm(tnfa_bd1_ln ~ f__Erysipelotrichaceae_g__Catenibacterium + sex + age + bmi_final, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Erysipelotrichaceae_g__Coprobacillus + sex + age + bmi_final, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Lachnospiraceae_g__Roseburia + sex + age + bmi_final, phen_sub)) # NS
+summary(lm(tnfa_bd1_ln ~ f__Lachnospiraceae_g__Butyrivibrio + sex + age + bmi_final, phen_sub)) # NS
+
+```
+
 The family christensenellaceae is widely inversely associated with BMI. A review is [here](https://bmcbiol.biomedcentral.com/articles/10.1186/s12915-019-0699-4).
 
-```{r Christensenellaceae genus and BMI}
-summary(lm(f__Christensenellaceae_g__ ~ bmi_final + sex + age, phen))
-```
+```{r Christensenellaceae family and BMI}
+# Note = ggData is the family level DESeq results 
+fam_res <- merge(ggData, phen, by = 0)
+rownames(fam_res) <- fam_res[,"Row.names"]
+fam_res <- fam_res[,-1]
+colnames(fam_res) <- gsub(" ", "_", colnames(fam_res))
+
+summary(lm(`_o__Clostridiales_f__Christensenellaceae` ~ bmi_final + sex + age, fam_res))
+```

### 2022-02-02 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Loop plot and name changes

@@
 taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
-taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
+taxaData_res$FamilyGenus <- paste0(taxaData_res$Family, taxaData_res$Genus)
 
@@
 ggData <- merge(taxaData_res, otuData_desRes, by = 0)
-rownames(ggData) <-  ggData$OrderFamilyGenus
+rownames(ggData) <-  ggData$FamilyGenus
 ggData <- ggData[,-(1:9)]
@@
 ggData <- subset(ggData, select = -c(Row.names))
-
-# Loop Plot
+# Save a copy of results for plotting
+ggData_G <- ggData
+
+# Loop Plot to view
 nres <- length(resOTU)
+comparisons <- list(c("Tertile1", "Tertile2"), c("Tertile1", "Tertile3"), c("Tertile2","Tertile3"))
 for (i in 1:nres) { 
-    print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
-            geom_boxplot()+
-            ylab(colnames(ggData)[i])) 
+    print(ggplot(ggData_G, aes(x = tmao_tertile, y = log(ggData_G[,i]))) +
+            geom_boxplot(aes(color=tmao_tertile)) +
+            geom_point() +
+            theme_minimal() +
+            ylab(paste0("Log of", colnames(ggData_G[i]))) +
+            xlab("TMAO Tertile") +
+            scale_color_brewer(palette = "Dark2") +
+            labs(color = "TMAO Tertile") +
+            stat_compare_means(comparisons=comparisons)
+    )
+}
+
+# Loop Plot to save
+nres <- length(resOTU)
+comparisons <- list(c("Tertile1", "Tertile2"), c("Tertile1", "Tertile3"), c("Tertile2","Tertile3"))
+for (i in 1:nres) { 
+    temp_plot <- ggplot(ggData_G, aes(x = tmao_tertile, y = log(ggData_G[,i]))) +
+            geom_boxplot(aes(color=tmao_tertile)) +
+            geom_point() +
+            theme_minimal() +
+            ylab(paste0("Log of \n", colnames(ggData_G[i]))) +
+            xlab("TMAO Tertile") +
+            scale_color_brewer(palette = "Dark2") +
+            labs(color = "TMAO Tertile") 
+            #stat_compare_means(comparisons=comparisons)
+    tiff(paste0("../../Outputs/plots/microbiome/DESeq2_Genus_TMAOTertile_Boxplot_", i,"_2201.tiff"), width = 4, height = 4, units = 'in', res = 300)
+    print(temp_plot)
+    dev.off()
 }
@@
 ggData <- subset(ggData, select = -c(Row.names))
-
-# Loop Plot
+# Save a copy of results for plotting
+ggData_F <- ggData
+
+# Loop Plot to view
 nres <- length(resOTU)
 for (i in 1:nres) { 
-    print(ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+    print(ggplot(ggData_F, aes(x = tmao_tertile, y = log(ggData_F[,i]))) +
             geom_boxplot(aes(color=tmao_tertile)) +
@@
             theme_minimal() +
-            ylab(paste0("Log of", colnames(ggData[i]))) +
+            ylab(paste0("Log of", colnames(ggData_F[i]))) +
             xlab("TMAO Tertile") +
@@
 
-# Loop Plot
+# Loop Plot to save
 nres <- length(resOTU)
 for (i in 1:nres) { 
-    temp_plot <- ggplot(ggData, aes(x = tmao_tertile, y = log(ggData[,i]))) +
+    temp_plot <- ggplot(ggData_F, aes(x = tmao_tertile, y = log(ggData_F[,i]))) +
             geom_boxplot(aes(color=tmao_tertile)) +

### 2022-03-12 Rmarkdown/analysis/Response_to_Reviewers.Rmd
Response to reviewer comments

---
title: "Response to Reviewers"
author: "Kristen James"
date: "3/7/2022"
output: html_document
---

This markdown was created in response to the Reviewer's comments during the peer-review process of our manuscript submitted to Nutrients. The workflow is borrowed in large part from "Characteristics_Diet_CardiometabolicRisk_211111.Rmd" with subtle changes.

Load libraries

```{r Libraries, warning=FALSE, message=FALSE}
library(plyr)
library(tableone)
```

Load data

```{r Load data}
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")

# Complete TMAO data only
phen <- phen[complete.cases(phen$tmao == TRUE),]

# Get rid of NA ethnicity category
phen$ethnicity <- droplevels(phen$ethnicity)

# Factor and recode sex
phen$sex <- as.factor(phen$sex)
phen$sex <- plyr::revalue(phen$sex, c("1" = "Male", "2" = "Female"))

# Make TMAO log variable
phen$tmao_log <- log(phen$tmao)

```

# Reviewer 1

Provide the number of participants in each age-sex-BMI bin to answer Reviewer 1's comment about statistical power.

```{r}
table(phen$bin_number)
```

Reviewer 1 didn't understand that the relationship between TMAO and age differed by sex. We will make it more clear by stating the restults of the statistical test.

```{r sex*age interaction}
summary(lm(tmao_log ~ sex*age  + cystatinc_bd1, phen))
```

# Reviewer 2

## Fiber Comment
Reviewer 2 asked why we did not include fiber in the analysis. We will add total fiber intake from foods and supplements (fibe_tnfs) to our analysis.

First, lets set up the dataframe. This code is borrowed from "Characteristics_Diet_CardiometabolicRisk_211111.Rmd".

```{r Load data and preprocess}
phen <- readRDS("../../data/processed/phenotype/Phenotypes_211108.rds")

# Complete TMAO data only
phen <- phen[complete.cases(phen$tmao == TRUE),]

# Get rid of NA ethnicity category
phen$ethnicity <- droplevels(phen$ethnicity)

# Factor and recode sex
phen$sex <- as.factor(phen$sex)
phen$sex <- plyr::revalue(phen$sex, c("1" = "Male", "2" = "Female"))

# Make TMAO log variable
phen$tmao_log <- log(phen$tmao)

# Convert crp units
phen$crp_bd1.mgdL <- phen$crp_bd1 * 0.0001
```

Create quantile and tertile classifiers. Transform data to comply to normality (Shapiro-Wilk test W statistic > 0.95).

```{r Quantile and Tertile TMAO}
# Quantile
tmao_quantile <- quantile(phen$tmao)
phen$tmao_quantile <- ifelse(phen$tmao <= tmao_quantile[2], "Quantile1",
                             ifelse(phen$tmao > tmao_quantile[2] & phen$tmao <= tmao_quantile[3], "Quantile2",
                                   ifelse(phen$tmao > tmao_quantile[3] & phen$tmao <= tmao_quantile[4], "Quantile3", "Quantile4")))

# Check it
table(phen$tmao_quantile)
phen$tmao_quantile <- as.factor(phen$tmao_quantile)
str(phen$tmao_quantile)

# Tertile
tmao_tertile <- quantile(phen$tmao, c(0:3/3))
phen$tmao_tertile <- ifelse(phen$tmao <= tmao_tertile[2], "Tertile1",
                             ifelse(phen$tmao > tmao_tertile[2] & phen$tmao <= tmao_tertile[3], "Tertile2", "Tertile3"))
# Factor it
phen$tmao_tertile <- factor(phen$tmao_tertile)
table(phen$tmao_tertile)

# Print tertiles for manuscript
tmao_tertile

# Checked all variables for normality via shapiro.test() in console. TG, and glc ere the only ones that didn't pass.
phen$glc_bd1_ln <- log(phen$glc_bd1)
phen$tg_bd1_ln <- log(phen$tg_bd1)
phen$crp_bd1_ln <- log(phen$crp_bd1)
phen$tnfa_bd1_ln <- log(phen$tnfa_bd1)
phen$il6_bd1_ln <- log(phen$il6_bd1)
phen$insulin_bd1_ln <- log(phen$insulin_bd1)
```

Load the diet data.

```{r Load ASA24 averaged data}
food <- readRDS(file = "../../data/processed/asa24/ASA24_meanIntake_withTransformations_211111.rds")

# Combine food with phenotype data
full <- merge(phen, food, by = 0)
row.names(full) <- full[,"Row.names"]
full <- subset(full, select = -c(Row.names))

# Make ASA24 choline log variable
full$choln_tnfs_log <- log(full$choln_tnfs)

# Make ASA24 fiber log variable
shapiro.test(log(full$fibe_tnfs))
full$fibe_tnfs_log <- log(full$fibe_tnfs)
```

Make a list of foods and covariate variables to keep. Add fiber this time!

```{r Foods to keep, results='hide'}
transformed_keep_tbl <- c("sex", 
#"sex_numeric",
"kcal_tnfs", 
"age", 
"cystatinc_bd1", 
"ethnicity",
"tmao", 
"tmao_log", 
#"TMAO.0",
#"TMAO.30",
#"TMAO.180",
#"TMAO.360",
#"lnTMAO.0",
#"lnTMAO.30",
#"lnTMAO.180",
#"lnTMAO.360",
#"TMAO_0_30",
#"TMAO_0_180",
#"TMAO_0_360",
#"Ind.SO4",
#"p.Cresol.SO4",
#"TMAO_quart_num",
#"T.pf_mps_totals_sqrRt",
"T.pf_meat_totals_sqrRt",
"T.pf_curedmeat_totals_sqrRt",
"T.pf_poult_totals_sqrRt",
"T.pf_seafd_hi_totals_Binary",
"T.pf_seafd_low_totals_Binary",
"T.pf_eggs_totals_sqrRt",
"T.pf_nutsds_totals_sqrRt",
"T.pf_legumes_totals_sqrRt",
"T.pf_total_totals_natLog",
#"T.pf_soy_totals_Binary",
"T.v_drkgr_totals_sqrRt",
"T.v_redor_total_totals_sqrRt",
#"T.v_redor_tomato_totals_sqrRt",
#"T.v_redor_other_totals_sqrRt",
"T.v_starchy_total_totals_sqrRt",
#"T.v_starchy_potato_totals_sqrRt",
#"T.v_starchy_other_totals_sqrRt",
"T.v_other_totals_sqrRt",
"T.v_total_totals_sqrRt",
#"T.v_legumes_total_sqrRt",
"T.g_whole_totals_sqrRt",
"T.g_refined_totals_sqrRt",
"g_total_totals",
"T.d_total_totals_sqrRt",
"T.d_milk_totals_sqrRt",
"T.d_cheese_totals_sqrRt",
"T.d_yogurt_totals_Binary",
"choln_tnfs_log",
"hei_asa24_totalscore",
"fibe_tnfs_log"
)
```

Now run the multiple linear regression. Note, we will not include ethnicity in these regressions because although TMAO levels significantly differ between ethnic groups, there was no relationship between TMAO and ethnicity. Additionally, diet patterns differ by ethnicity so we could be over adjusting.

```{r TMAO ~ food - LCMS BD1 - lm 356 ppl, results='hide'}
# Make a smaller df from master (the old data)
full_sub <- full[, colnames(full) %in% transformed_keep_tbl]
food <- subset(full_sub, select = -c(sex, kcal_tnfs, age, cystatinc_bd1, ethnicity, tmao, tmao_log))
n <- ncol(food)

# LCMS method
# Run linear model
my_lm <- lapply(1:n, function(x) lm(tmao_log ~ food[,x] + sex*age + cystatinc_bd1 + kcal_tnfs, full_sub )) # include kcal

# Organize results
my_lm2 <- sapply(my_lm, function(x){summary(x)$adj.r.squared})
my_lm2.5 <- sapply(my_lm, function(x){summary(x)$coefficients[2,1]}) # [food row, estimate] corresponds to food B 
my_lm2.6 <- sapply(my_lm, function(x){summary(x)$coefficients[2,2]}) # [food row, Std.Error] corresponds to food Std. Error 
my_lm2.7 <- sapply(my_lm, function(x){summary(x)$coefficients[2,3]}) # [food row, T value] corresponds to food t value
my_lm3 <- sapply(my_lm, function(x){summary(x)$coefficients[2,4]}) # [food row, P] corresponds to food P value
my_lm356_lcms <- cbind(my_lm2, my_lm2.5, my_lm2.6, my_lm2.7, my_lm3)
colnames(my_lm356_lcms) <- c("adj.r.squared", "estimate", "Std.Error", "T.value","P")
row.names(my_lm356_lcms) <- colnames(food)
res_food <- as.data.frame(my_lm356_lcms)

# BH correction
res_food$p.adj <- stats::p.adjust(res_food$P, method = "BH", n = n) 
```

Save with the extension "_withFiber"

```{r save food linear regression table}
# Save table
write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24_withFiber.csv")
```

We also need to add fiber to our food by TMAO tertile table.

```{r Food Intake by TMAO tertile}
# TMAO table by age bin - females
factorVars <- c("age_cat", "tmao_quantile")
vars <- c("pf_meat_totals", 
          "pf_curedmeat_totals",
          "pf_poult_totals",
          "pf_seafd_hi_totals",
          "pf_seafd_low_totals",
          "pf_eggs_totals",
          "pf_nutsds_totals",
          "pf_legumes_totals",
          "pf_total_totals",
          #"pf_soy_totals_Binary",
          "v_drkgr_totals",
          "v_redor_total_totals",
          #"v_redor_tomato_totals",
          #"v_redor_other_totals",
          "v_starchy_total_totals",
          #"v_starchy_potato_totals",
          #"v_starchy_other_totals",
          "v_other_totals",
          "v_total_totals",
          #"v_legumes_total",
          "g_whole_totals",
          "g_refined_totals",
          "g_total_totals",
          "d_total_totals",
          "d_milk_totals",
          "d_cheese_totals",
          "d_yogurt_totals",
          "hei_asa24_totalscore",
          "choln_tnfs",
          "fibe_tnfs", # fiber added!
          "age",
          "ethnicity" ,
          "bmi_final",
          "tmao",
          "choline",
          "betaine",
          "carnitine")
table_food.ft <- CreateTableOne(vars = vars,
                                    strata = "tmao_tertile",
                                    data = full[full$sex == "Female",],
                                    factorVars = factorVars)
table_food.ft

# TMAO table by age bin - males
table_food.mt <- CreateTableOne(vars = vars,
                                    strata = "tmao_tertile",
                                    data = full[full$sex == "Male",],
                                    factorVars = factorVars)
table_food.mt

# TMAO table by sex
table_food.allt <- CreateTableOne(vars = vars,
                                    strata = "tmao_tertile",
                                    data = full,
                                    factorVars = factorVars)
table_food.allt

# Do BH correction
food_allt_ContTable <- attr(table_food.allt$ContTable, "pValues")
n <- length(food_allt_ContTable$pNormal)
# BH correction
food_allt_ContTable$padj <- stats::p.adjust(food_allt_ContTable$pNormal, method = "BH", n = n) 
food_allt_ContTable
```

Save with the "_withFiber" extension.

```{r Save food tertile tables}
table_save.ft <- print(table_food.ft, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.ft, file = "../../Outputs/tables/ParticipantChar/Food_Female_byTMAOTertile_withFiber.csv")

table_save.mt <- print(table_food.mt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.mt, file = "../../Outputs/tables/ParticipantChar/Food_Male_byTMAOTertile_withFiber.csv")

table_save.allt <- print(table_food.allt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.allt, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOTertile_withFiber.csv")

write.csv(food_allt_ContTable, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOTertile_PadjTable_withFiber.csv")
```

Reviewer 2 asked us to clarify what transformations were used in the endothelial analysis. Ultimately, we used AI in Table 1 and  AI75 in the linear regression analysis. This was because when we use AI75 in the descriptive table, the mean values are negative and positive, which is confusing/not intuitive. The following chunk demonstrates this point.

```{r Participant characterisitics by TMAO tertile}
# TMAO table by tmao_tertile - females
factorVars <- c("age_cat", "tmao_tertile")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.ft <- CreateTableOne(vars = vars,
                                    strata = "tmao_tertile",
                                    data = phen[phen$sex == "Female",],
                                    factorVars = factorVars)
table_tmao.ft

# TMAO table by tmao_tertile - males
factorVars <- c("age_cat", "tmao_tertile")
vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.mt <- CreateTableOne(vars = vars,
                                    strata = "tmao_tertile",
                                    data = phen[phen$sex == "Male",], # note 6103 and 7098 have high il6
                                    factorVars = factorVars)
table_tmao.mt

# TMAO table by tmao_tertile - all, add sex
factorVars <- c("tmao_tertile", "sex")
vars <- c("age", "sex", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
table_tmao.allt <- CreateTableOne(vars = vars,
                                    strata = "tmao_tertile",
                                    data = phen,
                                    factorVars = factorVars)
table_tmao.allt

# Do BH correction for continuous variables
allt_ContTable <- attr(table_tmao.allt$ContTable, "pValues")
n <- length(allt_ContTable$pNormal)
# BH correction
allt_ContTable$padj <- stats::p.adjust(allt_ContTable$pNormal, method = "BH", n = n) 
allt_ContTable

# Do BH correction for categorical variables
allt_CatTable <- attr(table_tmao.allt$CatTable, "pValues")
n <- 23
# BH correction
allt_CatTable$padj <- stats::p.adjust(allt_CatTable$pApprox, method = "BH", n = n) 
allt_CatTable
```

Save with "_al75" extension; however, we did not use these tables (we used the AI version originally created).

```{r Save characteristic tertile tables}
table_save.ft <- print(table_tmao.ft, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.ft, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byTMAOTertile_al75.csv")

table_save.mt <- print(table_tmao.mt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.mt, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byTMAOTertile_al75.csv")

table_save.allt <- print(table_tmao.allt, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(table_save.allt, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOTertile_al75.csv")

# BH corrections for the all tertile table
write.csv(allt_ContTable, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOTertile_Padj_al75.csv")
```

Do some individuals have undiagnosed hyperglycemia or hyperlipidemia? Yes

```{r undiagnosed metabolic disease}
summary(phen$glc_bd1)
summary(phen$tg_bd1)
summary(phen$chol_bd1)

hist(phen$glc_bd1, breaks = 40)
hist(phen$tg_bd1, breaks = 40)
hist(phen$chol_bd1, breaks = 40)
```

### 2022-03-24 Rmarkdown/analysis/Response_to_Reviewers.Rmd
Round2 of review

@@
 library(tableone)
+library(ggplot2)
 ```
@@
 # Reviewer 1
+## Reviewer 1 - Response 1
 
@@
 summary(lm(tmao_log ~ sex*age  + cystatinc_bd1, phen))
+```
+
+## Reviewer 1 - Response 2
+Demonstrate the sex* age interaction in a visual plot for the supplement.
+```{r}
+colnames(phen)
+sd <- sd(phen$tmao_log)
+
+# too messy, point of interaction not obvious
+ggplot(phen, aes(x = age, y = tmao_log, group = sex, color = sex)) +
+  geom_line() +
+  geom_point() +
+  geom_errorbar(aes(ymin=tmao_log - sd, ymax=tmao_log+sd),
+                width = .2,
+                position = position_dodge(0.05)) +
+  theme_minimal() +
+  ylab("Log of TMAO (uM)") +
+  xlab("Age (years)") +
+  scale_color_brewer(palette = "Dark2") +
+  labs(color = "Sex")
+
+# try a simplified version - this is better
+p <- ggplot(phen, aes(x = age, y = tmao_log, group = sex, color = sex)) +
+  geom_point() +
+  geom_smooth(method = lm) +
+  theme_minimal() +
+  ylab("Log of TMAO (uM)") +
+  xlab("Age (years)") +
+  scale_color_brewer(palette = "Dark2") +
+  labs(color = "Sex")
+
+tiff(paste0("../../Outputs/plots/Sex_x_Age_Interaction.tiff"), width = 5, height = 4, units = 'in', res = 300)
+p
+dev.off()
 ```

### 2022-03-24 Rmarkdown/analysis/Response_to_Reviewers.Rmd
Round 2 of review

--- 2022-03-12 Rmarkdown/analysis/Response_to_Reviewers.Rmd
@@
 
-Reviewer 1 didn't understand that the relationship between TMAO and age differed by sex. We will make it more clear by stating the restults of the statistical test.
+Reviewer 1 didn't understand that the relationship between TMAO and age differed by sex. We will make it more clear by stating the results of the statistical test.
 
@@
 summary(lm(tmao_log ~ sex*age  + cystatinc_bd1, phen))
+```
+
+## Reviewer 1 - Response 2
+Demonstrate the sex* age interaction in a visual plot for the supplement.
+```{r}
+colnames(phen)
+sd <- sd(phen$tmao_log)
+
+# Plot
+p <- ggplot(phen, aes(x = age, y = tmao_log, group = sex, color = sex)) +
+  geom_point() +
+  geom_smooth(method = lm) +
+  theme_minimal() +
+  ylab("Log of TMAO (uM)") +
+  xlab("Age (years)") +
+  scale_color_brewer(palette = "Dark2") +
+  labs(color = "Sex")
+
+tiff(paste0("../../Outputs/plots/Sex_x_Age_Interaction.tiff"), width = 5, height = 4, units = 'in', res = 300)
+p
+dev.off()
 ```

### 2022-03-24 Rmarkdown/preprocessing/Phyloseq_Preprocessing_211104.Rmd
Limit to TMAO analysis

@@
 1) Processed QIIME2 inputs in R .rds and .csv formats
-2) Phyloseq object with log transformed Biocrates BD1 data [PSOlogBioc]
-3) Phyloseq object with TMAO releated phenotypes [PSOtmao]
+2) Phyloseq object with TMAO related phenotypes [PSOtmao]
 
@@
 These will be added to the project branch named "update-taxonomy-table" in this GitHub project.
+The branches will eventually be merged to one.
 
@@
 
-## Biocrates Phyloseq Object
-
-Make a phyloseq object for the log transformed biocrates BD1 data. First, we have to do additional preprocessing.
-
-```{r Biocrates BD1 log transformed phyloseq preprocess}
-# Load Mary's original memtadata file... that we converted into a csv in the above chunk. 
-md <- readRDS("../../data/processed/microbiome/phyloseq_inputs/FL100_mapping_file_complete_highest-seq-reps_nocontrols_min_withAlphaStatistics_211104.rds")
-rownames(md) <- md[,"Description"]
-
-# Load phenotype data HERE 
-df <- readRDS("../../data/processed/biocrates/FL100_biocrates_lnTransformed_BD1_211103.rds")
-colnames(df[,1:10])
-
-# Trim df to subjects WITH TMAO (primary variable)
-#df <- df[which(complete.cases(df$p.Cresol.SO4.0) == TRUE),]
-
-# First, check the structure of the data. Make sure everything is numeric/integer. 
-str(df[,1:15])
-
-# In the event that some variables are factors, run the following code. (Can run regardless, if the original is already numeric the output will be identical)
-# In order to make them numeric we first have to turn them to characters. 
-df1 <- df # save a copy
-# This loops says, "if you are a factor, then first become a character, then a number; if you aren't a factor, stay how you are"
-df1[] <- lapply(df1, function(x) {
-    if(is.factor(x)) as.numeric(as.character(x)) else x
-})
-
-# Make categorical variables
-input_data <- df1
-out <- list()
-for(i in seq_along(input_data)){
-  mdn <- lapply(input_data, 2, FUN = median) # find the median for each column
-  out[[i]] <- ifelse(input_data[,i] <= mdn[i], "below", "above") # assign below if value is less than or equal to median
-}
-res <- do.call("cbind",out)
-colnames(res) <- paste0(colnames(input_data), "_mdn")
-rownames(res) <- rownames(input_data)
-
-# Turn to a data frame and rename. Then turn variables into factors as they should be.
-str(res[,1:15])
-df2 <- as.data.frame(res)
-str(df2[,1:15])
-
-# Factor
-df3 <- df2 # back up
-df3[] <- lapply(df3, function(x) {
-  if(is.character(x)) as.factor(x) else x
-})
-# Check the levels. R will factor alphabetically but we want the "control" factor to be the below level
-str(df3[,1:15])
-
-# Relevel so that "below" is the "control". This matters downstream in regressions when contrasts matter.
-df3.5 <- df3
-df3.5[] <- lapply(df3, function(x) {
-  relevel(x, ref = "below")
-})
-str(df3.5)
-
-# Merge categorical data with numeric data
-df4 <- merge(df, df3.5, by = 0)
-rownames(df4) <- df4[,1]
-df4 <- df4[,-1]
-
-# Merge with metadata file
-md2 <- merge(md, df4, by.x = "Description", by.y = 0) # Description = subjectID in the md object
-# 103 subjects. These are subjects with biocrates AND microbiome data.
-
-# Assign 'Description' (subject ID in XXXX terms) to row names
-rownames(md2) <- md2$Description
-```
-
-### Save sample data
-
-```{r Save ln biocrates data}
-write.csv(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.csv")
-saveRDS(md2, "../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.rds")
-```
-
-### Make Phyloseq object - log biocrates BD1
-
-```{r Make Phyloseq object- log biocrates BD1}
-# Load files (metadata, count table, taxonomy table, tree)
-# Metadata
-metadata_a <- readRDS("../../data/processed/microbiome/phyloseq_inputs/lnBiocrates_mdn_sampledata_211108.rds")
-# Count table
-ASVr <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged_table_dada2_no_mito_chloro_highseqreps_0.2rm_subjectID.rds")
-# Taxonomy table
-taxtable <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
-# Phylogenetic Tree, rooted
-treeR <- read_qza(file = "../../data/raw/from_MaryKable/tmao-tree-files/rooted-tree-dada2_no-mito-chloro_0.2rm_TMAOsamps_mc1.qza",  tmp = "Q2Rtemp", rm = F)
-
-# Create phyloseq object
-PSOlogBioc <- phyloseq(
-  otu_table(ASVr, taxa_are_rows = T), 
-  phy_tree(treeR$data), 
-  tax_table(as.data.frame(taxtable) %>% dplyr::select(-Confidence) %>% column_to_rownames("Feature.ID") %>% as.matrix()), #moving the taxonomy to the way phyloseq wants it
-  sample_data(metadata_a)
-)
-print(PSOlogBioc)
-
-# Save the phyloseq object
-save(PSOlogBioc, file = "../../data/processed/microbiome/phyloseq_objects/PSOlogBioc_211108.RData")
-```
-
 ## TMAO Phyloseq Object
 
-```{r Biocrates BD1 log transformed phyloseq preprocess}
+```{r TMAO phyloseq preprocess}
 # Load Mary's original memtadata file... that we converted into a csv in the above chunk. 

### 2022-03-24 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Limit to TMAO analyses

@@
 
-1) Plots per section [expand as analysis goes final]
-2) Tables [expand as analysis goes final]
+1) Plots of differential abundance, and alpha and beta diversity results.
+2) Tables used in the supplemental files of the manuscript.
 
@@
 
-Table of read counts at the Phylum level.
-We will filter low abundant Phyla.
-We will also require that taxa are present in at least 10% of samples (at least 36 samples).
-We will then agglomerate at the Genus level.
+Table of read counts at the Phylum level. We will filter low abundant Phyla. We will also require that taxa are present in at least 5% of samples. We will then agglomerate at the Genus level.
 
@@
 
-## Family Plot of TMAO Quantile
-
-```{r Phylo Abundance Plot by TMAO Quantile}
-# Agglomerate taxa
-PSOtmao1F = tax_glom(PSOtmao1, "Family", NArm = TRUE)
-PSOtmao1F.top10 <- microbiome::aggregate_top_taxa(PSOtmao1F, "Family", top = 10)
-
-ps1F.10.family.comp <- microbiome::transform(PSOtmao1F.top10, transform="compositional")
-
-plot.composition.relAbun <- microbiome::plot_composition(ps1F.10.family.comp,
-                                             sample.sort = "Description",
-                                             x.label = "tmao_quantile",
-                                             group_by = "tmao_quantile",
-                                             average_by = "tmao_quantile") +
-  theme_minimal() +
-  scale_fill_brewer("Family", palette = "Paired") + 
-  ggtitle("Relative Abundance at Family Level")
-
-plot.composition.relAbun
-```
-
 ## Family Plot of TMAO Tertile
@@
 From the DESeq2 [documentation](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html), "Note: In order to benefit from the default settings of the package, you should put the variable of interest at the end of the formula and make sure the control level is the first level."
-
-From Dr. Kable's example,
-
-> DESeq2 Set Up
-> Test the impact of fiber cluster, accounting for covariates, using the log ratio test
->
-> dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
-> This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero.
->
-> dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
-> 
-> The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
-> 
-> resultsNames(dds.ratio.batcho)
-> res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))
 
@@
 
-## Factored by TMAO Quantile
-
-```{r DESeq2 analysis - TMAO quantile}
-# DESeq analysis - tmao_quantile
-mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_quantile)
-mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
-res.ratio <- results(mb.ratio, contrast = c("tmao_quantile", "Quantile1", "Quantile4"))
-alpha = 0.05
-sigtab = res.ratio[which(res.ratio$padj < alpha), ]
-sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
-sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
-head(sigtab)
-
-#~~~~~~~~~~~~~~~
-# Plot Results
-#~~~~~~~~~~~~~~~
-resOTU <- rownames(sigtab)
-otuData <- as.data.frame(otu_table(PSOtmao1G))
-otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
-#otuData_desRes <- as.data.frame(t(otuData_desRes))
-# Get sample data groupings
-sampleData <- as.data.frame(sample_data(PSOtmao1G))
-sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
-# Get tax table
-taxaData <- as.data.frame(tax_table(PSOtmao1G))
-taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
-taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)
-
-# Merge
-ggData <- merge(taxaData_res, otuData_desRes, by = 0)
-rownames(ggData) <-  ggData$OrderFamilyGenus
-ggData <- ggData[,-(1:9)]
-ggData <- as.data.frame(t(ggData))
-ggData <- merge(ggData, sampleData, by = 0)
-rownames(ggData) <- ggData[,"Row.names"]
-ggData <- subset(ggData, select = -c(Row.names))
-
-# Loop Plot
-nres <- length(resOTU)
-for (i in 1:nres) { 
-    print(ggplot(ggData, aes(x = tmao_quantile, y = log(ggData[,i]))) +
-            geom_boxplot()+
-            ylab(colnames(ggData)[i])) 
-}
-
-#~~~~~~~~~~~~~~~
-# Plot Results Fold Change
-#~~~~~~~~~~~~~~~
-sigtabgen = subset(sigtab, !is.na(Genus))
-# Phylum order
-x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
-x = sort(x, TRUE)
-sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
-# Genus order
-x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
-x = sort(x, TRUE)
-sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
-ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
-  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
-  geom_point(size=3) + 
-  ylab("Taxonomy") +
-  xlab("Log2 Fold Change") + 
-  scale_color_brewer(palette = "Dark2") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
-```
-
 # Alpha Diversity Analysis
@@
 
-Play around with the coloring to try to figure out what is discriminating these groups!
-
-```{r What defines the PCoA groups exploration}
-colnames(sample_data(PSOtmao1G))
-loopPheno <- c("pa_ee_total_mdn","endo_ln_rhi_mdn", "endo_al75_mdn", "betaine_mdn", "carnitine_mdn", "tmao_mdn",  "choline_mdn", "creatinine_mdn", "phosphocholine_mdn", "hei_asa24_totalveg_mdn", "hei_asa24_totalproteinfoods_mdn", "hei_asa24_sodium_mdn", "hei_asa24_refinedgrains_mdn", "hei_asa24_sfat_mdn",  "hei_asa24_addsug_mdn", "hei_asa24_wholegrain_mdn", "hei_asa24_totalscore_mdn", "tmao_log_mdn", "TMAO_5", "TMAO_6", "TMAO_7", "sex", "age_cat", "bmi_cat", "tmao_quantile", "tmao_tertile")
-
-# PCoA_WU, PCoA_UnWU, PCoA_BC plug in to view
-for(i in 1:length(loopPheno)){
-print(plot_ordination(PSOtmao1G, PCoA_WU, 
-                    color = loopPheno[i],
-                    shape = NULL,
-                    axes = 1:2) +
-  #scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
-  stat_ellipse(linetype = 1) +
-  labs(color = loopPheno[i]) +
-  theme_minimal() + 
-  theme(axis.text = element_text(size = 12), # x and y axis text size
-        axis.title = element_text(size = 12), # x and y label text size
-        legend.text = element_text(size = 12), # legend text size
-        legend.title = element_text(size = 12), # label legend text size
-        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5)))
-}
-
-# From the plots, Weighted UniFrac and endo_al75 and betaine look different
-bdis <- betadisper(dismax_wu, df$betaine_mdn)
-permutest(bdis) # NS, good
-boxplot(bdis, las = 2)
-
-# ADONIS (perMANOVA) analysis
-adonis2(dismax_wu ~ betaine + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # different
-
-```
-
-# ANCOM
-
-## ANCOM2 no-covariate analysis
-
-```{r ANCOM - genus level}
-source("../../Rscripts/ancom_v2.1.R")
-
-# OTU table
-MicroData <- otu_table(PSOtmao1G)
-MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
-# Add 1
-#MicroData2 <- MicroData +1
-#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
-#dim(MicroData2) # 103 x 98
-
-# Load the phenotype data
-#------------------------
-# Perform ANCOM analysis
-# Step 1: Data preprocessing
-# OTU data
-otu_data <- otu_table(PSOtmao1G)
-otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
-otu_id <- rownames(otu_data)
-
-# Metadata 
-#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
-#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
-# Subset
-MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
-MetaData2$Sample.ID <- rownames(MetaData2)
-
-feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
-out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
-# Function
-prepro = feature_table_pre_process(feature_table = feature_table,
-                                   meta_data = MetaData2,
-                                   sample_var = sample_var,
-                                   group_var = NULL,
-                                   out_cut = out_cut,
-                                   zero_cut = zero_cut,
-                                   lib_cut = lib_cut,
-                                   neg_lb = neg_lb)
-
-# Step 2: ANCOM
-feature_table = prepro$feature_table 
-meta_data = prepro$meta_data
-struc_zero = prepro$structure_zeros
-
-# Define variables for ANCOM
-main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
-adj_formula = NULL ; rand_formula = NULL
-# Run ANCCOM
-t_start = Sys.time()
-res = ANCOM(feature_table = feature_table, 
-            meta_data = meta_data, 
-            struc_zero = struc_zero, 
-            main_var = main_var, 
-            p_adj_method = p_adj_method, 
-            alpha = alpha, 
-            adj_formula = adj_formula, 
-            rand_formula = rand_formula)
-t_end = Sys.time()
-t_run = t_end - t_start 
-t_run # around 5s
-
-# Results
-res$out
-res$fig$data
-
-# Step 3: Volcano Plot
-# Number of taxa except structural zeros
-n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
-# Cutoff values for declaring differentially abundant taxa
-cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
-names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")
-
-# Annotation data
-dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")
-
-fig = res$fig +  
-  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
-  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
-            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
-fig
-
-# ORGANIZE PLOT
-
-TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
-rownames(TaxaData2) <- TaxaData2$Feature.ID
-
-ancomRes <- res$out
-
-test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)
-
-# Make plot 
-tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
-asv_id <- unique(tempdata$taxa_id)
-
-MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]
-
-# Rename to the taxonomy 
-#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
-TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
-# Make a unique naming column that incorporates multiple levels of taxonomy
-TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
-rownames(TaxaData3) <- rownames(TaxaData2)
-# Merge
-MicroData5 <- merge(MicroData4, TaxaData3,
-                    by=0, all.x = T, all.y = F)
-
-row.names(MicroData5) <- MicroData5$OrderFamilyGenus
-
-MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
-MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))
-
-# Merge with the phenotype data 
-MetaData3 <- MetaData2 %>% dplyr::select(tmao_mdn)
-
-ggData <- merge(MetaData3, MicroData5, by =0)
-row.names(ggData) <- ggData$Row.names
-ggData <- ggData %>% dplyr::select(-Row.names)
-
-# Gather the data
-library(tidyr)
-ggData2 <- gather(ggData,
-                  key = "Taxa",
-                  value = "Abundance",
-                  -tmao_mdn)
-
-ggData2$Abundance <- log(ggData2$Abundance)
-
-# Plot it
-p <- ggplot(data = ggData2,
-            aes(x = Taxa, 
-                y = Abundance,
-                fill = tmao_mdn), 
-            alpha = 0.1) +
-  geom_boxplot(lwd=.25) +
-  xlab(NULL) +
-  ylab("Log of Taxa Abundance") +
-  theme_bw() + # note order matters here - need bw to be before theme()
-  theme(axis.text.y = element_text( hjust = 1),
-        axis.text.x = element_text(angle = 60, hjust = 1),
-        text=element_text(size=11))  +
-  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
-  #scale_fill_viridis(discrete = TRUE) +
-  #coord_flip() +
-  labs(fill = "TMAO \n Classification")
-
-p
-```
-
-## ANCOM2 covariate analysis
-
-ANCOM2 lets you include covariates into the analysis. If we add the same covariates as in the DESeq2 analysis, do we get similar results?
-
-```{r ANCOM - genus level with covariates median}
-source("../../Rscripts/ancom_v2.1.R")
-
-# OTU table
-MicroData <- otu_table(PSOtmao1G)
-MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
-# Add 1
-#MicroData2 <- MicroData +1
-#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
-#dim(MicroData2) # 103 x 98
-
-# Load the phenotype data
-#------------------------
-# Perform ANCOM analysis
-# Step 1: Data preprocessing
-# OTU data
-otu_data <- otu_table(PSOtmao1G)
-otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
-otu_id <- rownames(otu_data)
-
-# Metadata 
-#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
-#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
-# Subset
-MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
-MetaData2$Sample.ID <- rownames(MetaData2)
-
-feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
-out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
-# Function
-prepro = feature_table_pre_process(feature_table = feature_table,
-                                   meta_data = MetaData2,
-                                   sample_var = sample_var,
-                                   group_var = NULL,
-                                   out_cut = out_cut,
-                                   zero_cut = zero_cut,
-                                   lib_cut = lib_cut,
-                                   neg_lb = neg_lb)
-
-# Step 2: ANCOM
-feature_table = prepro$feature_table 
-meta_data = prepro$meta_data
-struc_zero = prepro$structure_zeros
-
-# Define variables for ANCOM
-main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
-adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
-# Run ANCCOM
-t_start = Sys.time()
-res = ANCOM(feature_table = feature_table, 
-            meta_data = meta_data, 
-            struc_zero = struc_zero, 
-            main_var = main_var, 
-            p_adj_method = p_adj_method, 
-            alpha = alpha, 
-            adj_formula = adj_formula, 
-            rand_formula = rand_formula)
-t_end = Sys.time()
-t_run = t_end - t_start 
-t_run # around 5s
-
-# Results
-res$out
-res$fig$data
-
-# Step 3: Volcano Plot
-# Number of taxa except structural zeros
-n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
-# Cutoff values for declaring differentially abundant taxa
-cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
-names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")
-
-# Annotation data
-dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")
-
-fig = res$fig +  
-  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
-  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
-            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
-fig
-
-# ORGANIZE PLOT
-
-TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
-rownames(TaxaData2) <- TaxaData2$Feature.ID
-
-ancomRes <- res$out
-
-test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)
-
-# Make plot 
-tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
-asv_id <- unique(tempdata$taxa_id)
-
-MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]
-
-# Rename to the taxonomy 
-#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
-TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
-# Make a unique naming column that incorporates multiple levels of taxonomy
-TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
-rownames(TaxaData3) <- rownames(TaxaData2)
-# Merge
-MicroData5 <- merge(MicroData4, TaxaData3,
-                    by=0, all.x = T, all.y = F)
-
-row.names(MicroData5) <- MicroData5$OrderFamilyGenus
-
-MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
-MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))
-
-# Merge with the phenotype data 
-MetaData3 <- MetaData2 %>% dplyr::select(tmao_mdn)
-
-ggData <- merge(MetaData3, MicroData5, by =0)
-row.names(ggData) <- ggData$Row.names
-ggData <- ggData %>% dplyr::select(-Row.names)
-
-# Gather the data
-library(tidyr)
-ggData2 <- gather(ggData,
-                  key = "Taxa",
-                  value = "Abundance",
-                  -tmao_mdn)
-
-ggData2$Abundance <- log(ggData2$Abundance)
-
-# Plot it
-p <- ggplot(data = ggData2,
-            aes(x = Taxa, 
-                y = Abundance,
-                fill = tmao_mdn), 
-            alpha = 0.1) +
-  geom_boxplot(lwd=.25) +
-  xlab(NULL) +
-  ylab("Log of Taxa Abundance") +
-  theme_bw() + # note order matters here - need bw to be before theme()
-  theme(axis.text.y = element_text( hjust = 1),
-        axis.text.x = element_text(angle = 60, hjust = 1),
-        text=element_text(size=11))  +
-  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
-  #scale_fill_viridis(discrete = TRUE) +
-  #coord_flip() +
-  labs(fill = "TMAO \n Classification")
-
-p
-```
 # Stats for paper
@@
 summary(ppl_occurrences_T3)
-```
-
-How many OTUs are present in 5%, 25%, 50% of samples?
-
-```{r}
-
 ```
@@
 
-THIS CHUNK WORK IN PROGRESS
-
 ```{r DESeq2 analysis - tnf-a tertile}

### 2022-03-24 Rmarkdown/preprocessing/ASA24_Preprocessing_211111.Rmd
limit to useful code

@@
 
-Many of these variables do not have a normal distribution. We will use this data to run parametric statistics so go ahead and transform key variables for future use.
+Many of these variables do not have a normal distribution.
+They have been tested in an external program for the best transformation.
+We will use this data to run parametric statistics so go ahead and transform key variables for future use.
 
@@
 ```
-
-Little exploration below. Pay no mind.
-```{r}
-# Count number of subjects with 2 vs 3 recalls
-dat2_count <- dat2[complete.cases(dat2$lang_tnf),] # 384
-dat3_count <- dat3[complete.cases(dat3$lang_tnf),] # 378
-dat4_count <- dat4[complete.cases(dat4$lang_tnf),] # 347
-
-# What is the distribution of meat intake across the recalls?
-summary(dat2$pf_meat_totals)
-summary(dat3$pf_meat_totals)
-summary(dat4$pf_meat_totals)
-
-summary(dat2$pf_curedmeat_totals)
-summary(dat3$pf_curedmeat_totals)
-summary(dat4$pf_curedmeat_totals)
-
-t.test(dat2$pf_curedmeat_totals, dat4$pf_curedmeat_totals) # not different, good
-
-# List of TMAO subjects
-
-# List of subjects with at least 2 recalls and TMAO data
-
-# List of subjects with at least 3 recalls and TMAO data
-```

### 2022-03-24 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
Limit to paper

--- 2021-12-08 Rmarkdown/analysis/Characteristics_Diet_CardiometabolicRisk_211111.Rmd
@@
 
-1)
-
-**Sources:**
+1) Descriptive tables describing participants
+2) Diet-TMAO analysis results
+3) CVD-TMAO analysis results
 
@@
 ```
+
+Create a table of participant characteristics by age category in males, females, and by sex.
 
@@
 
-```{r Participant characterisitics by TMAO quantile}
-# TMAO table by age bin - females
-factorVars <- c("age_cat", "tmao_quantile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
-table_tmao.fq <- CreateTableOne(vars = vars,
-                                    strata = "tmao_quantile",
-                                    data = phen[phen$sex == "Female",],
-                                    factorVars = factorVars)
-table_tmao.fq
-
-# TMAO table by age bin - males
-factorVars <- c("age_cat", "tmao_quantile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
-table_tmao.mq <- CreateTableOne(vars = vars,
-                                    strata = "tmao_quantile",
-                                    data = phen[phen$sex == "Male",],
-                                    factorVars = factorVars)
-table_tmao.mq
-
-# TMAO table by sex
-factorVars <- c("tmao_quantile")
-vars <- c("age", "ethnicity" , "bmi_final", "waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1", "glc_bd1", "insulin_bd1", "tnfa_bd1", "il6_bd1", "crp_bd1.mgdL", "cystatinc_bd1", "tmao", "choline", "betaine", "carnitine")
-table_tmao.allq <- CreateTableOne(vars = vars,
-                                    strata = "tmao_quantile",
-                                    data = phen,
-                                    factorVars = factorVars)
-table_tmao.allq
-```
-
-Save
-
-```{r Save characteristic quantile tables}
-table_save.fq <- print(table_tmao.fq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Female_byTMAOQuantile_211203.csv")
-
-table_save.mq <- print(table_tmao.mq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Characteristics_Male_byTMAOQuantile_211203.csv")
-
-table_save.allq <- print(table_tmao.allq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Characteristics_All_byTMAOQuantile_211203.csv")
-```
-
-Make similar tables but for TMAO tertile
+Create TMAO tertile descriptive table
 
@@
 write.csv(res_food, "../../Outputs/tables/Diet/MultipleLinReg_TMAO_LCMS_n356_ASA24.csv")
-```
-
-Make TMAO food intake by quantile table
-```{r Food Intake by TMAO quantile}
-# TMAO table by age bin - females
-factorVars <- c("age_cat", "tmao_quantile")
-vars <- c("pf_meat_totals", 
-          "pf_curedmeat_totals",
-          "pf_poult_totals",
-          "pf_seafd_hi_totals",
-          "pf_seafd_low_totals",
-          "pf_eggs_totals",
-          "pf_nutsds_totals",
-          "pf_legumes_totals",
-          "pf_total_totals",
-          #"pf_soy_totals_Binary",
-          "v_drkgr_totals",
-          "v_redor_total_totals",
-          #"v_redor_tomato_totals",
-          #"v_redor_other_totals",
-          "v_starchy_total_totals",
-          #"v_starchy_potato_totals",
-          #"v_starchy_other_totals",
-          "v_other_totals",
-          "v_total_totals",
-          #"v_legumes_total",
-          "g_whole_totals",
-          "g_refined_totals",
-          "g_total_totals",
-          "d_total_totals",
-          "d_milk_totals",
-          "d_cheese_totals",
-          "d_yogurt_totals",
-          "hei_asa24_totalscore",
-          "hei_asa24_totalveg",
-          "hei_asa24_sfat",
-          "choln_tnfs",
-          "age",
-          "ethnicity" ,
-          "bmi_final",
-          "tmao",
-          "choline",
-          "betaine",
-          "carnitine")
-table_food.fq <- CreateTableOne(vars = vars,
-                                    strata = "tmao_quantile",
-                                    data = full[full$sex == "Female",],
-                                    factorVars = factorVars)
-table_food.fq
-
-# TMAO table by age bin - males
-table_food.mq <- CreateTableOne(vars = vars,
-                                    strata = "tmao_quantile",
-                                    data = full[full$sex == "Male",],
-                                    factorVars = factorVars)
-table_food.mq
-
-# TMAO table by sex
-table_food.allq <- CreateTableOne(vars = vars,
-                                    strata = "tmao_quantile",
-                                    data = full,
-                                    factorVars = factorVars)
-table_food.allq
-```
-
-```{r Save food quantile tables}
-table_save.fq <- print(table_food.fq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.fq, file = "../../Outputs/tables/ParticipantChar/Food_Female_byTMAOQuantile.csv")
-
-table_save.mq <- print(table_food.mq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.mq, file = "../../Outputs/tables/ParticipantChar/Food_Male_byTMAOQuantile.csv")
-
-table_save.allq <- print(table_food.allq, quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
-write.csv(table_save.allq, file = "../../Outputs/tables/ParticipantChar/Food_All_byTMAOQuantile.csv")
 ```
@@
 ```
-```
-
-"waistavg_cm", "sysbpavgv2", "diabpavgv2", "endo_rhi", "endo_al75", "hdl_bd1", "ldl_bd1", "chol_bd1", "tg_bd1_ln", "glc_bd1_ln","tnfa_bd1_ln", "il6_bd1_ln", "crp_bd1_ln",
-
-BS section below...
-```{r}
-full$low_mps_meat <- ifelse(full$pf_mps_total_totals <= median(full$pf_mps_total_totals), "low", "high")
-full$low_mps_meat <- ifelse(full$pf_mps_total_totals <= 9.091, "low", "high")
-summary(lm(glc_bd1_ln ~  tmao_log + sex*age + cystatinc_bd1, full[full$low_mps_meat == "high",]))
-summary(lm(glc_bd1_ln ~  tmao_log + sex*age + cystatinc_bd1, full[full$low_mps_meat == "high",]))
-```
